The three empirical studies presented in Chapters~\ref{chp:pinning}, \ref{chp:fake-stars}, and~\ref{chp:cursor} demonstrate the feasibility of explicit causal credibility assessments in software engineering by applying the four-step process---making causal theory explicit, defining the estimand and identifying assumptions, honestly acknowledging limitations, and navigating alternative explanations---to produce trustworthy evidence that challenges widely held software engineering myths.
In each study, the four-step process guided the research design from the outset, ensuring that causal claims were appropriately calibrated to the strength of the evidence.

Building on these contributions, the proposed work extends the thesis by demonstrating a complementary use of the framework: systematically assessing the causal credibility of \emph{existing} empirical research to identify specific methodological gaps and guide the design of future studies that yield stronger evidence.
To this end, I revisit the longstanding debate on whether programming language choice causally affects software defect proneness~\cite{DBLP:conf/sigsoft/RayPFD14, DBLP:journals/cacm/RayPDF17, DBLP:journals/toplas/BergerHMVV19}---a controversy introduced in Chapter~\ref{chp:intro} and used as a running example throughout Chapter~\ref{chp:background}.
I first apply the four-step framework to the debate surrounding the seminal study by \citet{DBLP:conf/sigsoft/RayPFD14} (Section~\ref{sec:proposed-assessment}), showing how explicit causal reasoning can clarify what the existing evidence supports and where the field needs stronger designs.
Then, informed by the assessment results, I propose two complementary empirical studies to address the identified gaps: a panel fixed-effects analysis exploiting within-developer and within-repository variation (Section~\ref{sec:proposed-fe}), and a difference-in-differences study of JavaScript-to-TypeScript migration events (Section~\ref{sec:proposed-did}).
Together, these proposed contributions further demonstrate the feasibility of causal credibility assessments---not only for designing new studies but also for critically evaluating existing research and charting a principled path toward stronger evidence on important software engineering myths.

\section{Revisiting the Programming Language and Defect Proneness Debate}
\label{sec:proposed-language}

\subsection{Background: A Longstanding Controversy}

The question of whether programming language choice causally affects software defect rates is among the most contentious debates in empirical software engineering.
The seminal study by \citet{DBLP:conf/sigsoft/RayPFD14}, published at FSE 2014 and later extended in Communications of the ACM~\cite{DBLP:journals/cacm/RayPDF17}, analyzed 729 projects across 17 programming languages on GitHub and reported that certain language classes---such as functional and statically typed languages---are associated with fewer defect-fixing commits than other classes.
Importantly, the original authors were careful to frame their findings as correlational associations, describing the language effect as ``small but significant'' while acknowledging limitations of their observational design.

Nevertheless, a rigorous reproduction study by \citet{DBLP:journals/toplas/BergerHMVV19} challenged both the robustness and the interpretation of these findings.
The reproduction identified errors in language classification, inadequate handling of polyglot repositories, and concerns about the statistical models employed.
After correcting these issues, the reproduction found that many of the original effects shrank dramatically or failed to replicate entirely.

The ensuing debate~\cite{DBLP:journals/corr/abs-1911-07393, DBLP:journals/corr/abs-1911-11894} exposed a deeper methodological impasse.
The reproduction team argued that an ``analysis of GitHub is necessarily measuring correlation and not establishing causation, as it does not constitute an experiment''~\cite{DBLP:journals/corr/abs-1911-11894}, while the original authors maintained that their statistical findings are robust despite corrections.
This impasse is revealing: Even though Ray et al.\ presented their results as correlational, the question that practitioners and researchers actually care about---\emph{should I choose Language A over Language B to reduce defects?}---is fundamentally causal.
When a study investigates a causal question using correlational methods but without an explicit framework for assessing what the findings can and cannot credibly claim, the resulting ambiguity invites precisely the kind of unproductive controversy that followed.
Neither side had the vocabulary to articulate \emph{under which specific conditions} the reported associations would or would not support causal interpretation, and the debate devolved into generic positions on ``correlation versus causation'' rather than advancing toward concrete, actionable next steps.

I now demonstrate how the four-step causal credibility assessment framework proposed in this thesis can move beyond this impasse.
The goal is not to criticize Ray et al.\ for failing to make causal claims---their correlational framing was appropriate given their design and generated important insights for open-source projects---but to show that applying explicit causal reasoning to a fundamentally causal question clarifies what the existing evidence can support, where its limitations lie, and what specific study designs would yield stronger evidence.

\subsection{Applying the Causal Credibility Assessment Framework}
\label{sec:proposed-assessment}

\paragraph{Step 1: Make the causal theory explicit as a DAG.}
Consistent with their correlational framing, \citet{DBLP:conf/sigsoft/RayPFD14} do not present an explicit causal model.
Their statistical approach---a negative binomial regression of defect-fixing commits on language, controlling for project age, project size, and number of developers---is designed to estimate conditional associations rather than causal effects.
This choice is defensible on its own terms, but it leaves an important question unanswered: What \emph{would} the data-generating process need to look like for the reported associations to reflect genuine causal effects from the choice of programming languages?

Making the causal theory explicit as a DAG (as developed in Figure~\ref{fig:pl-dag} in Chapter~\ref{chp:background}) helps answer this question.
The DAG encodes several confounders---Domain, Team Skill, Org Culture, and Codebase Age---each of which plausibly affects both language choice and defect proneness through independent pathways.
The back-door criterion (Section~\ref{sec:formalism}) shows that conditioning on all four would be necessary to block every confounding path.
Yet the covariates available in the regression model---project age, project size, and developer count---are at best crude proxies for some of these confounders and entirely miss others.
Specifically, the project age partially captures Codebase Age, and the developer count is a noisy proxy for Team Skill;  neither Domain nor Org Culture is measured in the regression model.

The absence of an explicit DAG had concrete consequences for the subsequent debate.
Without a shared visual representation of the assumed data-generating process, neither the original authors nor the reproduction team could identify precisely \emph{which} confounding paths remained open, leading to vague disagreements about whether the findings ``adequately controlled for confounders'' rather than productive engagement with specific threats.
A DAG would have channeled the debate into concrete questions---Is Team Skill adequately proxied by developer count? Is Domain a relevant confounder in this sample?---rather than the generic ``correlation versus causation'' impasse that ultimately resulted.

\paragraph{Step 2: Define the causal estimand, estimation method, and identifying assumptions.}
Ray et al.'s regression coefficient on language type is best understood as a conditional association: the average partial association between programming language and defect-fixing commit counts, conditional on the included covariates.
The authors do not claim this coefficient is causal, and their correlational framing is appropriate given the design.
However, the question that motivates the study and drives practitioner interest---whether choosing one language over another would reduce defects---is inherently causal.
The potential outcomes framework helps clarify the gap between the correlational estimate and the causal quantity of interest.

For the regression coefficient to carry a causal interpretation, the study would require the conditional ignorability assumption (Equation~\eqref{eq:cond-ignorability}): that conditional on project age, size, and developer count, language choice is as-if randomly assigned.
This assumption is implausible given the DAG in Figure~\ref{fig:pl-dag}.
Even if the included covariates perfectly measured the concepts they proxy for, they would fail to block the back-door paths through Domain (application domains differ across languages and independently affect defect rates), Team Skill (more experienced developers select into certain languages and independently write fewer bugs), and Org Culture (organizations with stronger quality cultures adopt certain languages and independently produce fewer defects).
Furthermore, the study's cross-sectional design limits identification to \emph{between-project} comparisons: Python projects are compared with Haskell projects, Java projects with Ruby projects.
The potential outcomes framework makes explicit why this is problematic for causal interpretation: The study estimates $E[Y | D = \text{Haskell}, X] - E[Y | D = \text{Python}, X]$, but this equals the causal effect only if selection bias (Equation~\eqref{eq:selection-bias}) vanishes conditional on $X$---which, given the unblocked back-door paths, it almost certainly does not.
Making these assumptions explicit does not invalidate the correlational findings, but it precisely delineates the conditions under which they would---or would not---support the causal conclusions that practitioners seek.

\paragraph{Step 3: Honestly acknowledge and mitigate limitations.}
Beyond the identification considerations, the study faces several data and measurement challenges that affect how strongly the findings---whether interpreted correlationally or causally---can be trusted.

First, the treatment---``programming language''---is operationalized using GitHub's dominant-language classification, which assigns each repository a single language based on byte counts.
\citet{DBLP:journals/toplas/BergerHMVV19} demonstrated that many repositories are polyglot, containing substantial code in secondary languages, and that reclassifying these repositories substantially changed the estimated effects.
This measurement error in the treatment variable violates the SUTVA requirement that treatment be well-defined: A repository classified as ``JavaScript'' that contains 40\% TypeScript code does not cleanly represent either language's effect.

Second, the outcome---``defect proneness''---is proxied by bug-fix commits identified through regular expression matching on commit messages.
This heuristic is noisy: Not all commits matching bug-fix patterns represent genuine defects, and not all defects are reflected in matching commits.
More critically, different language communities may follow different commit conventions.
If Haskell developers write more descriptive commit messages that are less likely to match bug-fix patterns, the observed association between Haskell and fewer ``bugs'' may partly reflect commit-message culture rather than actual defect rates.

Third, the study does not conduct sensitivity analysis to assess how strong unmeasured confounding would need to be to explain away the reported effects.
Tools such as $E$-values~\cite{vanderweele2017sensitivity} or Rosenbaum bounds~\cite{rosenbaum2002observational} would have provided readers with a quantitative assessment of the findings' robustness to unmeasured confounders, enabling more calibrated interpretation.

\paragraph{Step 4: Navigate alternative explanations.}
Because the study does not frame its investigation in causal terms, it does not systematically consider alternative data-generating processes that could produce the same observed associations.
An explicit causal framework would have enabled---and motivated---engagement with competing explanations, each of which implies a different DAG and different implications for the reported findings.
Four prominent alternatives illustrate how such engagement would work:
\begin{enumerate}[leftmargin=15pt]
    \item \emph{Developer expertise as confounder.} If more skilled developers preferentially adopt functional and statically-typed languages and independently write fewer bugs, the observed association is (partly or wholly) an artifact of developer selection. This alternative implies a DAG where Language $\leftarrow$ Team Skill $\to$ Defect Proneness is an open back-door path. The cross-sectional design cannot distinguish this from a genuine language effect.
    
    \item \emph{Domain confounding.} Certain application domains (e.g., web front-ends, systems programming, data science) cluster around specific languages and have inherently different defect profiles due to differences in complexity, error handling requirements, and testing infrastructure. If domain assignment is correlated with language choice, the language coefficient absorbs domain effects. The study's covariates do not include any domain indicators.
    
    \item \emph{Ecosystem quality as alternative mechanism.} A critic might argue that the apparent benefit of statically-typed languages operates entirely through the maturity of the associated tooling ecosystem (linters, testing frameworks, IDE support) rather than through the type system itself. If ecosystem quality independently improves with language maturity, it acts as a confounder rather than a mediator, and conditioning on it (or failing to) affects the interpretation of the language coefficient.
    
    \item \emph{Reverse causality.} Quality-conscious teams may select ``safer'' languages \emph{because} they care about minimizing defects, reversing the assumed causal direction. Cross-sectional data cannot discriminate Language $\to$ Defects from Quality Consciousness $\to$ Language and Quality Consciousness $\to$ Defects.
\end{enumerate}

Each alternative implies distinct testable predictions---for instance, the developer-expertise alternative predicts that the language association should vanish when comparing the \emph{same developer's} contributions across languages, while the domain-confounding alternative predicts that controlling for application domain should substantially attenuate the association.
These predictions could, in principle, be tested empirically, but neither the original study nor the reproduction derives or examines them.
This is not a criticism of the original authors' stated goals---they did not set out to establish causation---but it illustrates how the absence of explicit causal reasoning left the field without a roadmap for resolving the disagreement that followed.

\subsection{Assessment Summary and Proposed Path Forward}

The four-step assessment is not meant as a criticism of Ray et al., whose correlational framing was appropriate for their study design.
Rather, it illustrates a broader pattern: When empirical software engineering research investigates fundamentally causal questions without explicit causal reasoning, three consequences tend to follow.
\begin{enumerate}[leftmargin=15pt]
    \item \textbf{Ambiguous interpretation}: Without an explicit DAG, readers cannot determine which confounding paths the statistical controls address and which remain open. The resulting ambiguity invites both over-interpretation (treating associations as evidence of causation) and over-skepticism (dismissing all observational evidence as ``merely correlational'').
    \item \textbf{Limited identification}: A cross-sectional design with selection-on-observables provides the weakest form of evidence on the hierarchy discussed in Chapter~\ref{chp:background}. When the covariates are insufficient to block the back-door paths identified in the DAG, neither correlational nor causal interpretation of the regression coefficients is well-supported.
    \item \textbf{Unproductive debate}: Without testable predictions derived from competing causal structures, disagreements devolve into generic positions about ``correlation versus causation'' rather than progressing toward concrete evidence that would resolve the question.
\end{enumerate}

These consequences are not unique to Ray et al.---they reflect a common pattern in empirical software engineering research that investigates causal questions using correlational methods without explicit causal reasoning.
The causal credibility assessment framework proposed in this thesis offers a constructive path forward by identifying two specific methodological advances, each of which motivates a proposed study:
\begin{itemize}[leftmargin=15pt]
    \item \textbf{Consequence 2 $\to$ Proposed Study I (Section~\ref{sec:proposed-fe})}: The cross-sectional design's inability to control for developer- and repository-level confounders motivates a \emph{panel fixed effects analysis} that exploits within-developer and within-repository variation, controlling for all time-invariant confounders without needing to measure them.
    \item \textbf{Consequences 1, 3 $\to$ Proposed Study II (Section~\ref{sec:proposed-did})}: The absence of quasi-experimental variation and the inability to discriminate among alternative causal structures motivate a \emph{difference-in-differences study of JavaScript-to-TypeScript migration events}, which exploits within-repository language transitions as natural experiments and generates testable predictions that distinguish language effects from selection effects.
\end{itemize}

Together, the two proposed studies provide triangulation: If both designs---one controlling for time-invariant confounders through fixed effects, the other exploiting quasi-experimental variation through DiD---converge on similar conclusions, confidence in the causal interpretation grows beyond what either study alone could provide.
If they diverge, the discrepancy itself is informative about which alternative explanations are most plausible.

\section{Proposed Study I: Panel Fixed Effects Analysis}
\label{sec:proposed-fe}

The credibility assessment reveals that the cross-sectional design used by Ray et al.\ cannot disentangle language associations from time-invariant developer and project characteristics that drive both language choice and defect rates.
To address this gap, I propose a panel fixed effects analysis that leverages longitudinal software repository data to control for unobserved heterogeneity at both the developer and repository levels.

\subsection{Research Design}

The key insight is that many developers and repositories work with multiple programming languages over time, creating \emph{within-unit variation} that can be exploited for identification.
Rather than comparing Haskell developers with Python developers (a between-unit comparison confounded by developer characteristics), this design compares the \emph{same developer's} contributions when working in different languages (a within-unit comparison that eliminates time-invariant developer effects).

Specifically, I propose estimating fixed effects models of the form:
\begin{equation}
\textit{Defects}_{ijt} = \alpha_i + \gamma_j + \lambda_t + \beta \cdot \textit{Language}_{ijt} + \Gamma' Z_{ijt} + \epsilon_{ijt}
\end{equation}
where $\textit{Defects}_{ijt}$ represents defect-related commits by developer $i$ in repository $j$ at time $t$; $\alpha_i$ represents developer fixed effects capturing time-invariant developer characteristics (skill level, coding style, attention to quality); $\gamma_j$ represents repository fixed effects capturing time-invariant project characteristics (complexity, domain, team culture); $\lambda_t$ represents time fixed effects capturing temporal trends; and $\textit{Language}_{ijt}$ indicates the programming language used.

This specification controls for confounders that are constant within developers or within repositories.
Referring to the DAG in Figure~\ref{fig:pl-dag}, developer fixed effects absorb the back-door path through Team Skill (to the extent that developer expertise is stable over the study period), and repository fixed effects absorb paths through Domain and Org Culture (to the extent that these are stable within repositories).
The identifying variation comes from developers who contribute to multiple languages or repositories that use multiple languages, enabling within-unit comparisons that are less susceptible to selection bias than the between-unit comparisons in Ray et al.

\paragraph{Data Construction.}
The analysis requires constructing a panel dataset of developer--repository--month observations with associated language and defect information.
I will operationalize this using the dataset shared by \citet{DBLP:journals/toplas/BergerHMVV19}, identifying developers who contribute to repositories in at least two distinct programming languages during the observation period.
Defect-related commits will be identified using the commit-message heuristics of \citet{DBLP:conf/sigsoft/RayPFD14}, supplemented by issue-linked commit analysis where available.
Language assignment at the commit level---rather than the repository level---avoids the polyglot misclassification problem.

\subsection{Expected Challenges and Limitations}

Fixed effects alone cannot address time-varying confounders.
A developer who switches from JavaScript to TypeScript may do so precisely because they are working on a more complex feature requiring stronger typing---in which case the language switch is confounded with feature complexity.
This limitation is intrinsic to fixed effects designs and cannot be resolved without additional identifying variation, motivating the complementary DiD study in Section~\ref{sec:proposed-did}.

Second, the analysis requires sufficient within-unit variation: developers must contribute in multiple languages, and repositories must use multiple languages.
If the majority of developers specialize in a single language, the fixed effects estimator will have limited statistical power.
I will assess the available within-developer variation at the data construction stage and report the fraction of developers contributing to each pair of languages.

Third, the causal estimand in a fixed effects model is the \emph{within-unit} effect, which may differ from the population-average effect if language effects are heterogeneous.
Developers who work in multiple languages may be systematically different from single-language developers (e.g., more versatile, more senior), limiting the generalizability of findings.

\section{Proposed Study II: JavaScript-to-TypeScript Migration}
\label{sec:proposed-did}

While the fixed effects analysis addresses time-invariant confounding, it cannot control for time-varying confounders that change alongside language transitions.
To provide quasi-experimental evidence, I propose exploiting a specific class of language changes---repository-level migrations from JavaScript to TypeScript---as natural experiments for a difference-in-differences design.

\subsection{Research Design}

When a repository migrates from JavaScript to TypeScript, the migration decision is endogenous and cannot be treated as random.
However, the \emph{timing} of migration may exhibit quasi-random variation conditional on the decision to migrate: Some repositories migrate early in the observation period, others migrate late.
This staggered timing creates variation that can be exploited using the difference-in-differences framework applied in Chapter~\ref{chp:cursor}.

Formally, let $E_j$ denote the time of migration for repository $j$.
I propose estimating:
\begin{equation}
\textit{Defects}_{jt} = \gamma_j + \lambda_t + \sum_{h=-k}^{m} \tau_h \cdot \mathbf{1}[t = E_j + h] + \Gamma' Z_{jt} + \epsilon_{jt}
\end{equation}
where $\tau_h$ represents the treatment effect $h$ periods relative to migration.
For $h < 0$, these estimates serve as pre-trend tests; for $h \geq 0$, they estimate the dynamic effect of language migration on defect rates.

This design requires several assumptions for causal interpretation:
\begin{enumerate}[leftmargin=15pt]
    \item \textbf{Parallel trends}: Absent migration, repositories that migrate early would have followed similar defect trajectories as repositories that migrate late.
    \item \textbf{No anticipation}: Defect rates do not change in anticipation of migration (testable via pre-trend analysis).
    \item \textbf{Exogenous timing}: Conditional on the decision to migrate, the timing of migration is not correlated with time-varying determinants of defect rates.
\end{enumerate}

While these assumptions cannot be directly verified, they can be assessed through pre-trend analysis, placebo tests, and robustness checks.
I will strengthen the parallel trends assumption's plausibility through propensity score matching on pre-migration trajectories, following the approach developed in Chapter~\ref{chp:cursor}.

\paragraph{Why JavaScript to TypeScript.}
The JavaScript-to-TypeScript migration offers several methodological advantages for this design.
First, TypeScript is a strict superset of JavaScript, enabling gradual adoption that creates rich within-repository variation and well-defined migration events.
Second, migration events are observable through file extensions and compilation configurations, enabling reliable identification.
Third, the comparison directly tests whether static typing reduces defects---one of the central claims in Ray et al.~\cite{DBLP:conf/sigsoft/RayPFD14}.
Fourth, the JavaScript/TypeScript pair is among the most common in the Cursor study dataset (Chapter~\ref{chp:cursor}), providing a natural connection between the proposed and completed work and opportunities to reuse existing data infrastructure.

\paragraph{Operationalization.}
The analysis will proceed as follows:
\begin{enumerate}[leftmargin=15pt]
    \item \textbf{Identify migration events}: For each repository, track the proportion of code in TypeScript versus JavaScript over time. Define migration as a sustained shift from majority-JavaScript to majority-TypeScript.
    \item \textbf{Construct treatment and comparison groups}: Treated repositories are those that undergo migration during the observation period; comparison repositories are those that remain JavaScript-dominant throughout.
    \item \textbf{Measure defect outcomes}: Operationalize defects using bug-fix commits (identified via commit message patterns~\cite{DBLP:conf/sigsoft/RayPFD14}), static analysis warnings (using the SonarQube infrastructure from Chapter~\ref{chp:cursor}), or issue reports labeled as bugs.
    \item \textbf{Apply DiD estimation}: Estimate the effect of migration using the Borusyak imputation estimator~\cite{borusyak2024revisiting}, which has shown robust performance in Chapter~\ref{chp:cursor}.
\end{enumerate}

\subsection{Expected Challenges and Limitations}

Language migration is endogenous: repositories migrate for reasons (perceived quality benefits, ecosystem maturity, team expertise) that may independently affect defect rates.
While the DiD design addresses some selection concerns by controlling for time-invariant repository characteristics and common temporal trends, it cannot fully eliminate confounding from time-varying factors correlated with both migration timing and defect rates.

The parallel trends assumption may be violated if migrating repositories are on different trajectories than non-migrating repositories before migration.
Pre-trend tests can assess plausibility but cannot definitively validate the assumption.

The generalizability of findings from JavaScript/TypeScript migration to broader language comparisons is limited.
TypeScript and JavaScript share syntax and runtime, making migration relatively seamless compared to more drastic language changes (e.g., from C++ to Rust).
Findings should therefore be interpreted as evidence about the specific effect of adding static typing to a dynamically-typed language, rather than about programming language effects in general.

\section{Expected Contributions}

The proposed work is expected to contribute in several ways.
First, the causal credibility assessment of the programming language debate will demonstrate concretely how the four-step framework can clarify the strengths and limitations of existing research on fundamentally causal questions, identifying specific study designs that would yield stronger evidence---thereby illustrating the framework's value not only for designing new studies (as demonstrated in Chapters~\ref{chp:pinning}, \ref{chp:fake-stars}, and~\ref{chp:cursor}) but also for constructively engaging with existing work.

Second, the two proposed studies will provide new empirical evidence on a longstanding and consequential question, using methods that move beyond the correlational evidence in prior work.
The fixed effects analysis controls for time-invariant developer and project characteristics that confound cross-sectional comparisons, while the DiD study exploits quasi-experimental variation that supports stronger causal claims than selection-on-observables approaches.

Third, the analysis will produce concrete comparisons between correlational and causal inference approaches on the programming language question, complementing the methodological comparisons across the completed empirical studies.
Together, these analyses will provide further evidence for the thesis argument that causal credibility assessments---with their emphasis on explicit theory, transparent assumptions, and honest limitations---are feasible and productive in empirical software engineering research.

Finally, regardless of whether the analysis finds that programming language causally affects defect proneness, the methodological contribution stands.
If the proposed studies find effects consistent with prior work~\cite{DBLP:conf/sigsoft/RayPFD14}, they would provide stronger warrants for the claimed relationships.
If they find null effects consistent with the reproduction~\cite{DBLP:journals/toplas/BergerHMVV19}, they would suggest that prior findings were artifacts of confounding.
Either outcome advances the field by replacing vague debates about ``correlation versus causation'' with specific, evidence-based conclusions grounded in explicit causal reasoning.

\section{Timeline and Feasibility}
\label{sec:proposed-timeline}

Table~\ref{tab:proposed-timeline} presents the anticipated timeline for completing the proposed work.
Both proposed studies build on methodological foundations and data infrastructure developed through the completed empirical studies (Chapters~\ref{chp:pinning}, \ref{chp:fake-stars}, \ref{chp:cursor}), reducing implementation risk and enabling efficient execution.

\begin{table}[h]
\centering
\small
\caption{Proposed timeline for completing the thesis.}
\label{tab:proposed-timeline}
\begin{tabular}{llp{12cm}}
\toprule
\textbf{Phase} & \textbf{Timeline} & \textbf{Activities} \\
\midrule
Phase 1 & Months 1--2 & Construct panel dataset: identify polyglot developers and repositories from GHArchive, operationalize language assignment at the commit level, compute defect metrics across multiple operationalizations. \\
\addlinespace
Phase 2 & Months 2--3 & Implement panel fixed effects analysis (Section~\ref{sec:proposed-fe}); assess within-developer variation; conduct specification tests and robustness checks; write up findings. \\
\addlinespace
Phase 3 & Months 3--5 & Identify JavaScript-to-TypeScript migration events; construct matched treatment and control groups; implement DiD analysis (Section~\ref{sec:proposed-did}); conduct pre-trend tests and sensitivity analyses; write up findings. \\
\addlinespace
Phase 4 & Month 6 & Integrate findings across both studies; synthesize with the causal credibility assessment of Ray et al.; revise and finalize thesis chapters; prepare thesis defense. \\
\bottomrule
\end{tabular}
\end{table}

The proposed timeline reflects several efficiency considerations.
Phase 1 (data construction) serves both proposed studies, as the panel dataset of developers, repositories, languages, and defects provides the foundation for both the fixed effects analysis and the migration study.
The fixed effects analysis (Phase 2) can begin as soon as the panel data is assembled, while migration event identification (Phase 3) requires additional data processing that can proceed in parallel with the early stages of Phase 2.
The SonarQube infrastructure and GHArchive pipelines developed for Chapter~\ref{chp:cursor} can be reused for defect measurement, reducing implementation effort. Table~\ref{tab:risk-assessment} identifies the primary risks to the proposed work and outlines mitigation strategies for each.

\begin{table}
\centering
\footnotesize
\caption{Risk assessment for the proposed work.}
\label{tab:risk-assessment}
\begin{tabular}{p{4cm}ccp{8cm}}
\toprule
\textbf{Risk} & \textbf{Impact} & \textbf{Likelihood} & \textbf{Mitigation} \\
\midrule
Insufficient within-developer language variation for fixed effects & High & Medium & Assess variation early in Phase 1; if insufficient, relax to within-repository variation or developer random effects with Hausman test for model selection. \\
\addlinespace
Parallel trends violation in the DiD study & High & Medium & Apply propensity score matching on pre-migration trajectories following Chapter~\ref{chp:cursor}; conduct pre-trend tests; report sensitivity to alternative matching and estimation specifications. \\
\addlinespace
Too few JS$\to$TS migration events for statistical power & High & Low & Expand observation window; broaden migration definition (e.g., lower the TypeScript share threshold); supplement with related migration pairs (e.g., JavaScript to Flow). \\
\addlinespace
Defect measurement noise from commit heuristics & Medium & High & Triangulate across multiple defect operationalizations (commit heuristics, SonarQube warnings, issue labels); report sensitivity across metrics. \\
\addlinespace
Data construction takes longer than planned & Medium & Medium & Begin data collection immediately; validate pipeline on pilot samples before full-scale processing; parallelize Phases 1 and 2 where possible. \\
\addlinespace
GitHub API rate limits delay data collection & Low & Medium & Leverage GHArchive for historical data to minimize API calls; batch requests; use authenticated endpoints with higher rate limits. \\
\bottomrule
\end{tabular}
\end{table}

\section{Summary}

This chapter has outlined a proposed research program that demonstrates how the causal credibility assessment process demonstrated in this thesis can serve as both a diagnostic tool for existing research and a design guide for future studies.
By applying the four-step framework to the programming language and defect proneness debate, the assessment revealed how investigating a fundamentally causal question without explicit causal reasoning---even when findings are carefully framed as correlational---leads to ambiguous interpretation, limited identification, and unproductive debate that the field struggles to resolve.

These gaps directly motivate two proposed empirical studies.
The panel fixed effects analysis (Section~\ref{sec:proposed-fe}) addresses the identification gap by exploiting within-developer and within-repository variation to control for time-invariant confounders, while the JavaScript-to-TypeScript migration study (Section~\ref{sec:proposed-did}) provides quasi-experimental evidence through a difference-in-differences design that generates testable predictions distinguishing language effects from selection effects.

Together with the completed empirical studies, the proposed work demonstrates the full scope of this thesis's contribution: that \emph{causal credibility assessments are feasible in empirical software engineering through an explicit articulation of causal theory and assumptions}.
The four-step process demonstrated throughout this thesis (1) guides the design of new studies to produce trustworthy evidence (Chapters~\ref{chp:pinning}, \ref{chp:fake-stars}, \ref{chp:cursor}), (2) enables constructive evaluation of existing research on fundamentally causal questions (Section~\ref{sec:proposed-assessment}), and (3) charts a principled path from identified limitations to improved study designs (Sections~\ref{sec:proposed-fe} and~\ref{sec:proposed-did}).
By providing both empirical insights and methodological guidance that emphasizes transparency over perfection, this thesis aims to contribute to the transformation of software engineering into a more evidence-based discipline---not through claims of epistemological certainty, but through the honest accumulation of credible evidence informed by explicit causal reasoning.
