This chapter presents three empirical studies that demonstrate the feasibility of explicit causal credibility assessments in empirical software engineering, using the four-step process introduced in Chapter~\ref{chp:intro}.
Each study challenges a widely held software engineering belief---what I have termed a \emph{software engineering myth}---by deriving an explicit causal theory from domain knowledge, defining the causal estimate of interest and specifying the assumption under which the estimate carries a causal interpretation, honestly acknowledging and mitigating limitations and known caveats in data collection, metric design, measurement infrastructure, and model specifications, and thoughtfully navigating alternative explanations (i.e., alternative causal structures or causal DAGs) that may also be supported with the current results.
Together, these studies illustrate that explicit causal credibility assessments are feasible and productive across diverse research settings in software engineering, even when perfect experimental designs are infeasible.

\section{Empirical Study I: The Impact of Dependency Pinning on Supply Chain Security}
\label{sec:finished-pinning}
\label{chp:pinning}

\subsection{Overview}

The first study addresses a contentious debate in software supply chain security: Should developers \emph{pin} their dependencies to specific versions or allow \emph{floating} version constraints that automatically incorporate updates?
This debate exemplifies the kind of software engineering myth that persists due to conflicting practitioner intuitions and the absence of rigorous quantitative evidence.
Specifically, security practitioners increasingly advocate for dependency pinning as protection against malicious package updates---the rationale being that pinned dependencies prevent automatic installation of compromised versions.
This recommendation has been codified in security scorecards and best practice guidelines, with the OpenSSF Scorecard project including ``pinned dependencies'' among its 18 security best practices~\cite{DBLP:journals/ieeesp/ZahanKHSW23}.
Conversely, software engineering practitioners argue that pinning creates substantial maintenance burdens, accumulates technical debt through outdated dependencies, and paradoxically increases exposure to known security vulnerabilities by blocking automatic propagation of security patches~\cite{DBLP:journals/tse/JafariCAST22}.
Prior research on this topic has been limited to collecting and synthesizing practitioner opinions~\cite{DBLP:journals/tosem/BogartKHT21, DBLP:journals/tse/JafariCAST22, DBLP:conf/sp/LadisaPMB23} or aggregating descriptive statistics from software repositories~\cite{DBLP:conf/msr/0001PSTB19, DBLP:conf/msr/PinckneyCGB23}.
Neither approach provides the quantitative, causal evidence necessary to substantiate causal claims about the impact of pinning, thereby resolving the controversy or informing evidence-based decision-making.

\subsection{Method}

To move beyond prior correlational and qualitative evidence, this study employs a \emph{counterfactual analysis} approach~\cite{morgan2015counterfactuals}---estimating the causal effect of pinning by comparing observed outcomes with simulated ``what-if'' scenarios where each project pins all its direct dependencies.
This design directly embodies the four-step causal credibility assessment process demonstrated in this thesis: The causal theory is derived from domain knowledge and made explicit through the counterfactual simulation structure, the causal estimate of interest is defined through potential outcomes reasoning and the identifying assumption is specified, limitations and known caveats are honestly acknowledged and mitigated, and alternative explanations for the surprising findings are thoughtfully navigated.

\paragraph{Data Collection}
The study curates a dataset of 10,000 npm packages and 10,000 GitHub repositories, ensuring representation of both libraries and applications.
Projects are selected based on popularity (top 5\% of package downloads, top 1\% of repository stars), active maintenance (commits within 2 years), and having production dependencies to analyze.
This sampling strategy ensures findings apply to actively maintained projects that practitioners care about, while avoiding toy projects and obsolete codebases that would introduce noise.

\paragraph{Simulation Setup}
The key methodological innovation involves leveraging an undocumented time-traveling feature of the npm dependency resolver~\cite{DBLP:conf/msr/PinckneyCGB23} to reconstruct historical dependency graphs.
For each project, I resolve dependency graphs at five time points spanning 12 months (September 2022 to September 2023), under two conditions: (1) the \emph{control} condition using actual version constraints from each project's \texttt{package.json}, and (2) the \emph{treatment} condition where all direct dependencies are pinned to specific versions.
This creates a balanced panel dataset with repeated observations under both conditions.

\paragraph{Outcome Metrics}
Five metrics capture the security and maintenance trade-offs commonly debated around the use of pinning and floating version constraints:
\begin{enumerate}[leftmargin=15pt]
    \item \emph{Number of floating dependencies}: The attack surface for malicious updates---packages where a compromised version would be automatically installed.
    \item \emph{Number of automatic updates}: Exposure to breaking changes, assuming such changes occur uniformly among releases.
    \item \emph{Number of security vulnerabilities}: Known vulnerabilities in the dependency graph based on GitHub Advisory entries.
    \item \emph{Number of outdated dependencies}: Direct dependencies with newer versions available, indicating maintenance burden.
    \item \emph{Number of bloated dependencies}: Packages with multiple versions simultaneously present, indicating dependency conflicts.
\end{enumerate}

\paragraph{Statistical Analysis}
The study employs \emph{fixed-effects panel regression}~\cite{bruderl2015fixed} to estimate the causal impact of pinning, controlling for unobserved project- and time-specific heterogeneity.
The model specification includes project-fixed effects ($\alpha$), time-fixed effects ($\beta$), and an interaction term with dependency graph size to test whether pinning's effects vary with project complexity:
\begin{equation}
\ln(M+1) \sim \text{\emph{pinning}} + \ln(size(G)) + \text{\emph{pinning}} \times \ln(size(G)) + \alpha + \beta
\end{equation}
This specification allows the examination of whether pinning's impact differs between projects with small and large dependency graphs---a critical nuance that prior studies did not address.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    var/.style={draw, rounded corners, minimum width=1.6cm, minimum height=0.55cm, font=\small, align=center},
    arr/.style={->, >=stealth, semithick},
    node distance=1cm and 2cm
]

% Causal chain (middle row)
\node[var] (pinning) {Local Version\\Constraints};
\node[var, right=1.5cm of pinning] (npm) {npm Resolution\\(Deduplication)};
\node[var, right=1.5cm of npm] (outcome) {Security \&\\Maint.\ Outcomes};

% Top confounders
\node[var, above left=1.6cm and 0.8cm of pinning] (project) {Project\\Characteristics};
\node[var, above right=1.6cm and -1.5cm of outcome] (time) {Ecosystem\\Evolution};

% Bottom: upstream practices
\node[var, below=1.8cm of npm] (upstream) {Upstream\\Versioning};

% === Causal path ===
\draw[arr] (pinning) -- (npm);
\draw[arr] (npm) -- (outcome);

% === Upstream versioning ===
\draw[arr] (upstream) -- (npm);

% === Project Characteristics ===
\draw[arr] (project) -- (pinning);
\draw[arr] (project) to[out=-5, in=155] (outcome);

% === Ecosystem Evolution ===
\draw[arr] (time) to[out=195, in=40] (npm);
\draw[arr] (time) -- (outcome);
\draw[arr] (time) to[out=215, in=50] (upstream);

\end{tikzpicture}
\caption{The causal DAG explained for the dependency pinning study.
Arrows represent hypothesized direct causal effects.
The causal path of interest runs Local Version Constraints $\to$ npm Resolution (Deduplication) $\to$ Security \& Maintenance Outcomes.
Project Characteristics and Ecosystem Evolution are confounders controlled by the counterfactual simulation design (the same project is observed under both pinned and unpinned conditions at the same time points) and by project-fixed effects ($\alpha$) and time-fixed effects ($\beta$) in the panel regression.
Upstream Versioning---how transitive dependencies version their own dependencies---feeds into the resolution process but is \emph{not} controlled by the project's local pinning decision.
Its interaction with npm's deduplication heuristics drives the counterintuitive finding that local pinning can increase attack surface in large dependency graphs.}
\label{fig:pinning-dag}
\end{figure}

\paragraph{Causal Credibility Assessment}
Following the four-step process demonstrated in this thesis, I assess the causal credibility of this study's findings.
\emph{(1) Causal theory:} The causal theory is made explicit as a DAG in Figure~\ref{fig:pinning-dag} and operationalized through the counterfactual simulation structure.
The treatment (pinning all direct dependencies) affects security and maintenance outcomes through the mediating mechanism of npm's dependency resolution, which also takes as input the upstream versioning practices that the project cannot control.
Project Characteristics and Ecosystem Evolution are confounders controlled by the simulation design (same project under both conditions) and by project- and time-fixed effects in the panel regression.
Crucially, the DAG makes visible the key structural feature driving the study's surprising results: Upstream Versioning feeds into npm Resolution alongside the project's local constraints, so the effect of local pinning is contingent on ecosystem-wide practices---a nuance that prior qualitative debates could not articulate.
\emph{(2) Causal estimate and identifying assumption:} The causal estimate of interest is the average treatment effect on the treated (ATT)---the within-project difference in each outcome metric between pinned and unpinned conditions---identified under the assumption that the counterfactual simulation faithfully reconstructs historical dependency resolution.
This assumption's plausibility is supported by the 83.6--87.7\% resolution success rate and validated against known npm behavior.
Additional identifying assumptions include that malicious attacks occur uniformly at random across packages~\cite{DBLP:conf/dimva/OhmPS020}, and that the conservative baseline of no manual developer intervention constitutes a meaningful lower bound on pinning's costs.
\emph{(3) Limitations:} The study honestly acknowledges that outcome metrics rely on proxy measures (e.g., counting floating dependencies as attack surface rather than actual attack events), that simulations assume no manual developer intervention (a conservative baseline), and that findings are specific to the npm ecosystem and may not generalize to ecosystems with different resolution mechanisms (e.g., Maven's single-version constraint).
\emph{(4) Alternative explanations:} The counterfactual simulation design is uniquely strong in ruling out standard observational confounders, since the same project is observed under both conditions with all other factors held constant.
The primary alternative explanation for the surprising finding---that pinning \emph{increases} attack surface in large dependency graphs---is that this effect is an artifact of npm's specific deduplication heuristics.
The study explicitly traces this mechanism, showing how pinning creates version conflicts that npm resolves by installing duplicate package versions with their own floating transitive dependencies, and argues that even in ecosystems without this specific mechanism, pinning would offer at best marginal security benefits given the dominance of transitive over direct dependencies.

\subsection{Key Findings}

The results challenge the conventional wisdom that pinning provides meaningful security benefits.
While pinning reduces automatic updates (as expected), the study reveals a counterintuitive finding: \emph{pinning direct dependencies can actually increase the attack surface for malicious package updates} in projects with large dependency graphs.
This surprising result stems from how npm's dependency resolver handles version conflicts---pinning creates more opportunities for duplicate package versions, each of which represents an additional attack vector.
For projects with 498 or more dependencies (affecting 26\% of GitHub repositories in the dataset), pinning becomes actively counterproductive for its primary security goal.

Meanwhile, the expected costs of pinning are confirmed: projects that pin their dependencies experience significantly higher exposure to known security vulnerabilities, more outdated dependencies requiring manual updates, and increased dependency conflicts.
These findings contradict security practitioners' recommendations but align with the views of npm developers, who have long regarded pinned dependencies as a ``smell'' indicating accumulated technical debt.

The study also explored alternative ecosystem-level interventions, demonstrating through simulation that coordinated pinning of carefully selected upstream packages could reduce the ecosystem-wide risk of malicious updates by up to 75\%---far more effective than individual developers independently pinning their own dependencies.

From a Bayesian perspective, findings that challenge established best practices require stronger evidence to be credible.
The combination of large-scale data (20,000 projects), longitudinal observation (12 months), counterfactual simulation, and robust statistical modeling with fixed effects provides the kind of compelling evidence necessary to warrant updating beliefs about dependency management practices.

\section{Empirical Study II: The Impact of Fake GitHub Stars on Repository Promotion}
\label{sec:finished-fake-stars}
\label{chp:fake-stars}

\subsection{Overview}

The second study challenges a software engineering myth rooted in the widespread but untested belief that artificially inflating popularity signals---specifically GitHub stars---provides sustainable promotional benefits for open-source projects.
This ``fake it until you make it'' belief drives a thriving black market where repositories purchase fake stars for growth hacking, under the assumption that higher star counts attract genuine attention, adoption, and investment.

GitHub stars serve as a widely used popularity signal when stakeholders evaluate, select, and adopt open-source projects~\cite{DBLP:journals/jss/BorgesV18, DBLP:journals/pacmhci/QiuLPSV19, DBLP:conf/sigsoft/VargasATBG20}.
However, this reliance creates perverse incentives: If star counts influence decision-making, actors have the motivation to manipulate them.
Prior reports from gray literature document repositories buying stars for growth hacking, r\'{e}sum\'{e} fraud, and even malware distribution~\cite{stargazer-goblin}, but systematic empirical investigation of fake stars' prevalence, characteristics, and actual effectiveness has been lacking---precisely the kind of evidence gap that enables software engineering myths to persist.

\subsection{Method}

This study presents the first systematic, global, and longitudinal measurement of fake GitHub stars, combining scalable anomaly detection with causal analysis of promotional effectiveness.
The methodological approach exemplifies the causal credibility assessment process demonstrated in this thesis---the detection tool makes explicit assumptions about what constitutes anomalous behavior, validates performance against known ground truth, and the causal analysis of promotional effectiveness employs panel regression with fixed effects while explicitly deriving a causal theory from domain knowledge, specifying the identifying assumption, honestly acknowledging limitations and known caveats, and thoughtfully navigating alternative explanations.

\paragraph{Detection Tool Design}
I developed \SystemName, a scalable detection tool that identifies two signatures of anomalous starring behavior across the entire GitHub event archive (63.9 million users, 331 million repositories, 326 million stars from 2019--2024):

\begin{itemize}[leftmargin=15pt]
    \item \emph{Low activity signature}: Stars from accounts that become stale after minimal activity---specifically, accounts with only one starring event plus at most one additional event (e.g., fork) on the same day.
    These likely represent throw-away bot accounts created by fake star merchants.
    
    \item \emph{Lockstep signature}: Clusters of accounts repeatedly starring repositories in coordinated time windows.
    Formally, the tool identifies groups of $\geq$50 accounts and $\geq$10 repositories where each repository receives stars from $\geq$25 of these accounts within a 30-day window.
    This pattern is unlikely to form naturally and is highly correlated with fraudulent activities in other social platforms~\cite{DBLP:conf/www/BeutelXGPF13}.
\end{itemize}

A postprocessing step identifies repositories with \emph{fake star campaigns}---those where fake stars constitute a substantial portion of total stars (at least one month with $>$50 fake stars comprising $>$50\% of that month's stars, and $>$10\% of all-time stars).
This distinguishes repositories that likely purchased fake stars from legitimate repositories that may have been incidentally starred by fake accounts.

\paragraph{Validation}
The detection tool's assumptions require empirical validation.
I assessed recall against the Stargazer Ghost Network dataset~\cite{stargazer-goblin}, a confirmed malware campaign using fake stars: \SystemName detected 81\% of the 847 repositories and 76\% of the 15,672 involved accounts.
For precision, I compared deletion rates between detected repositories/accounts and random samples: 90.42\% of detected repositories and 57.07\% of detected accounts had been deleted by GitHub at the time of analysis, compared to baseline deletion rates of 5.03\% and 3.54\%, respectively.
This 16x elevation in deletion rates provides strong evidence that detections correspond to genuinely problematic activity.

\paragraph{Causal Analysis of Promotional Effect}
To test whether fake stars achieve their purported promotional benefits, I employ \emph{panel autoregression} models~\cite{hanck2021introduction}, estimating the longitudinal effect of gaining fake stars on attracting real stars in subsequent months.
For each repository $i$ in each month $t$, I collect: fake stars gained ($\textit{fake}_{i,t}$), cumulative fake stars ($\textit{all\_fake}_{i,t}$), real stars gained ($\textit{real}_{i,t}$), cumulative real stars ($\textit{all\_real}_{i,t}$), repository age, release status, and authentic development activity.

The model specification uses both fixed-effect and random-effect estimators for robustness:
\begin{align*}
    \textit{real}_{i,t} \sim &\sum_{j=1}^k \textit{real}_{i, t-j} + \textit{all\_real}_{i, t - k - 1} + \sum_{j=1}^k \textit{fake}_{i, t-j} + \textit{all\_fake}_{i, t - k - 1} + \text{controls}
\end{align*}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    var/.style={draw, rounded corners, minimum width=1.6cm, minimum height=0.55cm, font=\small, align=center},
    arr/.style={->, >=stealth, semithick},
    node distance=1cm and 2cm
]

% Middle row: causal chain
\node[var] (fake) {Fake Star\\Acquisition};
\node[var, right=2.8cm of fake] (display) {Displayed\\Star Count};
\node[var, right=2.8cm of display] (real) {Real Star\\Gain};

% Top: confounder
\node[var, above left=1.6cm and 0.8cm of fake] (quality) {Maintainer /\\Project Quality};

% Bottom-left: prior real stars
\node[var, below=1.8cm of fake] (prior) {Prior Real\\Stars};

% Bottom-right: controls
\node[var, below=1.8cm of display] (activity) {Dev Activity,\\Age, Releases};

% === Causal path (hypothesized mechanism) ===
\draw[arr] (fake) -- (display);
\draw[arr] (display) -- (real);

% === Rich get richer ===
\draw[arr] (prior) to[out=30, in=-135] (display);

% === Confounder: Maintainer/Project Quality ===
\draw[arr] (quality) -- (fake);
\draw[arr] (quality) to[out=-5, in=155] (real);

% === Controls ===
\draw[arr] (activity) to[out=30, in=-135] (real);

\end{tikzpicture}
\caption{A causal DAG for the fake stars promotional effect analysis.
Arrows represent hypothesized direct causal effects.
The causal path of interest runs Fake Star Acquisition $\to$ Displayed Star Count $\to$ Real Star Gain:
purchased fake stars inflate the displayed count, which may attract genuine attention through the ``rich get richer'' dynamic.
Prior Real Stars also feed into Displayed Star Count and are controlled through autoregressive terms in the panel regression.
Maintainer / Project Quality is the primary confounder: low-quality maintainers may both purchase fake stars and independently fail to attract real attention.
The time-invariant component of this confounder is controlled by repository-fixed effects, but time-varying quality changes remain a potential source of endogeneity.
Dev Activity, Age, and Releases serve as observable control variables that partially proxy for project quality and maturation.}
\label{fig:fake-stars-dag}
\end{figure}

\paragraph{Causal Credibility Assessment}
Following the four-step framework, I assess the causal credibility of this study's findings regarding the promotional effectiveness of fake stars.
\emph{(1) Causal theory:} The causal theory is made explicit as a DAG in Figure~\ref{fig:fake-stars-dag}.
The hypothesized causal mechanism is that acquiring fake stars inflates the displayed star count, which in turn attracts genuine attention and real stars through the ``rich get richer'' dynamic prevalent in social platforms.
However, the DAG also encodes the primary alternative causal structure: Maintainer / Project Quality confounds the relationship, since low-quality or inexperienced maintainers may both purchase fake stars and independently fail to attract real attention.
The ``rich get richer'' effect of prior real stars is separately captured through autoregressive terms, ensuring that the estimated effect of fake stars is not conflated with the natural momentum of genuine popularity.
\emph{(2) Causal estimate and identifying assumption:} The causal estimate of interest is the longitudinal effect of fake star acquisition on subsequent real star gain, estimated through panel autoregression with both fixed-effect and random-effect specifications.
Fixed effects control for time-invariant repository characteristics (e.g., domain, maintainer quality), while autoregressive terms account for temporal dependencies and the natural ``rich get richer'' dynamics of star accumulation.
\emph{(3) Limitations:} The study honestly acknowledges that the panel regression approach provides suggestive but not definitive causal evidence: the time series are too short for formal Granger causality tests, and the detection tool may produce both false positives and false negatives.
Furthermore, measurement of ``fake'' versus ``real'' stars depends on the detection tool's heuristics, introducing potential measurement error that could attenuate estimated effects.
\emph{(4) Alternative explanations:} The most concerning alternative explanation is endogeneity from unobserved maintainer quality: if inexperienced maintainers both buy fake stars and produce low-quality projects that independently fail to attract real attention, the negative long-term coefficient on fake stars would reflect maintainer quality rather than a causal effect of faking.
The fixed-effects specification partially addresses this by controlling for time-invariant repository characteristics, and the short-term \emph{positive} effect of fake stars is harder to explain through this alternative mechanism alone (low-quality projects would not temporarily attract more real stars merely because their maintainers are inexperienced).
Nevertheless, the study positions its causal claims cautiously, acknowledging that future work with stronger identification strategies %(e.g., panel GMM to address endogeneity, as proposed in Chapter~\ref{chp:proposed-work}) 
is needed to establish more credible causal evidence.

\subsection{Key Findings}

Applying \SystemName to all GitHub data from 2019 to 2024 identified 6.0 million suspected fake stars across 18,617 repositories with fake star campaigns.
The measurement study reveals an alarming landscape: fake star campaigns surged dramatically in 2024, with up to 16.66\% of popular repositories showing evidence of fake star campaigns in peak months.
The accounts and repositories involved exhibit highly trivial activity patterns, with 90\% of detected repositories having been deleted by GitHub at the time of analysis.

Critically, the study finds that the majority of fake stars are used to promote short-lived malware repositories rather than legitimate projects seeking growth.
These malicious repositories leverage fake stars to appear trustworthy before distributing cryptocurrency stealers, credential phishing tools, and other malware through seemingly legitimate open-source channels.

Most relevant to the software engineering myth under investigation, the panel regression results directly contradict the ``fake it until you make it'' belief.
While fake stars produce a modest short-term promotional effect---a 1\% increase in fake stars correlates with a 0.07\% increase in real stars in the following month---this effect is approximately five times smaller than the effect of genuine stars (0.36\% increase).
More importantly, the cumulative effect of fake stars becomes \emph{negative} after two months: repositories with more historical fake stars receive fewer real stars in subsequent periods.
Fake stars become a liability rather than an asset, perhaps because repositories relying on fake stars lack the authentic activity and quality indicators that sustain long-term attention.

\section{Empirical Study III: The Impact of Cursor AI on Open-Source Projects}
\label{sec:finished-cursor}
\label{chp:cursor}

\subsection{Overview}

The third study addresses perhaps the most prominent software engineering myth of the current era: The belief that AI coding assistants, particularly modern LLM agents such as Cursor, produce unambiguous and sustained productivity improvements in real software projects.
This belief pervades industry discourse, with developers self-reporting ``multi-fold productivity increases'' and practitioners claiming transformative workflow impacts~\cite{cursor-reddit-survey, cursor-10x}---precisely the kind of experience-derived conventional wisdom that, per the thesis argument, requires rigorous empirical testing rather than uncritical acceptance.
Yet, substantial concerns persist regarding LLM-generated code quality and the long-term consequences of AI-heavy development workflows.
Prior studies have documented that AI coding assistants can produce code with security vulnerabilities~\cite{DBLP:conf/sp/PearceA0DK22, DBLP:conf/ccs/PerryS0B23}, performance issues~\cite{DBLP:conf/issre/LiC0X0S24}, and increased complexity~\cite{DBLP:journals/tse/LiuTLZZ24}.
However, these findings derive primarily from controlled experiments or benchmark analyses that may not generalize to real-world, longitudinal project outcomes.

\subsection{Method}

This study provides the first large-scale empirical investigation of how LLM agent assistants impact project-level development velocity and software quality over time.
The research design employs a \emph{difference-in-differences} (DiD) approach with staggered adoption~\cite{borusyak2024revisiting}---a state-of-the-art causal inference technique that exploits natural variation in when repositories adopted Cursor to identify causal effects.
This design most directly instantiates the four-step causal credibility assessment process introduced in Chapter~\ref{chp:intro}: The causal theory is derived from domain knowledge and encoded in the DiD structure with matched controls; the causal estimate of interest (ATT) is defined through the potential outcomes framework and the identifying assumption (parallel trends) is specified; the assumption's plausibility is assessed through pre-trend tests and strengthened by propensity score matching; and alternative explanations are thoughtfully navigated through matching, fixed effects, and panel GMM analysis.

\paragraph{Treatment Group Identification}
I identify Cursor-adopting repositories through configuration files in git history.
Using the GitHub code search API with adaptive partitioning to circumvent result limits, I discovered repositories with \texttt{.cursorrules} files or \texttt{.cursor} folders.
After filtering for non-fork repositories with $\geq$10 stars (a threshold achieving 97\% precision in identifying engineered projects~\cite{DBLP:journals/ese/MunaiahKCN17}), this yields \numtreated repositories with adoption dates between January 2024 and March 2025.

\paragraph{Control Group Construction via Propensity Score Matching}
A causal interpretation of DiD requires that treatment and control groups would follow similar outcome trajectories absent treatment (the parallel trends assumption).
To strengthen this assumption's plausibility, I construct a matched control group using \emph{propensity score matching}~\cite{austin2011introduction}.

The matching model captures \emph{dynamics} rather than static snapshots, recognizing that rapidly growing repositories may be more likely to adopt new tools.
For each Cursor adoption cohort, I estimate propensity scores via logistic regression:
\begin{equation}
\log\frac{P(\text{treat}|...)}{1-P(\text{treat}|...)} = \alpha + \beta T_{i, t-1} + \sum_{j=1}^{6}\Gamma_j X_{i,t-j} + \Theta\sum_{j=7}^{\infty}X_{i,t-j} + \epsilon_{i}
\end{equation}
where $T_{i,t-1}$ captures repository maturity, $\sum_{j=1}^{6}\Gamma_j X_{i,t-j}$ captures month-by-month evolution over six months (activity levels, stars, forks, issues, pull requests), and $\Theta\sum_{j=7}^{\infty}X_{i,t-j}$ captures historical baselines.
Matching additionally requires the same primary programming language.
This yields a matched control group of 1,380 repositories with similar conditional treatment probabilities (AUC 0.83--0.91).

\paragraph{Outcome Metrics}
For each repository-month, I collect two velocity metrics (commits, lines added) and three quality metrics measured via SonarQube static analysis:
\begin{itemize}[leftmargin=15pt]
    \item \emph{Static analysis warnings}: Total reliability, maintainability, and security issues detected.
    \item \emph{Duplicate line density}: Percentage of duplicated code in the codebase.
    \item \emph{Code complexity}: Cognitive complexity~\cite{DBLP:conf/icse/Campbell18}, quantifying code understandability.
\end{itemize}

\paragraph{DiD Estimation}
I employ the \citet{borusyak2024revisiting} imputation estimator, designed for staggered adoption settings where traditional two-way fixed effects may produce biased estimates.
The estimator proceeds in two steps:
\begin{enumerate}[leftmargin=20pt]
    \item \emph{Impute counterfactual outcomes}: Fit a model using only untreated observations (pre-adoption for treated repositories, all observations for never-treated controls):
    \begin{equation}
    \hat{Y}_{it}(0) = \hat{\mu}_i + \hat{\lambda}_t + \hat{\Gamma}' Z_{it} + \epsilon_{it}
    \end{equation}
    where $\hat{\mu}_i$ and $\hat{\lambda}_t$ are repository-fixed and month-fixed effects, and $Z_{it}$ includes time-varying covariates.
    
    \item \emph{Estimate treatment effects}: For post-adoption observations, compare actual outcomes to counterfactual predictions, yielding average treatment effects on treated ($ATT$) and horizon-specific effects ($ATT_h$).
\end{enumerate}

The repository-fixed effects control all time-invariant confounders (team culture, domain, language), while month-fixed effects control all repository-invariant temporal factors (industry trends, platform changes, seasonal patterns).

\paragraph{Testing Velocity-Quality Interactions}
To investigate whether velocity and quality outcomes interact dynamically---potentially explaining why initial velocity gains might dissipate---I employ \emph{panel generalized method of moments} (GMM)~\cite{arellano1991some}.
GMM uses lagged values as instrumental variables to identify causal relationships when variables may be endogenous:
\begin{equation}
Y_{it} = \hat\mu_i + \hat\lambda_t + \hat\rho Y_{i,t-1} + \hat\beta D_{it} + \hat\gamma X_{it} + \hat{\Gamma}' Z_{it} +\epsilon_{it}
\end{equation}
This allows testing whether technical debt accumulation ($X_{it}$) causally reduces future velocity ($Y_{it}$), and vice versa.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    var/.style={draw, rounded corners, minimum width=1.6cm, minimum height=0.55cm, font=\small, align=center},
    arr/.style={->, >=stealth, semithick},
    node distance=1cm and 2cm
]

% Middle row: causal chain
\node[var] (cursor) {Cursor\\Adoption};
\node[var, right=2.8cm of cursor] (velocity) {Development\\Velocity};
\node[var, right=2.8cm of velocity] (debt) {Technical\\Debt};

% Top confounders
\node[var, above left=1.6cm and 0.8cm of cursor] (project) {Team / Project\\Characteristics};
\node[var, above right=1.6cm and -2cm of debt] (trends) {Industry\\Trends};

% Bottom: other AI tools
\node[var, below=1.8cm of velocity] (ai) {Other AI\\Tool Usage};

% === Causal paths from Cursor ===
\draw[arr] (cursor) -- (velocity);
\draw[arr] (cursor) to[out=15, in=165] (debt);

% === Feedback: technical debt -> future velocity (GMM) ===
\draw[arr, dashed] (debt) to[out=-150, in=-30] node[below, font=\scriptsize] {$t \to t{+}1$} (velocity);

% === Team / Project Characteristics (confounder) ===
\draw[arr] (project) -- (cursor);
\draw[arr] (project) to[out=-5, in=155] (debt);

% === Industry Trends ===
\draw[arr] (trends) to[out=195, in=40] (velocity);
\draw[arr] (trends) -- (debt);

% === Other AI tools (contamination) ===
\draw[arr] (ai) -- (velocity);
\draw[arr] (ai) to[out=30, in=-90] (debt);

\end{tikzpicture}
\caption{A causal DAG for the Cursor adoption study.
Solid arrows represent hypothesized direct causal effects identified through the DiD design.
The treatment (Cursor Adoption) has two direct causal paths: a \emph{transient} positive effect on Development Velocity and a \emph{persistent} direct effect on Technical Debt (code complexity and static analysis warnings).
The dashed arrow represents the dynamic feedback from Technical Debt to future Development Velocity, identified through panel GMM analysis: accumulated technical debt causally reduces subsequent velocity, creating a self-reinforcing cycle that explains why initial velocity gains dissipate over time.
Team / Project Characteristics is the primary confounder, controlled by repository-fixed effects ($\mu_i$) and propensity score matching on pre-adoption trajectories.
Industry Trends are controlled by month-fixed effects ($\lambda_t$).
Other AI Tool Usage represents potential contamination from tools like GitHub Copilot present in both treatment and control groups; estimates thus reflect Cursor's \emph{additional} impact relative to current practice.}\label{fig:cursor-dag}
\end{figure}

\paragraph{Causal Credibility Assessment}
Of the three studies, this one most directly instantiates the four-step causal credibility assessment process, employing a quasi-experimental design from the econometrics literature.
\emph{(1) Causal theory:} The causal theory is made explicit as a DAG in Figure~\ref{fig:cursor-dag}.
Cursor adoption (treatment) has two direct causal paths: a transient positive effect on Development Velocity and a persistent direct effect on Technical Debt (code complexity and static analysis warnings).
The dashed feedback arrow from Technical Debt to future Development Velocity---identified through panel GMM analysis---represents the key dynamic interaction: accumulated technical debt causally reduces subsequent velocity, creating the self-reinforcing cycle that explains why initial velocity gains dissipate.
Potential confounders---Team / Project Characteristics (team culture, project maturity, developer expertise, application domain)---are controlled by repository-fixed effects ($\mu_i$) and propensity score matching on pre-adoption trajectories.
Industry Trends are controlled by month-fixed effects ($\lambda_t$).
Other AI Tool Usage represents contamination from tools like GitHub Copilot in both groups, meaning estimates reflect Cursor's additional impact relative to current practice.
\emph{(2) Causal estimate and identifying assumption:} The causal estimate of interest is the average treatment effect on the treated ($ATT$) and its horizon-specific decomposition ($ATT_h$), formalized through the potential outcomes framework: For each treated repository-month, the individual treatment effect is $Y_{it} - Y_{it}(0)$, where $Y_{it}(0)$ is the counterfactual outcome absent Cursor adoption, imputed from untreated observations via the \citet{borusyak2024revisiting} estimator.
Identification rests on the parallel trends assumption---absent Cursor adoption, treated and control repositories would have followed similar outcome trajectories---which is strengthened by propensity score matching on pre-adoption dynamics and assessed through pre-trend tests that all outcomes pass at the 0.05 significance level.
\emph{(3) Limitations:} The identification strategy relies on observable adoption through committed configuration files, which may not capture all Cursor usage---estimates thus represent intent-to-treat effects for repositories with deliberate, visible adoption.
The parallel trends assumption, while supported by pre-trend tests and matching, remains fundamentally untestable.
Quality metrics are limited to static-analysis-based proxies (warnings, complexity, duplication) that capture only one dimension of software quality.
Findings may not generalize to enterprise settings, other AI tools, or programming languages beyond the dataset's dominant TypeScript, Python, and JavaScript.
\emph{(4) Alternative explanations:} Two primary alternative explanations are considered.
First, potential contamination from alternative AI tools (e.g., GitHub Copilot) in both groups means that the estimated effects reflect Cursor's \emph{additional} impact relative to the current state-of-practice; this interpretation is conservative and does not undermine the finding of quality degradation, since it represents a lower bound on total AI-induced quality effects.
Second, selection effects---where repositories experiencing rapid growth might both adopt Cursor and subsequently slow down due to natural regression to the mean---are addressed through the matching design (which controls for pre-adoption growth trajectories) and the DiD framework (which differences out repository-specific baselines).
The panel GMM analysis provides additional evidence against this alternative by revealing a specific causal mechanism: Technical debt accumulation causally reduces future velocity, consistent with the proposed interpretation rather than regression to the mean.

\subsection{Key Findings}

The findings reveal a nuanced picture that challenges both unbridled optimism and categorical pessimism about AI-assisted coding.
Cursor adoption produces substantial but \emph{transient} velocity gains: Projects experience 3-5x increases in lines added in the first month post-adoption, but these gains dissipate entirely after two months, with velocity returning to baseline levels.

Concurrently, the study documents the \emph{persistent} accumulation of technical debt.
Static analysis warnings increase by 30\% and code complexity increases by 41\% on average post-adoption, and these quality degradations do not diminish over the observation period.
Panel GMM analysis reveals that accumulated technical debt subsequently reduces future development velocity, creating a self-reinforcing cycle in which initial productivity surges give way to maintenance burdens.
Notably, the code complexity increase persists even after controlling for velocity dynamics, providing evidence that code generated by LLM agents may be inherently more complex than human-written code---a ``complexity debt'' that compounds over time.

Perhaps most importantly, the study illustrates how pragmatic causal inference can reveal complex temporal dynamics that simpler methodological approaches would miss.
The transient velocity gains followed by persistent quality degradation, and the feedback loop between technical debt and subsequent velocity, emerged only through longitudinal analysis with appropriate statistical modeling.
Cross-sectional comparisons or short-term experiments would have produced misleading conclusions about the effectiveness of AI coding assistants.

\section{Summary}

These three empirical studies collectively demonstrate the feasibility and value of explicit causal credibility assessments for developing trustworthy evidence that challenges software engineering myths and informs evidence-based practice.
Each study targets a widely held belief---that dependency pinning improves security, that fake popularity signals provide sustainable promotional benefits, and that AI coding assistants produce unambiguous productivity improvements---and in each case, rigorous empirical analysis reveals a more nuanced reality than conventional wisdom suggests.

Importantly, the studies illustrate all four steps of the causal credibility assessment process introduced in Chapter~\ref{chp:intro}, albeit at different levels of formality appropriate to each research context.
Each derives an explicit causal theory from domain knowledge as a DAG---whether encoding how version constraints interact with npm's dependency resolution through a counterfactual simulation (pinning, Figure~\ref{fig:pinning-dag}), contrasting the hypothesized promotional dynamics with the alternative confounding structure from unobserved maintainer quality (fake stars, Figure~\ref{fig:fake-stars-dag}), or encoding the treatment mechanism with its two direct causal paths and the dynamic feedback loop between technical debt and velocity identified through panel GMM (Cursor, Figure~\ref{fig:cursor-dag}).
Each defines a causal estimate of interest through the potential outcomes framework---the within-project ATT of pinning (pinning), the longitudinal effect of fake on real star acquisition (fake stars), or the ATT and horizon-specific effects of Cursor adoption---and specifies the assumption under which the estimate carries a causal interpretation.
Each honestly acknowledges limitations and known caveats in data collection, metric design, measurement infrastructure, and model specifications, allowing readers to calibrate their confidence accordingly.
And each thoughtfully navigates alternative explanations: The counterfactual design rules out most observational confounding by construction; endogeneity from unobserved maintainer quality is explicitly considered and partially addressed through fixed effects; selection effects and contamination are addressed through matching, pre-trend tests, and panel GMM.

The three studies also occupy different positions in the hierarchy of evidence discussed in Chapter~\ref{chp:background}: The counterfactual simulation (pinning) provides strong internal validity within its simulation assumptions but limited external validity beyond the npm ecosystem; the panel regression (fake stars) provides suggestive but not definitive causal evidence under selection-on-observables assumptions, occupying a lower tier in the hierarchy; and the DiD design (Cursor) provides quasi-experimental evidence under a partially testable parallel trends assumption, sitting between the other two in causal credibility.
Together, they provide a template for how software engineering research can move beyond descriptive correlations and synthesized practitioner opinions toward credible causal evidence, while demonstrating that the appropriate level of methodological rigor depends on the research context, data availability, and the specific threats to validity present in each setting.
