\section{Motivation}

Software engineering is among the most consequential engineering disciplines of the modern era. 
Software systems underpin critical infrastructure across healthcare, finance, transportation, and communication, with failures potentially affecting millions of users, costing billions of dollars, and projecting real-world consequences~\cite{eghbal2016roads, solar-wind, xz-attack}. 
The decisions software engineers make, and the practices they adopt daily (e.g., dependency management strategies, tool adoption choices), accumulate into system-wide outcomes that shape organizational productivity, security posture, and long-term maintainability. 
Given these stakes, one might expect software engineering practice to be grounded in rigorous, evidence-based principles analogous to those in medicine~\cite{sackett1996evidence} or civil engineering~\cite{petroski1985engineer}. 
Unfortunately, this expectation remains largely unfulfilled.

In fact, software engineering practice today is characterized more by folklore, intuition, and contested conventional wisdom than by empirically validated guidance~\cite{DBLP:conf/icse/Devanbu0B16, DBLP:journals/ese/GarousiBO20, DBLP:journals/ese/ShrikanthNFM21, enoiu2026folklore}. 
While practitioners face consequential decisions on a routine basis, they very often find limited reliable evidence to inform their choices. 
To support their decisions, they draw on their own experiences or navigate a landscape of blog posts, podcasts, and conference talks to learn opinions of the ``said'' experts. 
This decision-making process is largely ad hoc, lacks empirical foundations, and may or may not lead to a consensus in the field.
As a result, the field abounds with what I may call \textbf{software engineering myths}, which are \emph{widely held beliefs about best practices derived from intuition, experience, and conventional wisdom, that have never been rigorously tested or, when tested, have produced inconclusive or contradictory results}.

This state of affairs persists despite decades of empirical research in software engineering. 
The field has generated thousands of studies examining key software engineering topics, including developer productivity, code quality, security practices, and the effectiveness of developer tools. 
Yet, this substantial body of work has failed to transform software engineering into an evidence-based discipline.
The problem lies not in the quantity of research but in its methodological limitations, particularly the pervasive reliance on (a) descriptive and correlational evidence from software repository data, and (b) qualitative evidence synthesizing practitioner opinions.

For example, consider the following problem: Should software projects allow flexibility in the versions of their dependencies, and to what extent?
A descriptive mining software repository study~\cite{DBLP:conf/msr/0001PSTB19} may conclude that the Java/Maven ecosystem is extensively pinning dependency versions while the JavaScript/npm ecosystem is extensively using semantic versioning, without providing any concrete guidance or discussions of the trade-offs around them.
A qualitative study surveying JavaScript developers~\cite{DBLP:journals/tse/JafariCAST22} may conclude that semantic versioning should be preferred over pinned dependencies as the latter is a ``dependency smell'' likely to cause maintainability issues in the future.
Another qualitative study interviewing security experts~\cite{DBLP:conf/sp/LadisaPMB23} may conclude that dependency pinning should be preferred over semantic versioning, as the former is critical for preventing software supply chain attacks.
In this case, the quantitative repository mining study fails to provide evidence-based guidance, while the two qualitative studies present contradictory opinions from practitioner cohorts with different experiences and perspectives; neither would be sufficient to formulate best practices for this dependency versioning problem.

Similarly, consider the longstanding debate over whether the choice of programming language causally affects defect-proneness. 
A large-scale correlational study~\cite{DBLP:conf/sigsoft/RayPFD14, DBLP:journals/cacm/RayPDF17} analyzed millions of commits across thousands of GitHub projects; the study concluded that programming language has a significant but modest effect on defect-proneness, with functional and statically-typed languages associated with fewer defects.
However, a rigorous reproduction study~\cite{DBLP:journals/toplas/BergerHMVV19} revisiting the same dataset with improved methodology—correcting errors in language classification, accounting for polyglot repositories, and applying more appropriate statistical models—found that the original effect sizes shrank dramatically and many findings failed to replicate. 
The original authors issued a rebuttal~\cite{DBLP:journals/corr/abs-1911-07393} defending their methodology, which prompted a counter-rebuttal~\cite{DBLP:journals/corr/abs-1911-11894} from the reproduction team. 
Putting the controversy around the specific research design aside, the counter-rebuttal~\cite{DBLP:journals/corr/abs-1911-11894} reveals a deeper methodological impasse behind why the field cannot reach consensus on a seemingly straightforward empirical question:
\begin{quote}
\emph{I fundamentally do not believe that the question the FSE authors set out to answer, namely, whether some languages are more defect-prone than others, can be meaningfully answered by scraping GitHub...An analysis of GitHub is necessarily measuring correlation and not establishing causation, as it does not constitute an experiment.}
\end{quote}

The problem extends beyond individual studies to create a broader crisis of confidence in empirical software engineering research. 
When studies merely repeat and synthesize practitioner opinion, practitioners justifiably will not pay attention to the research findings.
Even worse, when studies reach contradictory conclusions and fail to rule out confounding factors or alternative explanations, practitioners will justifiably become skeptical of the research findings altogether. 
The result is a field that produces voluminous research but limited actionable knowledge.

Modern causal inference methods offer a principled approach to addressing this problem~\cite{cunningham2021causal, pearl2009causality, hansen2022econometrics}. 
Developed primarily in econometrics and epidemiology over the past several decades, these methods provide frameworks for distinguishing genuine causal effects from spurious correlations in observational data---precisely the kind of data investigated in repository mining studies.
Importantly, the value of causal inference methods lies not in any claim to methodological perfection, but rather in its insistence on \emph{transparency}: Every causal design rests on assumptions that must be explicitly stated, examined, and debated.
When a difference-in-differences study claims a causal effect, it does so under the assumption of parallel trends---an assumption that can be partially tested and openly scrutinized~\cite{marcus2021role}.
When an instrumental variables analysis identifies an effect, it does so under exclusion restrictions that critics can directly challenge~\cite{deaton2010instruments}.
This methodological transparency addresses the core problem illustrated by the programming language vs. defect-proneness example: 
When studies produce contested results, progress requires not endless reanalysis but a shared framework for evaluating whether the findings and claims are plausible given the specific research context and, if not, how to develop a better design.

In this thesis, I argue that \emph{empirical software engineering research needs an explicit framework and vocabulary for causal credibility assessments so that it can better dispel software engineering myths and inform evidence-based practice}.
This pragmatic stance recognizes that causal inference in observational settings is inherently imperfect: No study eliminates all confounding, no natural experiment is as clean as a randomized controlled trial, and no finding generalizes to all contexts without qualification.
Yet, these limitations do not render causal inquiry futile.
Rather, they demand intellectual honesty about what we can and cannot conclude using available data, and careful calibration of confidence to convey the strength of the evidence to the audience.
Specifically, to assess causal credibility, one should: (1) make the causal theory explicit as directed acyclic graphs (DAGs), (2) define the causal estimate of interest through the potential outcome framework and specify the assumption under which the estimate would carry a causal interpretation, (3) honestly acknowledge and mitigate limitations in data collection, metric design, and model specifications, and (4) thoughtfully navigate alternative explanations (i.e,, alternative DAGs) that may also be supported with current results. 

This explicit framework and vocabulary are particularly important for research challenging well-known software engineering myths.
From a Bayesian perspective, if a study yields results contrary to practitioners' intuitions, it requires more compelling evidence to shift posterior beliefs than a study that merely confirms conventional wisdom.\footnote{This reflects the Bayesian epistemological principle often summarized as ``extraordinary claims require extraordinary evidence.'' When a claim has a low prior probability (i.e., it contradicts established beliefs or practitioner intuition), stronger evidence is needed to shift the posterior belief toward acceptance. Formally, if $P(H)$ is low, then $P(E|H)/P(E|\neg H)$---the likelihood ratio---must be correspondingly high for $P(H|E)$ to reach credible levels. For a rigorous treatment of Bayesian reasoning in scientific inference, see Howson and Urbach~\cite{howson2006scientific}.}
The value of this framework lies not in its capacity to provide causal evidence of epistemological certainty, but through the transparent articulation of theory and assumptions, the honest acknowledgment and mitigation of limitations, and the careful accumulation of consistent findings across complementary designs.
The three empirical studies in this thesis (Chapter~\ref{chp:pinning}, \ref{chp:fake-stars}, \ref{chp:cursor}) exemplify this approach: Each challenges a widely held belief and explicitly assesses the credibility of its causal claims---through which each provides stronger evidence on software engineering myths than a descriptive or correlational study without any explicit assessments of causal credibility.

\section{Thesis Statement}

\begin{framed}
Software engineering practice is characterized by \emph{myths}---widely held beliefs about best practices derived from intuition, experience, and conventional wisdom that have never been rigorously tested or, when tested, have produced inconclusive results.
This thesis argues that \emph{empirical software engineering research needs an explicit framework
and vocabulary for causal credibility assessments so that it can better dispel software engineering myths and inform evidence-based practice}.
Specifically, to assess causal credibility, one should: 
\begin{enumerate}[leftmargin=20pt]
    \item Make the causal theory explicit as directed acyclic graphs (DAGs), 
    \item Define the causal estimate of interest through the potential outcome framework and specify the assumption under which the estimate would carry a causal interpretation,
    \item Honestly acknowledge and mitigate limitations in data collection, metric design, and model specifications, 
    \item Thoughtfully navigate alternative explanations (i.e, alternative DAGs) that may also be supported with current results. 
\end{enumerate}

\end{framed}

\section{Organization and Contributions}

In this section, I summarize the organization and main contributions of this thesis.

\paragraph{A Primer on Causal Credibility (Chapter~\ref{chp:background})}
The thesis starts with a self-contained introduction to causal credibility assessments for software engineering researchers. 
The chapter begins by articulating the fundamental problem of causal inference---the impossibility of observing both potential outcomes for a single unit---and introduces three conceptual frameworks that have shaped modern causal reasoning: the potential outcomes framework, graphic causal models, and the design-based approach.
It then examines common pitfalls in descriptive and correlational research, explaining why correlation does not imply causation, why multivariate regression alone cannot establish causal claims, and when regression can support causal claims under explicitly stated assumptions.
After formalizing the key concepts through the potential outcomes framework and graphical causal models, the chapter concludes with a pragmatic approach to causal credibility assessments in empirical software engineering research---operationalizing the four-step framework articulated in the thesis statement, discussing the internal--external validity trade-off, and presenting a hierarchy of evidence for situating causal claims.

\paragraph{Three Cases of Empirical Studies}
Then, the thesis dives deep into three empirical studies I published in the past that challenged well-known software engineering myths through a pragmatic causal inference with explicit causal credibility assessments---each study making theory, estimand, and assumptions explicit, acknowledging limitations honestly, and navigating alternative explanations thoughtfully~\cite{DBLP:journals/corr/abs-2502-06662, DBLP:journals/corr/abs-2412-13459, DBLP:journals/corr/abs-2511-04427}.

\begin{itemize}[leftmargin=15pt]
    \item \textbf{The Impact of Dependency Pinning on Supply Chain Security (Chapter~\ref{chp:pinning})}: Recent high-profile incidents in open-source software have greatly increased practitioner attention on software supply chain attacks. 
    To guard against potential malicious package updates, security practitioners advocate \emph{pinning} dependency to specific versions rather than \emph{floating} in version ranges.
    However, it remains controversial whether pinning provides a meaningful security benefit that outweighs the cost of maintaining outdated, possibly vulnerable dependencies.
    In this chapter, I quantify, through counterfactual analysis and simulations, the security and maintenance impact of version constraints in the npm ecosystem. 
    By simulating dependency resolutions over historical time points, I find that pinning direct dependencies not only (as expected) increases the cost of maintaining vulnerable and outdated dependencies, but also (surprisingly) even increases the risk of exposure to malicious package updates in larger dependency graphs due to the specifics of npm's dependency resolution mechanism. 
    Finally, I explore collective pinning strategies to secure the ecosystem against supply chain attacks and suggest specific changes to npm to enable these interventions.
    These findings provide guidance to practitioners and tool designers on how to manage their supply chains more securely.
    \item  \textbf{The Impact of Fake GitHub Stars on Repository Promotion (Chapter~\ref{chp:fake-stars})}: GitHub, the de facto platform for open-source software development, provides a set of social-media-like features to signal high-quality repositories.
    Among them, the star count is the most widely used popularity signal, but it is also at risk of being artificially inflated (i.e., \emph{faked}), decreasing its value as a decision-making signal and posing a security risk to all GitHub users.
    In this chapter, I present a systematic, global, and longitudinal measurement study of fake stars in GitHub.
    To this end, I built \SystemName, a scalable tool able to detect anomalous starring behaviors across all GitHub metadata between 2019 and 2024.
    Analyzing the data collected using \SystemName, I find that: (1) fake-star-related activities have rapidly surged in 2024; (2) the accounts and repositories in fake star campaigns have highly trivial activity patterns; (3) the majority of fake stars are used to promote short-lived phishing malware repositories; the remaining ones are mostly used to promote AI/LLM, blockchain, tool/application, and tutorial/demo repositories; (4) while repositories may have acquired fake stars for growth hacking, fake stars only have a promotion effect in the short term (i.e., less than two months) and become a liability in the long term.
    This study has implications for platform moderators, open-source practitioners, and supply chain security researchers.
    \item \textbf{The Impact of Cursor AI on Open-Source Projects (Chapter~\ref{chp:cursor})}:   Large language models (LLMs) have demonstrated the promise to revolutionize the field of software engineering.
    Among other things, LLM agents are rapidly gaining momentum in software development, with practitioners reporting a multifold increase in productivity after adoption.
    Yet, empirical evidence is lacking around these claims.
    In this chapter, I estimate the \emph{causal} effect of adopting a widely popular LLM agent assistant, namely Cursor, on \emph{development velocity} and \emph{software quality}.
    The estimation is enabled by a state-of-the-art difference-in-differences design comparing Cursor-adopting GitHub projects with a matched control group of similar GitHub projects that do not use Cursor. 
    I find that Cursor adoption leads to a significant, large, but transient increase in project-level development velocity, along with a substantial and persistent increase in static analysis warnings and code complexity.
    Further panel generalized method of moments estimation reveals that the increase in static analysis warnings and code complexity acts as a major factor causing long-term velocity slowdown.
    This study carries implications for software engineering practitioners, LLM agent assistant designers, and researchers. 
\end{itemize}

\paragraph{Proposed Work (Chapter~\ref{chp:proposed-work})}
The proposed work extends the thesis by demonstrating a complementary use of the causal credibility assessment framework: systematically assessing \emph{existing} empirical research on a fundamentally causal question to clarify what the evidence supports, where its limitations lie, and what specific study designs would yield stronger evidence.
To this end, I revisit the longstanding debate on programming languages and defect proneness~\cite{DBLP:conf/sigsoft/RayPFD14, DBLP:journals/cacm/RayPDF17, DBLP:journals/toplas/BergerHMVV19}---a contested question that has resisted consensus despite extensive prior research.
I apply the four-step framework to the debate surrounding the seminal study by \citet{DBLP:conf/sigsoft/RayPFD14}, showing that when studies investigate causal questions using correlational methods without explicit causal reasoning---even when findings are carefully framed as correlational, as they were in Ray et al.---the resulting ambiguity invites confusion and unproductive controversy.
Informed by this assessment, I propose two complementary empirical studies: a panel fixed effects analysis exploiting within-developer and within-repository variation, and a difference-in-differences study of JavaScript-to-TypeScript migration events.
Together, these contributions illustrate how the framework can serve as both a diagnostic tool for evaluating existing research and a design guide for future studies, charting a principled path from identified limitations to stronger evidence.

\paragraph{Summary of Contributions} This thesis makes the following contributions:

\begin{itemize}[leftmargin=15pt]
   \item \textbf{Conceptual Contribution:} I introduce the notion of \emph{software engineering myths}---widely held beliefs about best practices that lack rigorous empirical validation---and argue that the prevalence of such myths stems from the field's over-reliance on descriptive evidence and inability to draw causal claims. I advocate for a \emph{pragmatic causal inference approach with explicit causal credibility assessments} that prioritizes transparency over perfection, with: (1) explicit definition of causal theory, potential outcomes, and assumptions for a causal interpretation, (2) honest acknowledgment and mitigation of methodological limitations, and (3) thoughtful navigation of alternative explanations. This conceptual framing provides a path for incremental accumulation of trustworthy evidence for software engineering practitioners, a lens for critically evaluating existing empirical software engineering research, and a principled guideline for future empirical software engineering research that would reach the same level of rigor as econometrics and epidemiology research today.
    
    \item \textbf{Methodological Contributions:} I demonstrate the practical application of causal inference methods to software engineering problems, with particular attention to the assumptions, limitations, and trade-offs involved in each design. The methods applied include: (a) counterfactual simulation for analyzing the effect of a practice where controlled experiments are infeasible but simulation is feasible; (b) fixed effect and panel regressions for controlling omitted variable bias from time- and subject-invariant characteristics; (c) difference-in-differences estimation with staggered adoption for evaluating the impact of developer tool adoption on software quality; (d) propensity score matching for constructing comparable treatment and control groups from observational repository data; and (e) panel generalized method of moments for disentangling bidirectional causal relationships. For each method, I showcase how to conduct explict casual credibility assessments as articulated above in each unique study setting.
    
    \item \textbf{Empirical Contributions:} I provide novel empirical evidence challenging three widely held software engineering myths: (a) that dependency pinning improves project security (Chapter~3), when in fact it paradoxically increases attack exposure in transitive dependency graphs under current ecosystem practices; (b) that artificially inflated GitHub star counts provide sustainable promotional benefits (Chapter~4), when in fact they become long-term liabilities; and (c) that AI coding assistants unambiguously improve developer productivity (Chapter~5), when in fact the initial velocity gains are offset by persistent technical debt accumulation. 
\end{itemize}