\section{The Causal Inference Problem}

The quest for rigorous causal reasoning traces back to the pioneering work of John Snow during the 1854 London cholera outbreak.
At a time when the prevailing ``miasma theory'' attributed disease to bad air, Snow meticulously mapped cholera deaths and traced them to a contaminated water pump on Broad Street.
By comparing mortality rates between households served by different water companies---one drawing from a polluted section of the Thames, the other from a cleaner source---he demonstrated a causal link between contaminated water and cholera, effectively conducting one of the first natural experiments in history~\cite{snow1855mode}.
Snow's approach foreshadowed the counterfactual thinking central to modern causal inference: What would have happened to these households had they received clean water instead?
Building upon such empirical foundations, the 20th century witnessed the formalization of causal reasoning.
Jerzy Neyman introduced the potential outcomes notation in 1923 for agricultural experiments~\cite{neyman1923application}, while Ronald Fisher developed the theory of randomized controlled trials~\cite{fisher1971design}.
Decades later, Donald Rubin extended the potential outcomes framework to observational studies~\cite{rubin1974estimating}, establishing what is now known as the Rubin Causal Model---the mathematical backbone of modern causal inference.

Armed with these rigorous frameworks, causal inference has transformed decision-making across virtually every domain of human inquiry.
In medicine, randomized controlled trials became the gold standard for drug approval, ushering in the era of evidence-based medicine that has saved countless lives~\cite{sackett1996evidence}.
In economics, causal methods revolutionized policy evaluation: Card and Krueger's difference-in-differences study on minimum wage~\cite{card1993minimum} overturned decades of conventional wisdom, while Angrist and Imbens formalized instrumental variable methods for extracting causal effects from observational data~\cite{angrist1996identification}.
The field's crowning recognition came in 2021, when the Nobel Prize in Economics was awarded jointly to David Card, Joshua Angrist, and Guido Imbens \emph{``for their methodological contributions to the analysis of causal relationships''}---a testament to how profoundly these methods have reshaped empirical research~\cite{nobel2021economics}.
Beyond academia, tech giants now run millions of A/B tests annually to causally evaluate product changes~\cite{kohavi2020trustworthy}, while public health agencies relied on causal inference frameworks to rapidly evaluate COVID-19 vaccine efficacy during the pandemic~\cite{polack2020safety}.
With research ranging from policy-making to medicine to application user experience, the ability to distinguish causation from correlation has become indispensable to modern science and industry alike.

Conceptually, causation can be defined through counterfactuals: $X$ causes $Y$ if $Y$ would have been different had $X$ not occurred.
Yet, we can never observe both realities for the same unit---a patient either receives a drug or does not; a developer either adopts a tool or does not; a project uses one programming language or another.
This ``fundamental problem of causal inference,'' that individual causal effects are inherently unobservable, means that causal claims can never be directly verified.
Different intellectual traditions have developed distinct approaches to navigate this challenge.
Specifically, three conceptual frameworks have shaped modern causal inference, each offering a distinct lens for reasoning about cause and effect:
\begin{itemize}[leftmargin=15pt]
    \item \textbf{The Potential Outcomes Framework:} Pioneered by Neyman in 1923 and formalized by Rubin~\cite{rubin1974estimating}, this framework defines causation through counterfactual comparisons. 
    For each unit $i$, there exist two potential outcomes: $Y_i(1)$ (outcome if treated) and $Y_i(0)$ (outcome if not treated). 
    The individual causal effect is $Y_i(1) - Y_i(0)$, but since only one outcome is ever observed, the framework focuses on estimating population-level quantities like the Average Treatment Effect $ATE = E[Y(1) - Y(0)]$.
    The central question becomes: Under what conditions can we use observed data to estimate these counterfactual quantities?
    The framework formalizes assumptions such as \emph{ignorability} (treatment assignment is independent of potential outcomes given covariates) and \emph{overlap} (all units have a positive probability of receiving either treatment) to justify causal estimation.
    Under this framework, \emph{Randomized controlled trials} are the golden standard: Random assignment ensures ignorability by design, providing uncontested identification.
    When randomization is infeasible, selection-on-observables methods---\emph{propensity score matching}, \emph{inverse probability weighting}, and \emph{doubly robust estimation}---attempt to achieve ignorability by conditioning on observed covariates.
    
    \item \textbf{The Structural Causal Model Framework:} Developed by Pearl~\cite{pearl2009causality}, this framework represents causal relationships as directed acyclic graphs (DAGs), where nodes are variables and edges represent direct causal effects.
    Causation is formalized through the ``do-operator'': $P(Y | do(X=x))$ represents the distribution of $Y$ when $X$ is externally set to $x$, distinct from the observational conditional $P(Y | X=x)$ which may reflect confounding.
    The framework provides graphical criteria---the back-door criterion, front-door criterion, and do-calculus---for determining when causal effects are \emph{identifiable} from observational data and which variables must be adjusted for.
    Its strength lies in making causal assumptions visually explicit: A DAG encodes what the researcher believes about the data-generating process, and critics can directly challenge specific edges or missing paths.
    Methods from this framework include \emph{covariate adjustment via the back-door criterion}, which identifies the minimal set of variables to condition on for blocking confounding paths; \emph{front-door adjustment}, which identifies causal effects through mediating variables when direct adjustment is impossible; \emph{instrumental variables} justified through graphical analysis of exogeneity; and \emph{mediation analysis}, which decomposes total effects into direct and indirect pathways using the causal graph structure.
    
    \item \textbf{The Design-Based (Credibility Revolution) Approach:} Championed by applied economists like Angrist and Imbens~\cite{angrist1996identification}, this approach is less a formal framework and more a philosophy of empirical practice.
    Rather than starting from abstract potential outcomes or graphical models, it asks: What feature of the real world generates plausibly exogenous variation in treatment?
    The emphasis is on finding \emph{natural experiments}---situations where nature, policy, or institutional rules create variation that approximates random assignment---and on transparently arguing why alternative explanations are implausible.
    A study's credibility rests not on mathematical formalism but on whether the identification strategy would convince a skeptical reader that confounding has been adequately addressed.
    Methods championed by this tradition include \emph{difference-in-differences}, which exploits differential timing of policy changes to compare treatment and control trajectories; \emph{regression discontinuity designs}, which leverage sharp eligibility thresholds that create quasi-random assignment near the cutoff; \emph{instrumental variables} motivated by institutional knowledge of exogenous shocks; \emph{synthetic control methods}, which construct counterfactual units as weighted combinations of untreated donors; and \emph{event studies}, which trace out dynamic treatment effects before and after intervention to assess parallel trends and effect persistence.
\end{itemize}
While these frameworks differ in formalism and emphasis, they share a common commitment: \emph{Making causal assumptions explicit and assessable so they can be examined and debated, rather than left implicit and buried in the underlying statistical models.}
Critics can then interrogate whether that assumption is plausible given the specific research context, rather than debating vague notions of ``correlation versus causation'' or beliefs of whether the findings ``seem valid.''
Based on different assumptions, these frameworks derive a family of \emph{identification methods}: Research designs that, under their stated assumptions, provide a complete path from observed data to causal effects.
Alongside these identification methods, researchers also frequently employ \emph{control methods}---techniques like fixed effects and covariate adjustment that address specific sources of confounding without claiming complete identification.\footnote{Propensity score matching occupies an intermediate position: In principle, it identifies causal effects if all confounders are observed (the ``selection on observables'' assumption), but because this assumption is untestable and often implausible, many practitioners treat matching as strengthening---rather than guaranteeing---causal credibility.}

To illustrate the promises and challenges of causal reasoning in software engineering, consider a deceptively simple question: Does the choice of programming language causally affect defect-proneness?
A large-scale study~\cite{DBLP:conf/sigsoft/RayPFD14, DBLP:journals/cacm/RayPDF17} analyzed millions of commits across thousands of GitHub projects and reported that functional and statically-typed languages are associated with fewer defects.
A subsequent reproduction study~\cite{DBLP:journals/toplas/BergerHMVV19}, however, found that effect sizes shrank dramatically after correcting methodological issues, sparking a series of rebuttals~\cite{DBLP:journals/corr/abs-1911-07393, DBLP:journals/corr/abs-1911-11894}.
The core difficulty lies in the nature of the data: Projects written in different languages differ systematically in ways that correlate with defect rates---team expertise, project maturity, application domain, and coding conventions all vary with language choice.
Observing that Haskell projects have fewer bugs than PHP projects tells us little about what would happen if the same team rewrote the same application in a different language.
This reflects a fundamental challenge in causal inference, \emph{confounding}: Observed associations may reflect selection effects rather than causal mechanisms.
Causal inference methods do not magically resolve this problem or provide definitive answers where none exist.
What they offer instead is a disciplined framework for assessing causal validity under explicitly stated assumptions.
Rather than claiming an effect is ``proven,'' causal designs articulate the conditions under which an estimate would be unbiased---parallel trends, conditional independence, exclusion restrictions---and provide tools to probe whether those conditions plausibly hold.
This transparency transforms debates from disputes over contested conclusions into examinations of clearly specified assumptions, enabling the research community to systematically evaluate where evidence is credible and where it remains fragile.

Importantly, not all empirical problems admit equally strong causal designs, and different methods offer varying degrees of credibility.
At one end of the spectrum, randomized controlled trials eliminate confounding by design, providing the strongest foundation for causal claims.
At the other end, naive correlational analyses make no attempt to address confounding and offer little causal credibility.
Between these extremes lies a hierarchy of quasi-experimental methods: Regression discontinuity designs exploit sharp thresholds that create local randomization; instrumental variables isolate exogenous variation when valid instruments exist; difference-in-differences leverages parallel trends assumptions that can be partially tested.
Fixed effects models occupy a middle position---they credibly control for time-invariant confounders but remain vulnerable to time-varying confounding and offer no built-in falsification test.

On the other hand, higher credibility does not always mean better evidence.
Randomized experiments, while eliminating all sources of confounding, may be impractical, unethical, or---crucially---may alter the setting so fundamentally that findings fail to generalize.
For example, a controlled lab experiment measuring how developers use a new tool may achieve perfect internal validity but tell us little about adoption in the messy reality of production environments with deadlines, legacy code, and team dynamics.
Conversely, observational studies of real-world behavior preserve ecological validity but much more challenging to rule out confounders and alternative explanations.
This creates an inherent trade-off: Designs that maximize internal validity often sacrifice external validity, and vice versa.
No single study, however well-designed, can definitively establish a causal claim across all contexts.
Instead, confidence in a causal relationship grows through \emph{triangulation}---when multiple studies using different designs, data sources, and populations converge on consistent findings.
A treatment effect estimated via difference-in-differences in one setting, replicated with regression discontinuity in another, and corroborated by a field experiment in a third, provides far stronger evidence than any single study alone.
The credibility of a causal theory ultimately rests not on any one method but on the accumulated weight of evidence across complementary approaches.

The remainder of this chapter develops the technical foundations for causal inference in software engineering research.
Section~\ref{sec:pitfalls} examines common pitfalls in descriptive and correlational research, explaining why correlation does not imply causation and why multivariate regression alone cannot establish causal claims.
Section~\ref{sec:formalism} formalizes the key concepts, presenting the potential outcomes framework and graphical causal models that underpin modern causal inference.
Section~\ref{sec:identification} surveys identification strategies---randomized controlled trials, difference-in-differences, regression discontinuity, and instrumental variables---that provide complete causal identification under stated assumptions.
Section~\ref{sec:control} discusses control strategies---matching, fixed effects, and generalized method of moments---that address specific sources of confounding without providing complete identification.
Section~\ref{sec:applicability} considers the applicability of these methods to software engineering research, addressing domain-specific challenges in data, measurement, and generalizability.
Finally, Section~\ref{sec:threats} discusses threats to validity in causal designs and the role of sensitivity analysis in strengthening credibility.

\section{Pitfalls in Descriptive and Correlational Research}
\label{sec:pitfalls}

Before presenting the causal inference toolkit, it is instructive to examine why naive approaches fail.
Understanding these pitfalls not only clarifies what causal methods must overcome but also reveals why the solutions are non-trivial---why simply ``being careful'' or ``controlling for more variables'' cannot substitute for genuine causal reasoning.

\subsection{Why Correlation Does Not Imply Causation}

The dictum ``correlation does not imply causation'' is perhaps the most frequently cited methodological principle in empirical research, yet it remains the most frequently violated.
The violation persists not because researchers are careless, but because the underlying mechanisms that generate spurious correlations are subtle and pervasive.
Let us examine these mechanisms through the lens of our programming language example.

Recall the question: Does choice of programming language causally affect defect-proneness?
Suppose we compute the correlation between language type (say, functional vs. imperative) and defect density across thousands of GitHub projects and find that functional language projects have 20\% fewer defects.
Can we conclude that switching to a functional language would reduce defects by 20\%?
The answer is almost certainly no, and understanding why requires distinguishing three mechanisms that can generate spurious correlations.

\paragraph{Confounding: The Hidden Common Cause.}
Confounding occurs when a third variable $Z$ causally affects both the treatment $X$ and the outcome $Y$, inducing an association between them even when no direct causal relationship exists.
In our programming language example, consider \emph{developer expertise}.
It is very likely that experienced developers are both more likely to choose functional languages (perhaps because these require more sophisticated programming knowledge) and more likely to write low-defect code (because expertise directly reduces errors).
If the above mechanism were true, the observed correlation between functional languages and fewer defects may simply reflect this selection effect: We are comparing code written by different types of developers, not the causal effect of the language itself.

To see this mathematically, let $Y$ denote defect density, $X$ denote language choice (1 for functional, 0 for imperative), and $Z$ denote developer expertise.
The true data-generating process might be:
\begin{align}
    X &= f(Z, U_X) \quad \text{(language choice depends on expertise)} \\
    Y &= g(Z, U_Y) \quad \text{(defects depend on expertise, not language)}
\end{align}
where $U_X$ and $U_Y$ are independent error terms.
Even though $X$ has no causal effect on $Y$, the correlation $\text{Cor}(X, Y) \neq 0$ because both are driven by the common cause $Z$.
More precisely, decomposing the observed association:
\begin{equation}
    E[Y | X=1] - E[Y | X=0] = \underbrace{\tau}_{\text{causal effect}} + \underbrace{E[Z | X=1] - E[Z | X=0]}_{\text{selection difference}} \cdot \underbrace{\frac{\partial Y}{\partial Z}}_{\text{effect of $Z$}}
\end{equation}
If the causal effect $\tau = 0$ but functional language adopters have higher expertise ($E[Z|X=1] > E[Z|X=0]$) and expertise reduces defects ($\partial Y / \partial Z < 0$), we observe a negative correlation between functional languages and defects---but it is a pure artifact of selection.

\paragraph{Reverse Causality: Getting the Arrow Backwards.}
Reverse causality occurs when the observed association reflects $Y$ causing $X$ rather than $X$ causing $Y$.
In our example, perhaps teams whose projects have persistent quality problems switch to more ``rigorous'' languages like Haskell in an attempt to impose discipline.
If this were true, we might observe that Haskell projects have \emph{more} defects---not because Haskell causes defects, but because defect-prone teams select into Haskell.
Alternatively, successful low-defect projects may attract functional programming enthusiasts who refactor the codebase, creating the opposite spurious correlation.

The fundamental problem is that cross-sectional data cannot distinguish between $X \to Y$ and $Y \to X$.
Both causal structures generate the same correlation.
Only temporal information (did language choice precede defect reduction?) combined with careful reasoning about the selection process can help disentangle these possibilities.

\paragraph{Selection Bias: Conditioning on a Consequence.}
Selection bias arises when the sample is not representative of the population of interest, or when the data-generating process induces spurious associations through conditioning on a \emph{collider}---a variable caused by both $X$ and $Y$.

Consider an extreme example: Suppose we analyze only ``successful'' GitHub projects (those with more than 1,000 stars).
A project becomes successful if it either (a) solves a valuable problem despite bugs, or (b) has exceptional code quality.
Among successful projects, we might observe a \emph{negative} correlation between functional languages and defects: Successful functional projects may have achieved success through quality, while successful imperative projects may have succeeded despite bugs by solving critical problems.
This correlation is entirely an artifact of conditioning on success---it tells us nothing about the causal effect of language choice.

More subtly, analyzing only projects that ``survived'' to a certain age conditions on a post-treatment variable (survival), potentially inducing bias.
If functional language projects fail faster when they have bugs (because functional programming communities have higher quality standards), the surviving functional projects are a selected sample, and comparisons with surviving imperative projects are confounded.

\subsection{Why Multivariate Regression (Usually) Does Not Suffice}

Confronted with confounding, researchers often reach for what seems like an obvious solution: ``control for'' the confounders using multivariate regression.
If developer expertise confounds the language-defect relationship, why not simply add expertise as a covariate?
While intuitive, this approach has fundamental limitations that render it insufficient for causal inference.

\paragraph{Omitted Variable Bias: The Inescapable Problem of Unobserved Confounders.}
The most fundamental limitation is that regression can only adjust for \emph{observed} confounders.
In practice, important confounders are often unmeasured or unmeasurable.
Returning to our example, consider the confounders that plausibly affect both language choice and defect rates: Developer expertise, yes, but also team culture, project complexity, application domain, code review practices, testing discipline, deadline pressure, and organizational incentives.
Can we measure all of these from repository data?
Certainly not.

The mathematical consequence is stark.
Suppose the true model is:
\begin{equation}
    Y_i = \alpha + \tau X_i + \beta Z_i + \gamma U_i + \epsilon_i
\end{equation}
where $Z$ is an observed confounder and $U$ is an unobserved confounder.
If we regress $Y$ on $X$ and $Z$ only, the estimated coefficient on $X$ is:
\begin{equation}
    \hat{\tau} \xrightarrow{p} \tau + \gamma \cdot \frac{\text{Cov}(X, U | Z)}{\text{Var}(X | Z)}
\end{equation}
The bias term $\gamma \cdot \text{Cov}(X, U | Z) / \text{Var}(X | Z)$ depends on the relationship between treatment and the omitted confounder \emph{after} controlling for $Z$.
Unless $\text{Cov}(X, U | Z) = 0$---which would require $Z$ to fully mediate the relationship between $X$ and $U$---the estimate remains biased.
Adding more controls helps only if those controls are themselves confounders; adding irrelevant controls simply reduces precision without reducing bias.

The uncomfortable implication is that \emph{no amount of control variables can substitute for a credible identification strategy}.
A regression with 50 covariates is no more causally credible than one with 5 if the key confounders remain unobserved.
The number of controls is irrelevant; what matters is whether the remaining variation in treatment is as-if random.

\paragraph{Bad Controls: When Adjustment Makes Things Worse.}
A subtler problem is that not all variables \emph{should} be controlled for, even if they are observed.
Including the wrong controls can introduce bias where none existed, or amplify existing bias.
Two cases are particularly important:

\begin{itemize}[leftmargin=15pt]
    \item \emph{Mediator bias.}
    A mediator is a variable on the causal pathway from treatment to outcome ($X \to M \to Y$).
    Controlling for a mediator blocks part of the causal effect and biases the estimate toward zero.
    In our example, suppose functional languages reduce defects partly by encouraging immutability, which in turn prevents certain bug classes.
    If we ``control for'' use of immutable data structures, we block this pathway and underestimate the true effect of language choice on defect proneness.
    \item \emph{Collider bias.}
    A collider is a variable caused by both treatment and outcome ($X \to C \leftarrow Y$).
    Conditioning on a collider \emph{creates} spurious associations between its causes, even when they are independent.
    Suppose a project's ``popularity'' depends on both the novelty of its language choice and its quality.
    Controlling for popularity induces a negative correlation between language novelty and quality among popular projects: If we know a project is popular, learning it uses a novel language makes us infer it must have lower quality (otherwise, why would quality alone not suffice for popularity?).
    This is a pure statistical artifact.
\end{itemize}
The lesson is that without a causal model specifying the data-generating process, researchers cannot determine which controls are appropriate.
Regression software will happily estimate coefficients regardless of whether the specification is causally meaningful.

\paragraph{Functional Form Misspecification.}
An even subtler problem for multivariate regressions is that most standard regressions assume linear, additive relationships.
If the true effect is nonlinear, if treatment effects are heterogeneous, or if confounders interact with treatment, the regression coefficient may not correspond to any meaningful causal quantity.
In our example, the effect of language choice may depend on project size, team composition, or application domain.
A single coefficient averaging across all contexts may be misleading for any particular context.

\paragraph{The Table 2 Fallacy: Misinterpreting ``Controlled'' Estimates.}
These problems manifest in a pervasive interpretive error that Westreich and Greenland term the ``Table 2 Fallacy''~\cite{westreich2013table}: The belief that if a coefficient ``remains significant after controlling for $Z$,'' the association must be causal.
This interpretation fails for multiple reasons:
\begin{itemize}[leftmargin=15pt]
    \item If $Z$ is a mediator, the controlled estimate underestimates the true effect.
    \item If $Z$ is a collider, conditioning on it introduces spurious association.
    \item If $Z$ is a confounder but imperfectly measured, residual confounding persists.
    \item If important confounders remain omitted, the estimate is still biased regardless of what was included.
\end{itemize}
The interpretation of ``controlling for $Z$'' depends entirely on the causal structure relating $X$, $Y$, and $Z$---a structure that regression alone cannot reveal.

\subsection{When Can Multivariate Regression Support Causal Claims}

The preceding discussion may leave the impression that multivariate regression is never useful for causal inference.
This would be an overstatement---and an impractical one.
The formal condition under which regression identifies causal effects is \emph{selection on observables} (also called conditional ignorability): $(Y(1), Y(0)) \perp D \mid X$, meaning that conditional on observed covariates, treatment is as-if randomly assigned.
This requires that all confounders be measured and correctly included, that no post-treatment variables contaminate the model, and that there is sufficient overlap between the treated and control groups.
Methods including propensity score matching, inverse probability weighting, and regression adjustment all rest on this same assumption---they differ only in implementation, not in the fundamental identification claim.

\paragraph{The Role of Theory.}
This is where causal inference rejoins substantive domain knowledge.
If a researcher can articulate a complete causal model---a DAG representing all relevant variables and their relationships---and the regression specification correctly implements the adjustment set dictated by that DAG, then the coefficient \emph{does} have a causal interpretation.
The regression is not blindly ``controlling for stuff''; it is operationalizing a theoretical claim about the data-generating process.
The credibility of the causal estimate then rests entirely on the credibility of the underlying theory: Have we correctly identified all common causes of treatment and outcome?
Have we avoided conditioning on mediators and colliders?
Is the proposed causal structure consistent with domain knowledge and prior evidence?

In this sense, \emph{all} causal inference from observational data is theory-dependent.
Design-based methods (difference-in-differences, regression discontinuity, instrumental variables) appear more credible not because they are assumption-free, but because their assumptions---parallel trends, continuity at the threshold, exclusion restrictions---are often more transparent and partially testable.
Selection-on-observables makes a stronger, less testable claim: That we have measured everything that matters.
But ``stronger'' does not mean ``never justified.''

\paragraph{A Pragmatic Perspective.}
If we insisted on ironclad causal identification for every empirical claim, much of applied research would grind to a halt.
Medicine advanced for centuries on observational evidence before randomized trials became standard.
Economics developed theories of market behavior, labor supply, and human capital from correlational data long before the credibility revolution.
Software engineering cannot always wait for perfect causally valid designs---practitioners need actionable insights now, even if those insights come with uncertainty.

The honest position is not that regression-based causal claims are invalid, but that they occupy a different place on the credibility spectrum.
A well-theorized regression study---one that explicitly articulates its causal model, defends the completeness of its adjustment set, addresses obvious threats, and conducts sensitivity analysis (e.g., Rosenbaum bounds~\cite{rosenbaum2002observational}, $E$-values~\cite{vanderweele2017sensitivity}, or coefficient stability tests~\cite{oster2019unobservable})---provides weaker but still informative evidence.
The key is transparency: Rather than claiming an effect is ``proven,'' the researcher specifies the conditions under which the estimate would be unbiased and invites scrutiny of whether those conditions hold.
Readers can then calibrate their confidence accordingly.
The danger lies not in using regression for causal questions, but in doing so naively.
The goal of this thesis is not to dismiss all prior work based on multivariate regressions, but to articulate clearly the assumptions underlying different methods, so that researchers can choose designs appropriate to their questions and honestly communicate the strength of their evidence.

\section{Formalizing Causal Inference}
\label{sec:formalism}

Having motivated the need for rigorous causal reasoning through the pitfalls of naive approaches, we now turn to the formal foundations.
The goal of formalization is not mathematical abstraction for its own sake, but precision: To state exactly what we mean by a ``causal effect,'' to articulate the assumptions required to estimate it, and to derive the logical consequences of those assumptions.
Without this precision, debates about causation devolve into disputes over intuition; with it, we can identify exactly where disagreements lie and what evidence would resolve them.

It is important to distinguish between \emph{frameworks} and \emph{methods}.
The frameworks presented in this section---potential outcomes and graphical causal models---are conceptual languages for reasoning about causation.
They define what causal effects mean, formalize the assumptions needed to identify them, and provide tools for determining whether identification is possible.
However, they are not themselves statistical procedures that one applies to data.
The \emph{methods} presented in subsequent sections---randomized experiments, difference-in-differences, regression discontinuity, instrumental variables---are concrete research designs and estimation strategies.
Each method makes specific assumptions that can be understood within either framework; the frameworks provide the theoretical foundation that justifies why the methods work when their assumptions hold.

Two complementary formal frameworks dominate modern causal inference.
The \emph{potential outcomes framework}, developed by Neyman~\cite{neyman1923application} and formalized by Rubin~\cite{rubin1974estimating}, defines causation through counterfactual comparison: What would have happened under alternative treatments?
The \emph{graphical causal model framework}, developed by Pearl~\cite{pearl2009causality}, represents causal relationships as directed graphs and provides algorithmic tools for determining when causal effects can be estimated from observational data.
Though they differ in their choice of formalism and emphasis, the two frameworks are mathematically compatible and often complementary.
The choice between them is largely one of convenience and communication: Potential outcomes excel at defining estimands and clarifying what randomization achieves; graphical models excel at encoding assumptions visually and deriving adjustment strategies.

\subsection{The Potential Outcomes Framework}

The potential outcomes framework begins with a simple question: What does it mean to say that $X$ causes $Y$?
The answer, formalized through counterfactual reasoning, is that $X$ causes $Y$ for a particular unit if $Y$ would have been different had $X$ been different, holding all else constant.
This definition transforms causation from a vague philosophical concept into a precise comparison between two states of the world---one in which treatment occurs and one in which it does not.

\paragraph{Notation and Setup.}
Consider a population of units indexed by $i = 1, \ldots, N$.
Each unit is characterized by a treatment indicator $D_i \in \{0, 1\}$, where $D_i = 1$ indicates that unit $i$ received treatment and $D_i = 0$ indicates control.
The key conceptual move is to posit that each unit has \emph{two} potential outcomes:
\begin{itemize}[leftmargin=15pt]
    \item $Y_i(1)$: The outcome that \emph{would} obtain if unit $i$ receives treatment ($D_i = 1$)
    \item $Y_i(0)$: The outcome that \emph{would} obtain if unit $i$ does not receive treatment ($D_i = 0$)
\end{itemize}
These potential outcomes exist conceptually for every unit, regardless of which treatment the unit actually receives.
The observed outcome is determined by which potential outcome is ``revealed'' by the treatment assignment:
\begin{equation}
    Y_i = D_i \cdot Y_i(1) + (1 - D_i) \cdot Y_i(0)
\end{equation}
This switching equation makes explicit that we observe $Y_i(1)$ when $D_i = 1$ and $Y_i(0)$ when $D_i = 0$, but never both.
The individual causal effect for unit $i$ is defined as:
\begin{equation}
    \tau_i = Y_i(1) - Y_i(0)
\end{equation}
This quantity answers the question: How much would $i$'s outcome change if we switched $i$ from control to treatment?
The challenge is that $\tau_i$ is inherently unobservable.
We can observe at most one of $Y_i(1)$ or $Y_i(0)$ for any given unit; the other remains forever counterfactual.
A developer either adopts a tool or does not; a project either uses one language or another.
We cannot rewind time and observe what would have happened under the alternative.

This impossibility of observing individual causal effects has profound implications.
It means that causal inference is fundamentally a problem of \emph{missing data}: For every treated unit, we are missing the control outcome; for every control unit, we are missing the treated outcome.
The entire enterprise of causal inference consists of finding ways to fill missing potential outcomes---or, more precisely, to estimate population summaries without needing to observe them individually.

\paragraph{Causal Estimands.}
Since individual effects are unobservable, we focus on population-level summaries.
The most common estimand is the \emph{Average Treatment Effect} (ATE):
\begin{equation}
    \text{ATE} = E[Y(1) - Y(0)] = E[Y(1)] - E[Y(0)]
\end{equation}
The ATE answers: On average across the population, how much does treatment change outcomes?
This is the quantity of interest when we ask whether a policy should be adopted universally.

An alternative estimand is the \emph{Average Treatment Effect on the Treated} (ATT):
\begin{equation}
    \text{ATT} = E[Y(1) - Y(0) \mid D = 1]
\end{equation}
The ATT answers: Among those who actually received treatment, what was the average effect?
This is relevant when evaluating an existing program: Given that certain people chose to participate, did participation help them?
Note that ATE and ATT coincide only when treatment effects are homogeneous or when treatment assignment is independent of potential outcomes.

Other estimands include the \emph{Average Treatment Effect on the Controls} (ATC), \emph{quantile treatment effects} (effects on specific percentiles of the outcome distribution), and \emph{conditional average treatment effects} (CATEs) that allow effects to vary with observed characteristics.
The choice of estimand should be guided by the substantive question, not statistical convenience.

\paragraph{Why Simple Comparisons Fail.}
A naive approach to estimating the ATE would compare mean outcomes between treated and control groups:
\begin{equation}
    E[Y \mid D=1] - E[Y \mid D=0]
\end{equation}
Expanding this in terms of potential outcomes reveals the problem:
\begin{align}
    E[Y \mid D=1] - E[Y \mid D=0] &= E[Y(1) \mid D=1] - E[Y(0) \mid D=0] \\
    &= \underbrace{E[Y(1) \mid D=1] - E[Y(0) \mid D=1]}_{\text{ATT}} \\ &+ \underbrace{E[Y(0) \mid D=1] - E[Y(0) \mid D=0]}_{\text{Selection Bias}}
\end{align}
The first term is the ATT---the causal effect among the treated.
The second term is \emph{selection bias}: The difference in baseline potential outcomes between those who select into treatment and those who do not.
If treatment is self-selected, these groups may differ systematically even in the absence of treatment.
High-ability developers may adopt new tools precisely because they are high-ability; comparing their outcomes with those of non-adopters confounds the tool's effect with pre-existing ability differences.
The observed difference equals the causal effect only when selection bias is zero---that is, when $E[Y(0) \mid D=1] = E[Y(0) \mid D=0]$.
This holds when treatment is randomly assigned, but rarely otherwise.

\paragraph{Identifying Assumptions.}
To bridge the gap between observable quantities and causal estimands, we require assumptions that discipline how potential outcomes relate to treatment assignment.
Three assumptions are foundational:

\begin{enumerate}[leftmargin=15pt]
    \item \textbf{Stable Unit Treatment Value Assumption (SUTVA).}
    This assumption has two components.
    First, \emph{no interference}: The potential outcomes for unit $i$ depend only on $i$'s own treatment status, not on the treatment status of other units.
    If one developer's adoption of a tool changes another developer's productivity (through collaboration, knowledge sharing, or competition), SUTVA is violated.
    Second, \emph{no hidden variations of treatment}: There is only one version of treatment, so $Y_i(1)$ is well-defined.
    If ``adopting a tool'' means different things for different developers (some use it intensively, others superficially), the single potential outcome $Y_i(1)$ may be poorly defined.
    SUTVA is often violated in networked settings.
    If developers on the same team influence each other, individual-level analysis may be inappropriate; the relevant unit may be the team, not the developer.
    
    \item \textbf{Ignorability (Unconfoundedness, Conditional Independence).}
    Conditional on observed covariates $X$, treatment assignment is independent of potential outcomes:
    \begin{equation}
        (Y(1), Y(0)) \perp D \mid X
    \end{equation}
    This assumption asserts that, among units with identical observed characteristics, treatment is as-if randomly assigned.
    Any systematic differences between treated and control groups are fully captured by $X$; once we condition on $X$, no unmeasured confounders remain.
    Ignorability is the critical assumption for observational studies.
    It is inherently untestable---because we cannot observe the unmeasured confounders, we cannot verify that all confounders have been measured.
    Its plausibility rests entirely on the researcher's knowledge of the data-generating process---or the plausibility of the causal theory based on the researcher's knowledge.
    
    \item \textbf{Overlap (Positivity, Common Support).}
    For all values of $X$ in the population, both treatment states have positive probability:
    \begin{equation}
        0 < P(D=1 \mid X) < 1 \quad \text{for all } X
    \end{equation}
    This ensures that for any combination of covariates, we can find both treated and untreated units, enabling comparison.
    If some covariate combinations are treated with probability 1 or 0, we have no counterfactual basis for estimating effects in that region.
    Overlap is testable: We can usually examine whether the distributions of covariates (or propensity scores) in treated and control groups have common support.
    Violations suggest that certain treatment effects are not identified from the data.
\end{enumerate}

\paragraph{Identification Under Ignorability.}
Under SUTVA, ignorability, and overlap, the ATE is identified from observed data.
The key insight is that ignorability allows us to substitute observable conditional expectations for unobservable potential outcome means:
\begin{align}
    E[Y(1)] &= E_X[E[Y(1) \mid X]] = E_X[E[Y(1) \mid X, D=1]] = E_X[E[Y \mid X, D=1]] \\
    E[Y(0)] &= E_X[E[Y(0) \mid X]] = E_X[E[Y(0) \mid X, D=0]] = E_X[E[Y \mid X, D=0]]
\end{align}
The first equality uses iterated expectations.
The second uses ignorability: $E[Y(1) \mid X] = E[Y(1) \mid X, D=1]$ because treatment is independent of potential outcomes given $X$.
The third uses the switching equation: Among the treated, observed $Y$ equals $Y(1)$.

Therefore:
\begin{equation}
    \text{ATE} = E_X[E[Y \mid D=1, X] - E[Y \mid D=0, X]]
\end{equation}
This expression involves only observable quantities: Conditional means of $Y$ given treatment status and covariates, averaged over the covariate distribution.
Various estimation strategies---regression adjustment, propensity score matching, inverse probability weighting, doubly robust estimation---implement this identification result using different statistical approaches.

\subsection{Graphical Causal Models}

While the potential outcomes framework excels at defining estimands and clarifying the role of randomization, it is less helpful for reasoning about \emph{which} variables to condition on.
The graphical causal model framework, developed by Pearl and colleagues, addresses this gap by representing causal assumptions as directed acyclic graphs (DAGs) and providing algorithmic criteria for determining when causal effects are identifiable~\cite{pearl2009causality}.

\paragraph{Directed Acyclic Graphs.}
A DAG $G = (V, E)$ consists of a set of nodes $V$ representing variables and a set of directed edges $E$ representing direct causal relationships.
An edge $X \to Y$ asserts that $X$ has a direct causal effect on $Y$---changing $X$ while holding all other causes of $Y$ fixed would change $Y$.
The graph is ``acyclic'' because no variable can be its own ancestor: Following directed edges, one can never return to the starting node.
This rules out simultaneous causation (which requires more complex frameworks) but permits chains of causation over time.

The power of DAGs lies in what they make explicit.
Every arrow represents a causal claim; every absent arrow asserts the absence of a direct effect.
Critics can challenge a causal model by questioning specific edges: ``Why do you assume $X$ affects $Y$ directly?'' or ``Have you considered that $Z$ might cause both $X$ and $Y$?''
This transparency is impossible with regression equations, where causal assumptions are buried in the choice of variables and functional forms.

\paragraph{Paths and Associations.}
In a DAG, variables can be associated through multiple mechanisms.
A \emph{path} is any sequence of edges connecting two variables, regardless of direction.
Three types of paths are fundamental:

\begin{enumerate}[leftmargin=15pt]
    \item \emph{Causal paths} (or directed paths): $X \to \cdots \to Y$.
    All edges point from $X$ toward $Y$.
    These represent the causal effect of $X$ on $Y$, possibly mediated through intermediate variables.
    
    \item \emph{Back-door paths} (or confounding paths): $X \leftarrow \cdots \to Y$ or $X \leftarrow \cdots \leftarrow Z \to \cdots \to Y$.
    These paths have an arrow \emph{into} $X$ and connect $X$ to $Y$ through a common cause.
    They represent spurious association due to confounding.
    
    \item \emph{Collider paths}: $X \to \cdots \to C \leftarrow \cdots \leftarrow Y$.
    These paths pass through a \emph{collider}---a variable with arrows pointing into it from both directions.
    Collider paths are naturally ``blocked'' and do not transmit association unless we condition on the collider (or its descendants), at which point they become ``open.''
\end{enumerate}

\paragraph{d-Separation: When Variables Are Independent.}
The criterion of \emph{d-separation} (``d'' for directional) determines when two variables are statistically independent given a conditioning set.
A path between $X$ and $Y$ is \emph{blocked} by a set $Z$ if either: (1) The path contains a non-collider (a variable with at most one arrow pointing in) that is in $Z$, or (2) The path contains a collider that is \emph{not} in $Z$ and has no descendants in $Z$.
Variables $X$ and $Y$ are d-separated by $Z$ if all paths between them are blocked by $Z$.
D-separation implies conditional independence: If $X$ and $Y$ are d-separated by $Z$ in the graph, then $X \perp Y \mid Z$ in any distribution compatible with the graph.

The practical implication is that we can read off conditional independence relations directly from the graph, without computing probabilities.
This allows researchers to check whether their assumed causal structure is consistent with observed data (by testing the implied independencies) and to identify which variables must be conditioned on to block confounding.

\paragraph{The do-Operator and Interventional Distributions.}
The central distinction in Pearl's framework is between \emph{observing} that $X = x$ and \emph{intervening} to set $X = x$.
These correspond to different probability distributions:
\begin{itemize}[leftmargin=15pt]
    \item $P(Y \mid X = x)$: The conditional distribution of $Y$ given that we \emph{observe} $X = x$.
    This may reflect both the causal effect of $X$ on $Y$ and confounding from common causes.
    
    \item $P(Y \mid do(X = x))$: The distribution of $Y$ when we \emph{intervene} to set $X = x$, overriding whatever process normally determines $X$.
    This isolates the causal effect, as the intervention breaks the connection between $X$ and its causes.
\end{itemize}

Graphically, the intervention $do(X = x)$ corresponds to \emph{graph surgery}: We delete all arrows pointing into $X$ (since the intervention replaces $X$'s natural causes) while preserving all arrows pointing out of $X$ (since $X$ still affects its descendants).
In the modified graph, $X$ becomes an exogenous variable set to $x$, and the resulting distribution is the interventional distribution.

The fundamental goal of causal inference is to estimate $P(Y \mid do(X = x))$ from observational data, which only provides $P(Y \mid X = x)$.
When are these equal?
Precisely when there is no confounding---that is, when all paths from $X$ to $Y$ except the causal paths are blocked.

\paragraph{The Back-Door Criterion.}
Pearl's back-door criterion provides a graphical test for when conditioning on a set of variables $Z$ suffices to identify the causal effect.
A set $Z$ satisfies the back-door criterion relative to $(X, Y)$ if:
\begin{enumerate}[leftmargin=15pt]
    \item No variable in $Z$ is a descendant of $X$ (to avoid conditioning on mediators or colliders on the causal path).
    \item $Z$ blocks all back-door paths from $X$ to $Y$ (paths with an arrow into $X$).
\end{enumerate}

When the back-door criterion is satisfied, the interventional distribution is identified by adjustment:
\begin{equation}
    P(Y \mid do(X = x)) = \sum_z P(Y \mid X = x, Z = z) P(Z = z)
\end{equation}
This formula has a natural interpretation: We compute the effect of $X$ on $Y$ within each stratum defined by $Z$, then average over the distribution of $Z$ in the population.
The result is the causal effect, purged of confounding.

The back-door criterion illuminates the ``bad controls'' problem discussed earlier.
Conditioning on a mediator (a descendant of $X$ on the causal path) violates condition 1 and blocks part of the causal effect.
Conditioning on a collider opens a previously blocked path and introduces bias.
The graphical framework provides a systematic way to determine which variables should and should not be included in an adjustment set.

\paragraph{Beyond the Back-Door Criterion.}
When no set of observed variables satisfies the back-door criterion---perhaps because key confounders are unmeasured---causal effects may still be identifiable through alternative strategies.
Pearl's do-calculus provides a complete set of rules for determining identifiability from any combination of observational and experimental data.
The \emph{front-door criterion}, for instance, identifies causal effects through mediating variables when the direct path is confounded but the mediator-outcome relationship is not.
Instrumental variables, discussed in Section~\ref{sec:identification}, can be understood graphically as exploiting exogenous variation that affects treatment but not outcomes directly.

\subsection{Using the Frameworks in Practice}

How should an applied researcher use these frameworks?
In practice, neither framework is a statistical method that one ``runs'' on data.
Rather, they are \emph{conceptual tools for reasoning about causation}---languages for articulating what causal effect one seeks to estimate, what assumptions would justify that estimate, and what could go wrong.
The actual estimation is performed by specific statistical methods---difference-in-differences, regression discontinuity, instrumental variables, matching---which the following sections present.
Each method makes specific assumptions (parallel trends, continuity at the threshold, exclusion restrictions) that can be understood and justified within either framework.

The practical workflow typically proceeds as follows.
First, the researcher uses the potential outcomes framework to clarify the research question: What is the treatment?
What is the outcome?
What causal estimand (ATE, ATT, LATE) is of interest?
Second, the researcher draws a causal graph encoding domain knowledge about the data-generating process, identifying potential confounders, mediators, and colliders.
This step forces explicit articulation of assumptions that might otherwise remain implicit.
Third, the researcher selects an identification strategy---a research design that, under stated assumptions, provides a credible path from observed data to causal effects.
The choice depends on what natural experiments or exogenous variation the setting provides.
Finally, the researcher implements the chosen method using appropriate statistical techniques, conducts robustness checks, and interprets results in light of the assumptions made.

Though developed independently, the two frameworks are mathematically compatible and often used together.
The potential outcomes condition of ignorability---$(Y(1), Y(0)) \perp D \mid X$---corresponds precisely to d-separation in an appropriately constructed graph.
In practice, potential outcomes excel at the first step (defining estimands and clarifying what we want to learn), while graphical models excel at the second step (encoding assumptions visually and determining what to adjust for).
The frameworks thus serve as the \emph{theoretical foundation} upon which specific methods rest: A difference-in-differences analysis implicitly assumes that the parallel trends assumption blocks all back-door paths; an instrumental variable strategy assumes the instrument satisfies exclusion restrictions that can be depicted graphically.
Understanding both frameworks enables researchers to see what each method achieves and where it might fail---knowledge essential for choosing appropriate designs and interpreting results honestly.

\section{Identification Strategies}
\label{sec:identification}

This section presents research designs that, under their stated assumptions, provide credible causal identification.
Each method is characterized by its identifying assumption and the alternative explanations it rules out.
Unlike the control strategies discussed in Section~\ref{sec:control}, identification strategies provide a complete path from observed data to causal effects---if their assumptions hold, the resulting estimates would have a causal interpretation.

\subsection{Randomized Controlled Trials}

\paragraph{Historical Context.}
The randomized controlled trial (RCT) emerged from agricultural research in the early 20th century.
Ronald Fisher, working at the Rothamsted Experimental Station at that time, developed the theoretical foundations of randomization in his landmark 1935 book \emph{The Design of Experiments}~\cite{fisher1971design}.
Fisher's key insight was that random assignment of treatments to experimental units provides a principled basis for statistical inference: It ensures that treatment and control groups are comparable in expectation across all characteristics, observed and unobserved, and it justifies the use of probability theory to quantify uncertainty.
The method revolutionized medical research after Austin Bradford Hill applied it to the first modern clinical trial of streptomycin for tuberculosis in 1948, establishing the RCT as the ``gold standard'' for evaluating medical interventions~\cite{bothwell2016assessing}.

\paragraph{Intuition.}
The power of randomization lies in its simplicity: By flipping a coin to determine who receives treatment, we break the link between treatment assignment and all potential confounders.
Before randomization, treated and control groups might differ systematically---perhaps more motivated patients seek treatment, or better-funded teams adopt new tools.
These pre-existing differences confound any comparison of outcomes.
Randomization eliminates this problem by ensuring that, on average, the only difference between groups is the treatment itself.
Any difference in outcomes can therefore be attributed to the treatment rather than to selection effects.

\paragraph{Formal Framework.}
Let $D_i \in \{0, 1\}$ denote random treatment assignment for unit $i$.
Randomization ensures that treatment is independent of potential outcomes:
\begin{equation}
    (Y_i(1), Y_i(0)) \perp D_i
\end{equation}
This is the strongest form of ignorability---it holds unconditionally, without requiring adjustment for covariates.
Under this assumption, the ATE is identified by a simple difference in means:
\begin{equation}
    \text{ATE} = E[Y_i(1)] - E[Y_i(0)] = E[Y_i | D_i = 1] - E[Y_i | D_i = 0]
\end{equation}
The sample analogue provides an unbiased estimator:
\begin{equation}
    \widehat{\text{ATE}} = \frac{1}{n_1} \sum_{i: D_i = 1} Y_i - \frac{1}{n_0} \sum_{i: D_i = 0} Y_i
\end{equation}
where $n_1$ and $n_0$ are the numbers of treated and control units, respectively.
Standard errors follow from the randomization distribution or, asymptotically, from the central limit theorem.

\paragraph{RCTs in Computer Science Research.}
Randomized experiments have a rich tradition across computer science, though their prevalence varies by subfield.
In human-computer interaction (HCI) research, controlled experiments comparing interface designs, interaction techniques, or visualization methods are standard practice, as the field inherited experimental traditions from psychology and human factors engineering~\cite{DBLP:conf/chi/Vaithilingam0G22}.
In computing education research, randomized trials routinely evaluate pedagogical interventions---comparing teaching methods, programming environments, or curricular structures---with studies randomly assigning students to different instructional conditions.
In the IT industry, A/B testing has become ubiquitous for user experience research~\cite{kohavi2020trustworthy}.
Large technology companies run thousands of online experiments annually, randomly assigning users to different interface designs, recommendation algorithms, or feature variants, and measuring outcomes like engagement, conversion, or retention.
These experiments provide clear causal evidence for product decisions because randomization eliminates the selection effects that plague observational comparisons.

\paragraph{Application in Software Engineering.}
Within software engineering research, RCTs are less common but increasingly influential.
A prominent recent example is the METR study by Becker et al., which conducted a randomized controlled trial with experienced open-source developers working on their own real-world projects~\cite{DBLP:journals/corr/abs-2507-09089}.
The study found that AI tools did not significantly reduce completion time for experienced developers---a result that diverged from earlier experiments showing large productivity gains.
Such experiments provide clear causal evidence while maintaining relevance to real-world practice.

While the appeal of RCTs in software engineering is clear, they also face significant practical limitations in this context.
First, many important questions are not amenable to randomization and are impractical for an RCT design.
It would be prohibitively expensive, unethical, or simply infeasible to run an experiment randomly assigning programming languages to projects in production, assigning developers to adopt tools they resist, or experimentally manipulating organizational culture.
Second, practical constraints limit the duration and scale of experiments: Software projects unfold over months or years, whereas experiments typically run for days or weeks. 
A tool that initially slows developers (as they learn it) but eventually accelerates them would show negative effects in a week-long experiment, but positive effects over months.
Finally, perhaps the most fundamental limitation of RCTs in software engineering is external validity~\cite{DBLP:journals/ese/KalliamvakouGBS16}.
For example, laboratory experiments with student participants completing artificial tasks may not generalize to professional developers working on production systems under real-world pressures.
Yet, these limitations do not diminish the value of RCTs when they are feasible and well-executed---they provide an important source of triangulation on one side of the trade-off: limited external validity in exchange for the highest internal validity.

\subsection{Difference-in-Differences}
\label{sec:chap2-did}

\paragraph{Historical Context.}
Difference-in-differences (DiD) has roots in 19th-century epidemiology---John Snow's comparison of cholera rates across London water companies implicitly used the logic of DiD---but its modern form crystallized in labor economics.
The method gained prominence through Ashenfelter's 1978 study of job training programs~\cite{ashenfelter1978estimating} and achieved canonical status with Card and Krueger's 1994 study of minimum wage effects in New Jersey and Pennsylvania~\cite{card1993minimum}.
This study compared employment changes in New Jersey fast-food restaurants (where the minimum wage increased) with those in neighboring Pennsylvania (where it did not), finding no negative employment effect and overturning decades of conventional economic wisdom.
The ``credibility revolution'' in economics, recognized by the 2021 Nobel Prize, built substantially on the widespread adoption of DiD and related methods~\cite{cunningham2021causal}.

\paragraph{Intuition.}
The core insight of DiD is that we can control for unobserved confounders by exploiting \emph{changes over time}.
For example, suppose we want to know whether adopting continuous integration (CI) improves code quality.
Simply comparing projects with and without CI conflates the treatment effect with selection: Perhaps better-organized teams both adopt CI and write better code.
The naive cross-sectional comparison captures this selection effect along with any true causal effect. 
DiD addresses this by adding a temporal dimension: We observe projects before and after CI adoption and compare the \emph{change} in quality of these projects with that of similar projects that did not adopt CI over the same period.
If both groups had followed parallel trajectories in the absence of treatment, then any differential change can be attributed to CI adoption.
The method effectively ``differences out'' time-invariant confounders (team quality, project complexity) by focusing on within-unit changes, and ``differences out'' common time trends (industry-wide quality improvements) by comparing to a control group.

\paragraph{Formal Framework.}
Consider the canonical $2 \times 2$ case with two groups (treatment and control) and two periods (before and after).
Let $Y_{g,t}$ denote the mean outcome for group $g \in \{0, 1\}$ at time $t \in \{0, 1\}$, where $g = 1$ indicates the treatment group and $t = 1$ indicates the post-treatment period.
The DiD estimator is:
\begin{equation}
    \widehat{\tau}_{\text{DiD}} = (Y_{1,1} - Y_{1,0}) - (Y_{0,1} - Y_{0,0})
\end{equation}
The first difference $(Y_{1,1} - Y_{1,0})$ captures the change in the treatment group; the second difference $(Y_{0,1} - Y_{0,0})$ captures the counterfactual change that would have occurred in the absence of treatment.

The identifying assumption is \emph{parallel trends}: In the absence of treatment, both groups would have experienced the same change in outcomes.
Formally, letting $Y_{g,t}(0)$ denote potential outcomes under control:
\begin{equation}
    E[Y_{1,1}(0) - Y_{1,0}(0)] = E[Y_{0,1}(0) - Y_{0,0}(0)]
\end{equation}
Under this assumption, the DiD estimator identifies the ATT:
\begin{equation}
    \tau_{\text{ATT}} = E[Y_{1,1}(1) - Y_{1,1}(0)] = \widehat{\tau}_{\text{DiD}}
\end{equation}

\paragraph{Panel Data and Staggered Adoption.}
When multiple entities are observed over multiple periods with treatment adoption occurring at different times, DiD extends to the panel setting~\cite{athey2022design}.
The standard two-way fixed effects (TWFE) specification is:
\begin{equation}
\label{eq:base-did}
    Y_{it} = \alpha + \tau D_{it} + \mu_i + \lambda_t + \epsilon_{it}
\end{equation}
where $\mu_i$ are entity fixed effects (absorbing time-invariant characteristics), $\lambda_t$ are time fixed effects (absorbing common shocks), and $D_{it}$ indicates whether entity $i$ is treated at time $t$.

\paragraph{Event Studies.}
To examine dynamic treatment effects and assess the parallel trends assumption, researchers estimate event study specifications:
\begin{equation}
\label{eq:panel-event}
    Y_{it} = \alpha + \sum_{k \neq -1} \tau_k \cdot \mathbf{1}[K_{it} = k] + \mu_i + \lambda_t + \epsilon_{it}
\end{equation}
where $K_{it}$ denotes the number of periods relative to treatment adoption (negative for pre-treatment, positive for post-treatment).
The coefficients $\tau_k$ for $k < 0$ serve as placebo tests: Significant pre-trends suggest that treatment and control groups were already diverging before treatment, violating parallel trends.
Post-treatment coefficients $\tau_k$ for $k \geq 0$ trace out the dynamic treatment effect after adoption.

\paragraph{Recent Advances.}
Recent methodological advances have revealed that TWFE can produce biased estimates when treatment effects vary across units or over time~\cite{goodman2021difference, de2020two}.
The problem arises because TWFE implicitly uses already-treated units as controls for newly-treated units, potentially introducing negative weights.
Modern estimators address this through clean control comparisons~\cite{callaway2021difference}, imputation methods~\cite{borusyak2024revisiting}, or stacking approaches~\cite{baker2022much}.
Best practice now involves reporting results from multiple estimators and conducting sensitivity analyses~\cite{roth2023s}.

\paragraph{Application in Software Engineering.}
While currently still not common in software engineering research, DiD designs are potentially well-suited for a family of empirical software engineering problems on the effects of tool adoptions, policy changes, or platform updates in software development.
Recent examples illustrate this potential: Fang et al.\ used a DiD design to estimate the causal effect of tweets on open-source project popularity and contributor growth~\cite{DBLP:conf/icse/FangLHV22}; Casanueva et al.\ applied DiD to estimate how the COVID-19 pandemic disproportionately affected women's contributions to public code~\cite{casanueva2025impact}; and Chapter~\ref{chp:cursor} of this thesis applies DiD with staggered adoption to estimate the effects of Cursor AI on development velocity and software quality.
These studies demonstrate that DiD is broadly applicable in software engineering contexts, where natural variation in adoption timing creates quasi-experimental conditions, to support causal claims to important software engineering problems.

\subsection{Regression Discontinuity}

\paragraph{Historical Context.}
The regression discontinuity design (RDD) was invented by psychologists Donald Thistlethwaite and Donald Campbell in 1960 to evaluate the effect of National Merit Scholarships on career aspirations~\cite{thistlethwaite1960regression}.
In their setting, students who were just above the scholarship cutoff received awards; those who were just below did not.
By comparing outcomes for students on either side of this arbitrary threshold, the authors could estimate the scholarship's effect without the selection bias that would plague comparisons between scholarship recipients and non-recipients.
The method languished in relative obscurity until economists rediscovered it in the 1990s, after which it became one of the most credible quasi-experimental designs available.
Angrist and Lavy's 1999 study of class-size effects, exploiting a maximum class-size rule, exemplifies the modern application of RDD in the social sciences~\cite{angrist1999using}.

\paragraph{Intuition.}
The power of RDD derives from the arbitrariness of cutoffs.
When treatment is assigned based on whether a continuous variable (the ``running variable'') exceeds a threshold, units just above and just below the cutoff are essentially identical---they differ only by an infinitesimal amount on the running variable.
If the cutoff is set at 70 points, a student scoring 70.1 is virtually indistinguishable from one scoring 69.9 in all respects except their treatment status.
Comparing outcomes across these nearly identical units approximates a local randomized experiment.

The key requirement is that units cannot precisely manipulate their position relative to the cutoff.
If students can ensure they score exactly above the threshold, those who do so may differ systematically from those who do not, reintroducing selection bias.
But if scoring involves some randomness or imprecision, manipulation becomes impossible, and the design remains valid.

\paragraph{Formal Framework.}
Let $R_i$ denote the running variable for unit $i$, $c$ the cutoff, and $D_i = \mathbf{1}[R_i \geq c]$ the treatment indicator.
The sharp RDD identifies the treatment effect at the cutoff as the discontinuity in the conditional expectation function:
\begin{equation}
    \tau_{\text{RDD}} = \lim_{r \downarrow c} E[Y_i | R_i = r] - \lim_{r \uparrow c} E[Y_i | R_i = r]
\end{equation}

The identifying assumption is \emph{continuity}: The conditional expectations of potential outcomes are continuous at the cutoff.
Formally:
\begin{equation}
    \lim_{r \downarrow c} E[Y_i(d) | R_i = r] = \lim_{r \uparrow c} E[Y_i(d) | R_i = r] \quad \text{for } d \in \{0, 1\}
\end{equation}
This rules out any discontinuity in outcomes at $c$ except that caused by the treatment itself.

Estimation typically employs local polynomial regression, fitting separate polynomials on each side of the cutoff and estimating the discontinuity as the difference in predicted values at $c$.
Bandwidth selection---how much data around the cutoff to use---involves a bias-variance tradeoff: Narrower bandwidths reduce bias from model misspecification but increase variance.
Optimal bandwidth selectors and robust confidence intervals are now standard~\cite{calonico2014robust}.

\paragraph{Fuzzy RDD.}
In many applications, crossing the threshold increases the \emph{probability} of treatment rather than determining it completely.
The fuzzy RDD extends the design to this setting by using the discontinuity in treatment probability as an instrument:
\begin{equation}
    \tau_{\text{Fuzzy}} = \frac{\lim_{r \downarrow c} E[Y_i | R_i = r] - \lim_{r \uparrow c} E[Y_i | R_i = r]}{\lim_{r \downarrow c} E[D_i | R_i = r] - \lim_{r \uparrow c} E[D_i | R_i = r]}
\end{equation}
This identifies a local average treatment effect for compliers at the threshold.

\paragraph{Application in Software Engineering.}
To the best of our knowledge, a true RDD has not yet been applied in software engineering research.
Some studies claim to use ``regression discontinuity'' but actually employ \emph{interrupted time series} (ITS) designs---a fundamentally different method that compares outcomes before and after an event in time, rather than comparing units just above and below a threshold on a continuous running variable.
ITS exploits temporal discontinuities (e.g., before versus after a policy change), whereas RDD exploits cross-sectional discontinuities at a cutoff (e.g., projects just above versus just below a threshold).
Conflating these methods obscures their distinct identifying assumptions and threatens the validity of causal claims.

In principle, RDD could apply in software engineering whenever platforms use threshold-based rules for resource allocation, feature access, or automated decisions.
A tantalizing near-miss occurred with GitHub Copilot's 2021 technical preview, which initially invited only a subset of developers to access the tool.
If GitHub had used a clear, observable threshold for invitation---say, developers with more than 1,000 contributions receiving access while those just below did not---comparing productivity outcomes for developers just above and below this cutoff would constitute a valid RDD.
In practice, however, GitHub's selection criteria were opaque and likely multidimensional, precluding a clean discontinuity design.
This illustrates the fundamental challenge: Even when platforms implement access restrictions, the thresholds are rarely transparent, sharp, or based on a single running variable---requirements essential for RDD validity.
The identification of genuine RDD opportunities in software engineering problems remains an open challenge for future research.

\subsection{Instrumental Variables}

\paragraph{Historical Context.}
Instrumental variables estimation has deep roots in economics, originating with Philip Wright's 1928 study of supply and demand for butter and flaxseed~\cite{wright1928tariff, stock2003retrospectives}.
Wright recognized that ordinary regression of quantity on price conflates supply and demand responses; weather shocks that affect supply but not demand provide an instrument for identifying the demand curve.
The method was formalized by the Cowles Commission economists in the 1940s and became central to structural econometrics~\cite{stock2003retrospectives}.
Its application to causal inference in program evaluation was systematized by Angrist and Imbens in their influential 1996 paper, which introduced the Local Average Treatment Effect (LATE) interpretation and earned them the 2021 Nobel Prize~\cite{angrist1996identification}.

\paragraph{Intuition.}
Instrumental variables address situations where treatment is endogenous---correlated with unobserved factors that also affect outcomes.
The strategy is to find a source of variation in treatment that is ``exogenous''---unrelated to the confounders---and use only that variation for estimation.
This exogenous variation is the ``instrument.''

Consider the challenge of estimating whether completing a computer science degree increases earnings.
Simply comparing graduates to non-graduates conflates the degree's effect with selection: Students who complete degrees may have higher ability, motivation, or family resources that would have led to higher earnings regardless.
An instrument must affect degree completion but influence earnings only through its effect on completing the degree.
Geographic proximity to a college might work: Students who grow up near colleges are more likely to attend (relevance), but proximity itself does not directly affect later earnings (exclusion restriction).
By focusing on the earnings variation explained by proximity---variation that is plausibly unrelated to unobserved ability---we can isolate the degree's causal effect.

\paragraph{Formal Framework.}
Let $D_i$ denote the endogenous treatment, $Y_i$ the outcome, $Z_i$ the instrument, and $U_i$ unobserved confounders.
The core problem is that $Cov(D_i, U_i) \neq 0$, so regressing $Y$ on $D$ yields biased estimates.
A valid instrument satisfies three conditions:
\begin{enumerate}[leftmargin=15pt]
    \item \textbf{Relevance}: $Cov(Z_i, D_i) \neq 0$.
    The instrument must affect treatment.
    This condition is testable by examining the first-stage regression of $D$ on $Z$.
    
    \item \textbf{Independence (Exogeneity)}: $Cov(Z_i, U_i) = 0$.
    The instrument must be uncorrelated with unobserved confounders.
    This condition is fundamentally untestable but must be argued on substantive grounds.
    
    \item \textbf{Exclusion Restriction}: $Z_i$ affects $Y_i$ only through $D_i$.
    There is no direct effect of the instrument on outcomes.
    This condition is also untestable and must be justified by theory or institutional knowledge.
\end{enumerate}

Under these assumptions, the IV estimator identifies the causal effect:
\begin{equation}
    \beta_{\text{IV}} = \frac{Cov(Y_i, Z_i)}{Cov(D_i, Z_i)}
\end{equation}
This is the ratio of the reduced-form effect of $Z$ on $Y$ to the first-stage effect of $Z$ on $D$.
In the two-stage least squares (2SLS) implementation, we first regress $D$ on $Z$ to obtain predicted values $\hat{D}$, then regress $Y$ on $\hat{D}$.
The second-stage coefficient is the IV estimate.

\paragraph{Local Average Treatment Effect.}
When treatment effects are heterogeneous, IV does not identify the ATE.
Angrist and Imbens showed that IV identifies the \emph{Local Average Treatment Effect} (LATE): The average effect for ``compliers''---units whose treatment status is changed by the instrument~\cite{angrist1996identification}.
Units who would always take treatment regardless of the instrument (always-takers) and those who would never take treatment (never-takers) contribute no identifying variation.
Defiers---who do the opposite of what the instrument encourages---are typically assumed not to exist (monotonicity assumption).

The LATE interpretation has important implications: The estimated effect may not generalize to the full population if compliers differ systematically from always-takers and never-takers.
Understanding who the compliers are is essential for interpreting IV results.

\paragraph{Application in Software Engineering.}
To the best of our knowledge, IV designs have not yet been applied in software engineering research, likely because finding valid instruments is particularly challenging in this domain.
Nonetheless, plausible opportunities exist.
Consider estimating whether attending a coding bootcamp improves employment outcomes.
Bootcamp attendance is endogenous: Motivated individuals both seek training and find jobs.
An instrument might be the opening of a bootcamp campus in a city---this plausibly increases local attendance (relevance) but does not directly affect employment except through training (exclusion restriction).

Another example involves platform features.
To estimate whether enabling two-factor authentication reduces account compromises, one might use the timing of reminder emails as an instrument.
Receiving a reminder increases the probability of enabling 2FA (relevance), and the timing of automated reminders is plausibly random with respect to individual security outcomes (independence).
The exclusion restriction requires that reminders affect security only through their effect on 2FA adoption, not through other channels like increased security awareness.

The fundamental difficulty with IV is justifying the exclusion restriction: How do we know the instrument affects outcomes only through treatment?
In software engineering contexts, instruments often have plausible alternative pathways.
A new bootcamp in a city may affect local labor markets beyond training; reminder emails may increase general security awareness.
Credible IV analyses require transparent argument about why these alternative channels are negligible and, where possible, empirical tests of the exclusion restriction's implications.


\section{Control Strategies}
\label{sec:control}

Unlike identification strategies, control strategies address specific sources of confounding without providing complete causal identification.
They are valuable tools but should be combined with substantive arguments about remaining threats.
The distinction is important: Identification strategies (RCTs, DiD, RDD, IV) provide a complete logical path from data to causal effects under their stated assumptions; control strategies reduce confounding but leave open the possibility that unmeasured confounders or other threats remain.

\subsection{Matching and Propensity Scores}

\paragraph{Historical Context.}
Matching---the idea of comparing treated units to similar untreated units---has ancient roots in experimental design, but its modern statistical formalization emerged in the 1970s and 1980s.
Donald Rubin's 1973 paper on matching in observational studies provided theoretical foundations, and his collaboration with Paul Rosenbaum in 1983 introduced the propensity score, which revolutionized matching methods~\cite{rosenbaum1983central}.
The propensity score reduced the ``curse of dimensionality'' in matching: Instead of finding exact matches on dozens of covariates (typically impossible), researchers could match on a single scalar summary.
This innovation made matching practical for studies with many confounders and propelled its widespread adoption in epidemiology, economics, and the social sciences.

\paragraph{Intuition.}
The fundamental challenge in observational studies is that treated and control groups differ systematically.
If we simply compare outcomes between GitHub repositories that adopted continuous integration (CI) and those that did not, any difference might reflect CI's effect or might reflect pre-existing differences between adopters and non-adopters.
Perhaps repositories with more active maintainers both adopt CI and have better outcomes, regardless of CI.

Matching addresses this by constructing a comparison group that ``looks like'' the treatment group on observed characteristics.
For each treated repository, we find untreated repositories with similar characteristics---similar size, age, programming language, contributor count---and compare outcomes within these matched pairs.
If matching succeeds in balancing all confounders, the matched comparison approximates what a randomized experiment would have produced.

The propensity score provides an elegant way to accomplish this.
Rather than matching on each covariate separately, we model the probability of treatment given covariates and match on this single number.
Rosenbaum and Rubin proved that if matching on all covariates would eliminate confounding, then matching on the propensity score does too.
Intuitively, units with the same propensity score are equally likely to be treated based on observed characteristics, so treatment assignment among them is ``as if random'' with respect to those characteristics.

\paragraph{Formal Framework.}
Let $X_i$ denote a vector of pre-treatment covariates for unit $i$, $D_i \in \{0,1\}$ the treatment indicator, and $Y_i$ the outcome.
The propensity score is defined as:
\begin{equation}
    e(X_i) = P(D_i = 1 \mid X_i)
\end{equation}
In practice, the propensity score is estimated via logistic regression or machine learning methods.
Rosenbaum and Rubin's balancing theorem states that if treatment assignment is ignorable given $X$---that is, $(Y(1), Y(0)) \perp D \mid X$---then it is also ignorable given the propensity score:
\begin{equation}
    (Y(1), Y(0)) \perp D \mid e(X)
\end{equation}

Several matching methods use propensity scores:
\begin{itemize}[leftmargin=15pt]
    \item \textbf{Nearest-neighbor matching}: Each treated unit is matched to the control unit(s) with the closest propensity score.
    \item \textbf{Caliper matching}: Matches are only accepted if the propensity score difference is below a threshold.
    \item \textbf{Stratification}: Units are grouped into strata based on propensity score quantiles; effects are estimated within each stratum and pooled.
    \item \textbf{Inverse probability weighting (IPW)}: Rather than matching, each unit is weighted by the inverse of its probability of receiving the treatment it actually received.
\end{itemize}

After matching, covariate balance is assessed by comparing means (or distributions) of covariates between matched treatment and control groups.
Standardized mean differences below 0.1 are often considered acceptable.
The treatment effect is then estimated as the difference in mean outcomes between matched groups.


\paragraph{Application in Software Engineering.}
Matching is well-suited for evaluating the effects of tool adoptions, process changes, or organizational decisions when randomization is infeasible.
Consider studying whether adopting a static analysis tool reduces bug rates.
Repositories that adopt static analysis may differ systematically from those that do not---perhaps they are larger, more active, maintained by more experienced teams, and place greater emphasis on quality assurance than the average repository.
These differences would confound any naive comparison.

A matching study would proceed as follows.
First, the researcher would identify treated repositories (those that adopted static analysis) and potential controls (those that did not).
Second, the researcher would estimate propensity scores using pre-treatment characteristics: repository age, programming language, number of contributors, prior commit frequency, prior bug rates, and any other available confounders.
Third, the researcher would match each treated repository to one or more control repositories with similar propensity scores.
Fourth, the researcher would verify that the matching achieved covariate balance.
Finally, the researcher would compare post-adoption bug rates between matched treatment and control groups.

\paragraph{Limitations.}
The critical limitation of matching is that it only adjusts for \emph{observed} covariates and rests on the assumption that, conditional on the matched covariates, treatment assignment is as good as random.
This is a strong assumption---if important confounders are unmeasured, matching does not eliminate their influence, and estimates remain biased.
No amount of statistical sophistication can adjust for variables you have not yet measured.

In software engineering research, this limitation is particularly relevant.
Repository data captures code and process metrics but often misses crucial factors such as developer skill, organizational support, team culture, project importance, and strategic decisions.
If projects adopt a tool because they face unusual challenges (and would have had worse outcomes regardless), matching on observable characteristics will not remove this selection bias.
If unobserved confounders threaten the validity of matching under the causal theory at hand, researchers should acknowledge this limitation, conduct sensitivity analyses, and interpret results as suggestive rather than definitive.
Sensitivity analysis methods---particularly Rosenbaum bounds---may also help assess how robust conclusions are to unobserved confounders.

\subsection{Fixed Effects and Panel Models}

\paragraph{Historical Context.}
Fixed effects estimation emerged from the econometrics of panel data---datasets tracking the same entities over multiple time periods.
The method has roots in early twentieth-century work on agricultural experiments and economic indices, but its modern formalization came in the 1960s and 1970s with the work of economists like Mundlak and Balestra~\cite{mundlak1978pooling, balestra1966pooling}.
The Hausman test provided a principled way to choose between fixed and random effects specifications~\cite{hausman1978specification}, and subsequent decades saw widespread adoption in labor economics, industrial organization, and policy evaluation.
Panel methods became standard tools for controlling for unobserved heterogeneity in settings where randomization is impossible.

\paragraph{Intuition.}
Fixed effects exploit a simple but powerful idea: If we observe the same unit (person, project, organization) over time, we can control for all characteristics of that unit that do not change over time, even if we cannot directly measure them.
Consider the challenge of estimating whether adopting TypeScript improves code quality.
Repositories using TypeScript might have better quality because TypeScript's type system catches bugs, or because repositories with disciplined development practices both adopt TypeScript and write better code.
Developer skill, organizational culture, and codebase complexity are difficult to measure but may confound comparisons.
Fixed effects address this by focusing on \emph{within-unit} variation.
Rather than comparing TypeScript repositories to JavaScript repositories, we compare each repository to itself over time: How did its code quality change when it transitioned from JavaScript to TypeScript?
By using each unit as its own control, we eliminate all time-invariant differences between units.
If a repository has especially skilled maintainers, this affects both its pre-TypeScript and post-TypeScript periods equally and cannot explain any change at the transition.
The key insight is that many confounders are time-invariant or change slowly.
Developer ability, project complexity, organizational resources---these factors may differ dramatically across projects but remain relatively stable within a project over the study period.
Fixed effects automatically control for all such factors without requiring us to find a meaningful metric to measure them.

\paragraph{Formal Framework.}
Consider a panel dataset with $N$ entities observed over $T$ periods.
Let $Y_{it}$ denote the outcome for entity $i$ at time $t$, $D_{it}$ the treatment indicator, and $X_{it}$ a vector of time-varying covariates.
The fixed effects model is:
\begin{equation}
    Y_{it} = \alpha + \beta D_{it} + X'_{it}\gamma + \mu_i + \epsilon_{it}
\end{equation}
where $\mu_i$ represents the fixed effect for entity $i$---a unit-specific intercept capturing all time-invariant characteristics.
Estimation proceeds by ``demeaning'' the data: subtracting each variable's entity-specific mean.
Define $\bar{Y}_i = \frac{1}{T}\sum_t Y_{it}$ and similarly for other variables.
The within-transformed model is:
\begin{equation}
    (Y_{it} - \bar{Y}_i) = \beta (D_{it} - \bar{D}_i) + (X_{it} - \bar{X}_i)'\gamma + (\epsilon_{it} - \bar{\epsilon}_i)
\end{equation}
The fixed effects $\mu_i$ cancel out, and ordinary least squares on the transformed data yields consistent estimates of $\beta$ and $\gamma$.
What's more, time fixed effects can be added to control for common shocks affecting all units:
\begin{equation}
    Y_{it} = \alpha + \beta D_{it} + X'_{it}\gamma + \mu_i + \lambda_t + \epsilon_{it}
\end{equation}
This ``two-way fixed effects'' specification controls for both unit-specific and time-specific unobserved heterogeneity.
It is the workhorse model for panel data analysis and underlies the standard difference-in-differences estimator introduced before in Section~\ref{sec:chap2-did}.

\paragraph{Limitations.}
Fixed effects control only for \emph{time-invariant} confounders.
If treatment adoption correlates with time-varying unobservables---factors that change over time and affect both treatment and outcomes---then estimates remain biased.
Consider a repository that adopts a new tool precisely when it receives a funding boost that also increases contributor activity.
Fixed effects cannot disentangle the tool's effect from the funding's effect, because the funding-related improvement varies over time.
A second limitation is that fixed effects provide no built-in falsification check.
Unlike difference-in-differences, which allows testing parallel pre-trends, pure fixed effects estimation offers no clear diagnostic for whether the identifying assumption (no time-varying confounding) holds.
This is why researchers often prefer DiD specifications that explicitly examine pre-treatment dynamics.
Third, fixed effects discard between-unit variation, using only within-unit changes for identification.
This can reduce statistical power, particularly if treatment status varies little within units.
If most repositories either always have CI or never have CI, with few transitions, fixed effects estimation may be imprecise.

\paragraph{Application in Software Engineering.}
Fixed effects are appropriate when studying how outcomes change as repositories or developers adopt new practices, tools, or policies.
Consider evaluating whether mandatory code review improves code quality.
At some point, an organization mandates code review for all pull requests; we observe the same repositories before and after the mandate.
A fixed-effects analysis would estimate the change in code quality associated with the mandate, controlling for repository fixed effects (which absorb stable differences between repositories) and time fixed effects (which absorb industry-wide trends in quality).
The treatment effect is identified from within-repository changes at the time of the mandate.
The identifying assumption is that no other time-varying factors coincided with the mandate.
If the organization simultaneously improved its CI pipeline, hired senior engineers, or changed leadership, these confound the estimate.
Researchers should examine whether the mandate was exogenous to other changes and, where possible, look for falsification tests---for example, whether effects appear before the mandate (which would suggest confounding) or whether unrelated outcomes change (which would suggest broader organizational shifts).

\subsection{Generalized Method of Moments}

\paragraph{Historical Context.}
The generalized method of moments (GMM) was formalized by Lars Peter Hansen in his 1982 \emph{Econometrica} paper~\cite{hansen1982large}, work that contributed to his 2013 Nobel Prize in Economics.
GMM unified earlier estimation approaches---method of moments, instrumental variables, maximum likelihood---within a single framework based on moment conditions.
For panel data with dynamic relationships, the Arellano-Bond estimator (1991) applied GMM to handle lagged dependent variables and endogeneity~\cite{arellano1991some}.
This approach became central to empirical work in corporate finance, growth economics, and any field where the central problem of interest can be formalized as: Do the current outcomes depend on past outcomes?

\paragraph{Intuition.}
Many phenomena exhibit dynamic dependence: Today's code quality depends on yesterday's code quality; this month's contributor activity depends on last month's activity.
Standard regression methods struggle with such settings because the lagged dependent variable is correlated with the error term in fixed effects models, leading to biased estimates.

GMM addresses this by using lagged values of the dependent variable as instrumental variables for the first-differenced model.
The intuition is that once we difference the data to remove fixed effects, lagged levels from earlier periods are correlated with the differenced lagged dependent variable (relevance) but uncorrelated with the differenced error term if errors are serially uncorrelated (exogeneity).
This provides valid instruments for consistent estimation.

Beyond handling dynamics, GMM is useful when researchers suspect bidirectional causality or want to test specific causal orderings.
Does productivity affect tool adoption, or does tool adoption affect productivity?
By specifying different lag structures and testing overidentifying restrictions, researchers can probe which causal directions the data support.

\paragraph{Formal Framework.}
Consider a dynamic panel model:
\begin{equation}
\label{eq:dynamic-gmm}
    Y_{it} = \rho Y_{i,t-1} + \beta D_{it} + X'_{it}\gamma + \mu_i + \lambda_t + \epsilon_{it}
\end{equation}
where $\rho$ captures persistence in the outcome, and $\beta$ is the treatment effect of interest.
The fixed effects $\mu_i$ create a correlation between $Y_{i,t-1}$ and the composite error $(\mu_i + \epsilon_{it})$, biasing OLS estimates. The following first-differencing eliminates the fixed effects:
\begin{equation}
    \Delta Y_{it} = \rho \Delta Y_{i,t-1} + \beta \Delta D_{it} + \Delta X'_{it}\gamma + \Delta \lambda_t + \Delta \epsilon_{it}
\end{equation}
But now $\Delta Y_{i,t-1} = Y_{i,t-1} - Y_{i,t-2}$ is correlated with $\Delta \epsilon_{it} = \epsilon_{it} - \epsilon_{i,t-1}$ through the common term $\epsilon_{i,t-1}$.
The Arellano-Bond approach uses lagged levels $Y_{i,t-2}, Y_{i,t-3}, \ldots$ as instruments for $\Delta Y_{i,t-1}$.
Under the assumption that errors are serially uncorrelated (sequential exogeneity), these lagged levels are uncorrelated with $\Delta \epsilon_{it}$ but correlated with $\Delta Y_{i,t-1}$.
The moment conditions are:
\begin{equation}
    E[Y_{i,t-s} \cdot \Delta \epsilon_{it}] = 0 \quad \text{for } s \geq 2
\end{equation}

GMM estimation combines all available moment conditions and optimally weights them.
The Sargan-Hansen test of overidentifying restrictions assesses whether the instruments are jointly valid.
The Arellano-Bond test for serial correlation checks whether errors are indeed uncorrelated at the required lags.
System GMM extends this by adding level equations alongside differenced equations, using lagged differences as instruments for levels.
This can improve efficiency, particularly when the outcome is persistent and levels carry substantial information.

\paragraph{Limitations.}
GMM requires careful specification and diagnostic testing.
Weak instruments---lagged values that have low correlation with differenced variables---lead to imprecise and potentially biased estimates.
Too many instruments (a common problem with long panels) can cause overfitting and make the Sargan-Hansen test unreliable.
The sequential exogeneity assumption---that past outcomes affect current outcomes only through the specified dynamic structure, not through omitted time-varying confounders---is fundamentally untestable.

GMM is also less transparent than DiD or fixed effects.
The instrumental variables logic is subtle, and results can be sensitive to the number of lags used as instruments, the weighting matrix, and whether system GMM or difference GMM is employed.
Thus, researchers should report extensive sensitivity checks and diagnostics when using GMM models.

\paragraph{Application in Software Engineering.}
GMM is appropriate when studying dynamic relationships in software development---settings where current outcomes depend on past outcomes, and simple before-after comparisons may be misleading.
Consider investigating whether adopting an AI coding assistant improves productivity.
Productivity is persistent: A repository that was highly productive last month is likely to be productive this month.
Failing to account for this persistence overstates the tool's effect if adoption coincided with an already-increasing productivity trend.

A GMM analysis would model current productivity as depending on lagged productivity and treatment status:
\begin{equation}
    \text{Productivity}_{it} = \rho \cdot \text{Productivity}_{i,t-1} + \beta \cdot \text{AI\_Tool}_{it} + \mu_i + \lambda_t + \epsilon_{it}
\end{equation}
Using lagged productivity levels as instruments for differenced lagged productivity, GMM provides consistent estimates of both the persistence parameter $\rho$ and the treatment effect $\beta$.

GMM can also help disentangle causal direction.
Does using an AI tool increase productivity, or do more productive developers adopt AI tools?
By testing whether lagged productivity predicts current tool use (and vice versa), and by examining whether effects appear before or after adoption, researchers can probe which causal story the data best support.
While GMM cannot definitively establish causation, it provides more rigorous evidence than simple correlational analyses and can rule out certain causal orderings.

In this thesis, GMM complements DiD by examining dynamic mechanisms through which tool adoption affects long-term outcomes (Chapter~\ref{chp:cursor}).
DiD estimates the overall effect of adoption; GMM decomposes this effect into immediate impacts versus persistent changes, helping distinguish genuine productivity gains from temporary adjustment effects.

\section{Threats to Validity in Causal Designs}
\label{sec:threats}

Every causal design, no matter how carefully constructed, faces threats that could undermine its conclusions.
The literature on research methodology distinguishes three fundamental types of validity: internal validity, which concerns whether the estimated effect reflects the true causal relationship within the study sample; external validity, which concerns whether findings generalize beyond the specific study context; and construct validity, which concerns whether the measured variables actually capture the theoretical constructs of interest.
Understanding these threats is essential both for designing rigorous studies and for critically evaluating the research of others.

\subsection{Internal Validity}

A study possesses high internal validity when we can confidently attribute observed outcome differences to the treatment rather than to confounding factors.
The fundamental challenge is that we never observe the counterfactual---what would have happened to treated units had they not been treated---and must instead infer it from comparison groups under assumptions that may or may not hold.

The most pervasive threat to internal validity is \emph{confounding}, which occurs when an unmeasured variable affects both treatment assignment and outcomes, creating a spurious association that masquerades as a causal effect.
Consider a study examining whether adopting a new development tool improves code quality.
If repositories with more experienced maintainers are both more likely to adopt innovative tools and more likely to produce high-quality code regardless of the tool, the observed association between tool adoption and quality may reflect maintainer expertise rather than the tool's genuine effect.
Randomized experiments eliminate confounding by construction, since random assignment ensures that treatment and control groups are, in expectation, identical on all characteristics---both observed and unobserved.
Observational designs lack this guarantee and must either measure and adjust for potential confounders or rely on design-based strategies such as difference-in-differences, regression discontinuity, or instrumental variables that address confounding through alternative identifying assumptions.

Closely related to confounding is \emph{selection bias}, which arises when the process determining who enters the study or who receives treatment is systematically related to potential outcomes.
If only repositories already experiencing quality problems adopt a static analysis tool (hoping it will help), naive comparisons of adopters to non-adopters will conflate the tool's effect with the pre-existing quality differential.
Selection can also manifest as differential attrition: if struggling projects are more likely to cease activity and drop out of a longitudinal study, the surviving sample becomes increasingly unrepresentative, and estimates based on this sample may be severely biased.

Even with perfect identification of causal effects, \emph{measurement error} in treatments or outcomes can attenuate or distort estimates.
When treatment status is measured imprecisely---for example, inferring ``tool adoption'' from the presence of a configuration file that may not indicate actual sustained use---the resulting measurement error typically biases estimates toward zero, making true effects appear smaller than they are.
When outcomes are measured with systematic error correlated with treatment status, bias can operate in either direction and is considerably harder to diagnose.

Finally, the Stable Unit Treatment Value Assumption (SUTVA) requires that one unit's treatment status does not affect another unit's outcomes---an assumption frequently violated in software engineering contexts.
Developers share knowledge across teams, projects share dependencies through package ecosystems, and practices diffuse through professional networks and organizations.
If adopting a tool in one repository indirectly improves code quality in related repositories through shared learning or code reuse, simple treatment-control comparisons will underestimate the tool's total effect by ignoring these spillovers.
Alternatively, if control units are ``contaminated'' by exposure to treatment effects (perhaps through documentation or community discussions), the estimated treatment effect will be biased toward zero.

\subsection{External Validity}

A study can achieve perfect internal validity---correctly identifying the causal effect within its specific sample---yet provide limited guidance for other populations, settings, or time periods.
External validity concerns this generalizability, and its threats are often underappreciated in fields that prioritize identification over representativeness.

The most fundamental external validity concern is \emph{population generalizability}: effects estimated on one population may not apply to others with different characteristics.
Studies conducted on open-source GitHub repositories may not generalize to proprietary enterprise development, where organizational constraints, incentive structures, and development practices differ substantially.
Effects observed among early adopters of a technology---who tend to be more technically sophisticated and intrinsically motivated---may not hold for the broader population of later adopters.
Volunteer participants in controlled experiments may differ systematically from developers who would never consent to participate, limiting the applicability of experimental findings.

\emph{Temporal generalizability} presents particular challenges in rapidly evolving technological domains.
The productivity impact of early AI coding assistants, measured with 2023-era tools on 2023-era codebases, may bear little resemblance to the effects of more advanced tools deployed in the late 2020s.
Best practices validated through empirical research in one technological era may become obsolete---or even counterproductive---as the underlying technologies evolve.
This suggests that empirical software engineering findings should be understood as historically situated rather than as timeless truths.

Beyond population and temporal concerns, causal effects often exhibit substantial \emph{context dependence}, varying with organizational culture, developer expertise, project characteristics, programming language, or other factors in ways that limit straightforward transferability.
A tool that accelerates novice developers may impose coordination overhead that slows experts; a practice that enhances productivity in small, co-located teams may create bottlenecks at enterprise scale.
Recognizing this heterogeneity, researchers increasingly report not just average treatment effects but also how effects vary across observable subgroups, since understanding who benefits and who does not is often more actionable than knowing only the population average.

\subsection{Construct Validity}

Beyond the question of whether we have correctly identified a causal effect (internal validity) and whether it generalizes (external validity), construct validity asks whether the variables we measure actually capture the theoretical constructs we care about.
This challenge is particularly acute in software engineering, where many concepts of interest---productivity, code quality, technical debt, developer experience---are multidimensional abstractions that resist direct measurement.

The operationalization of treatments shapes what effects we estimate.
``Adopting continuous integration'' might be operationalized as setting up a CI system, actually running automated tests on every commit, or promptly responding to test failures---each capturing different aspects of the underlying practice.
Similarly, ``code quality'' might be measured through defect density, code review feedback, static analysis warnings, or maintainability indices, each reflecting different quality dimensions and potentially yielding different conclusions about treatment effects.

Even well-defined metrics may suffer from measurement imprecision.
Commit counts, used as a productivity proxy, conflate substantive work with commit granularity conventions that vary across developers and projects.
Issue counts conflate bugs with feature requests, user-reported problems with automated detections, and genuine defects with documentation gaps.
Lines of code added confounds productive output with code bloat.
These measurement limitations do not invalidate empirical research, but they counsel humility about the precision of our conclusions and underscore the value of triangulating across multiple operationalizations.

\subsection{Sensitivity Analysis and Robustness}

Because the identifying assumptions underlying causal inference are fundamentally untestable---we cannot directly verify that parallel trends would have held or that an instrument satisfies the exclusion restriction---rigorous research practice demands sensitivity analysis that explores how conclusions would change if assumptions were violated.

For matching and propensity score studies, Rosenbaum bounds quantify how strong an unmeasured confounder would need to be to explain away the observed effect~\cite{rosenbaum2002observational}.
If results remain robust to confounders implausibly stronger than any known variable, confidence in the findings increases; if a moderately strong confounder could reverse conclusions, appropriate caution is warranted.
For difference-in-differences, researchers examine sensitivity to pre-trend specifications: event study plots visualizing pre-treatment coefficients help assess whether the parallel trends assumption is plausible, and researchers increasingly test whether conclusions survive alternative trend extrapolations.

A powerful approach to robustness is employing multiple estimators with different identifying assumptions.
When two-way fixed effects, Callaway-Sant'Anna, and imputation estimators all yield consistent treatment effect estimates, the finding is unlikely to be an artifact of any single method's assumptions.
Conversely, divergent results across estimators signal that conclusions may depend on methodological choices and warrant careful investigation.

Placebo tests provide another diagnostic tool, examining whether effects appear where they should not.
If treatment effects appear \emph{before} treatment occurs, this suggests confounding rather than causation.
If treatment affects outcomes it theoretically should not influence---such as pre-determined characteristics fixed before treatment assignment---something is amiss with the research design.
Passing placebo tests does not prove validity, but failing them is highly informative about potential threats.

Ultimately, the credibility of causal claims rests not on asserting that assumptions are satisfied---a claim that can never be definitively verified---but on transparently examining their plausibility, demonstrating robustness to reasonable violations, and acknowledging the limitations that remain.

\section{A Pragmatic Approach to Causal Inference in Empirical Software Engineering}
\label{sec:pragmatic}

The preceding sections have presented causal inference with appropriate rigor, emphasizing the assumptions underlying each method and the threats that can undermine conclusions.
This rigor is essential---but it can also be paralyzing.
If every observational study suffers from potential confounding, if every natural experiment has debatable assumptions, and if even randomized experiments face external validity concerns, can we learn anything useful?

This section argues that we can, provided we approach causal inference pragmatically.
The goal is not epistemological perfection but actionable evidence that improves upon intuition and correlational analysis.
Software engineering researchers must navigate trade-offs thoughtfully, choosing appropriate methods for their contexts and communicating uncertainty honestly.

\subsection{The Internal-External Validity Trade-off}

The most fundamental trade-off in empirical research is between internal validity (did we correctly identify the causal effect in our study?) and external validity (does the effect generalize to settings we care about?).
These two goals are often in tension, and no study achieves perfection on both.

Consider a randomized controlled trial evaluating an AI coding assistant.
If we recruit volunteer participants, run the study in a controlled environment, and measure carefully defined tasks, we may achieve excellent internal validity---we know the tool caused the observed productivity difference.
But the participants may not represent typical developers; the tasks may not reflect real work; the environment may not capture the pressures and interruptions of actual practice.
The clean causal estimate may not generalize to contexts that matter.

Conversely, a large-scale observational study of AI tool adoption across thousands of GitHub repositories captures real-world variation and naturally represents diverse contexts.
But without randomization, we cannot be certain that observed productivity differences reflect the tool's effect rather than selection---perhaps more capable developers adopted the tool first because they are also more tech-savvy.
The estimate may be externally valid (it reflects what happens in the real world), but internally questionable (we do not know if the tool caused it).

This trade-off is not a flaw to be fixed but a reality to be navigated.
The goal is not to find the one perfect study but to build knowledge through multiple studies with different strengths, whose convergent findings provide confidence that transcends any single study's limitations.

\subsection{A Hierarchy of Evidence}

Rather than treating causal inference as all-or-nothing, it is useful to think of a spectrum of evidence quality.
Different research questions warrant different standards, and different contexts permit different designs.
At the top of the hierarchy sit randomized experiments with representative samples.
These studies provide strong internal validity and external validity when samples are representative.
Below this are natural experiments and quasi-experimental designs: regression discontinuity, instrumental variables, and well-executed difference-in-differences studies.
These achieve internal validity under specific assumptions that may be partially testable (e.g., parallel pre-trends, no manipulation at thresholds).
External validity depends on how representative the setting is.
Next come observational studies with careful control strategies: matching, propensity scores, fixed effects, and theory-driven covariate adjustment.
These reduce confounding but cannot eliminate it; internal validity depends on untestable assumptions about unobserved confounders.
At the bottom are descriptive and correlational studies, which document patterns but make no causal claims.
These are valuable for exploration, hypothesis generation, and understanding prevalence, but they cannot answer ``what if'' questions about interventions.

\paragraph{What the Hierarchy Does Not Mean.}
The hierarchy does not imply that lower-tier studies are worthless or that only randomized experiments deserve publication.
A thoughtful observational study of real-world tool adoption may provide more useful guidance than a randomized experiment on a contrived task with unrepresentative participants.
The hierarchy is about the strength of \emph{causal} evidence, not overall scientific value.
The hierarchy is also not about methodological purity.
A DiD study with obviously violated parallel trends is worse than a careful descriptive study that does not overclaim.
Execution quality matters as much as design choice.

\subsection{Principles for Pragmatic Practice}

Given these trade-offs, how should software engineering researchers approach causal inference in practice?
The answer lies not in a rigid hierarchy of methods but in thoughtful matching of research designs to research questions, contexts, and available resources.

The first principle is to \emph{match the design to the question and context}.
Not every research question requires---or can support---the strongest possible causal design.
For exploratory research aimed at generating hypotheses and identifying promising directions, descriptive and correlational studies remain entirely appropriate.
For consequential decisions with significant resource implications---should an organization mandate a particular tool? should a platform change its policies?---stronger causal evidence is warranted to justify the costs and risks of action.
The practical question is what designs are feasible given the specific setting: Is randomization possible, or would it be unethical or impractical? Does the context provide a natural experiment through policy discontinuities or staggered adoption? What data are available, and with what quality?
Researchers should choose the strongest design that remains credible in their specific context, rather than forcing an inappropriate method that offers theoretically stronger causal identification but rests on implausible assumptions.

The second principle is \emph{transparency about assumptions and limitations}.
Every causal claim, whether from an RCT or an observational study, rests on assumptions that cannot be definitively verified.
Rigorous research practice demands that these assumptions be stated explicitly, their plausibility discussed in the specific research context, and potential violations acknowledged forthrightly.
Transparency about limitations does not weaken a research contribution; rather, it strengthens it by demonstrating that the authors understand the nature of the evidence they are providing and the conditions under which their conclusions hold.

Third, because identifying assumptions are fundamentally untestable, researchers should \emph{conduct systematic sensitivity analysis} exploring how conclusions would change if assumptions were violated.
For matching studies, Rosenbaum bounds quantify the strength of unmeasured confounding required to overturn findings.
For difference-in-differences, researchers should examine robustness to alternative pre-trend specifications and conduct event study analyses.
Applying multiple estimators with different identifying assumptions---and reporting whether they yield consistent conclusions---provides additional evidence about the robustness of findings.
When conclusions survive these diagnostic checks, confidence in their validity grows; when they prove fragile, appropriate caution and qualification are warranted.

Fourth, researchers should \emph{embrace triangulation} as a strategy for building cumulative knowledge.
No single study, however well-designed and carefully executed, provides definitive evidence about causal relationships.
Scientific knowledge accumulates through replication across different contexts, methods, and research teams.
When randomized experiments, quasi-experimental observational studies, and qualitative research all point in the same direction, our collective confidence in the underlying causal relationship grows substantially.
When different approaches yield divergent conclusions, the discrepancy itself is informative: perhaps effects are genuinely context-dependent, varying across populations or settings in ways that merit investigation; or perhaps some studies contain unrecognized biases that others avoid.
Either way, the apparent contradiction advances understanding more than any single study could.

Fifth, researchers should \emph{report effect heterogeneity} rather than focusing exclusively on average treatment effects.
Population averages often conceal substantial variation across subgroups that is more actionable than the overall mean.
A tool that improves productivity by 10\% on average might help novices by 30\% while actually slowing experienced developers by 10\%---information far more useful to practitioners than the aggregate figure alone.
Examining how effects vary across observable characteristics---developer experience, project type, programming language, organizational context---provides insights that help readers assess whether findings are likely to generalize to their own situations.

Finally, researchers should \emph{communicate uncertainty honestly}.
Statistical significance is neither equivalent to practical importance nor a guarantee of correctness.
Reporting confidence intervals conveys the precision of estimates; discussing effect magnitudes in substantively meaningful terms helps readers evaluate practical relevance; acknowledging when evidence is suggestive rather than definitive maintains scientific integrity.
Overclaiming---presenting tentative findings as established facts---undermines the credibility of individual researchers and the field as a whole.
Appropriate epistemic humility, by contrast, builds trust and enables research to inform practice even when certainty remains elusive.

\subsection{The Value of Imperfect Evidence}

Given the stringent assumptions underlying causal inference and the many threats to validity that observational studies face, it might be tempting to conclude that without a perfect randomized experiment, we cannot learn anything meaningful about causation.
This conclusion, while superficially rigorous, is ultimately too pessimistic to serve as a guide for research practice.
Imperfect evidence, accumulated thoughtfully across multiple studies and methods, can guide decisions far better than no evidence at all or purely intuitive reasoning based on anecdote and conventional wisdom.

Consider a practical decision: Should a software organization adopt a new static analysis tool?
A correlational study finds that repositories currently using the tool exhibit 20\% fewer bugs than those that do not.
This association, while suggestive, is not causal evidence---perhaps better-organized teams both adopt sophisticated tooling and write higher-quality code, with no genuine effect of the tool itself.
But suppose a difference-in-differences study exploiting staggered adoption finds that bug rates drop by 15\% after repositories adopt the tool, with no evidence of differential pre-trends.
A propensity score matching study comparing adopters to similar non-adopters finds a 12\% reduction.
Qualitative interviews with developers reveal that the tool catches common coding mistakes that previously slipped through review.

No single study in this portfolio is methodologically definitive.
The DiD study might suffer from time-varying confounders; the matching study cannot account for unobserved selection; the interviews provide mechanism but not magnitude.
Yet together, these studies suggest that the tool probably provides genuine benefits, though the effect size might be smaller than naive correlations suggest and likely varies across contexts and development practices.
This is useful guidance for organizational decision-making---not certainty, but evidence that substantially improves upon intuition alone.

The goal of causal inference in software engineering is not to achieve epistemological perfection---an unattainable standard in any empirical science---but to provide better evidence than we would have otherwise.
By employing the most rigorous designs feasible in each context, maintaining transparency about assumptions and limitations, conducting systematic sensitivity analyses, and triangulating across complementary studies, researchers can contribute genuine knowledge to a field where randomized experiments remain rare and observational data inevitably imperfect.
The credibility of causal claims ultimately rests not on claiming that assumptions are satisfied, but on transparently examining their plausibility, demonstrating robustness to reasonable violations, and honestly communicating what the evidence can---and cannot---tell us about the phenomena under investigation.
