\documentclass[12pt, openany]{cmuthesis}

\usepackage{mathptmx} 
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage[skins]{tcolorbox}
\usepackage{fontawesome}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{calc}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{minted}
\usepackage{balance}
\usepackage[backend=biber, style=numeric]{biblatex}\let\citet\textcite
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
\usepackage[
    pageanchor=true,
    plainpages=false,
    pdfpagelabels, 
    bookmarks,
    bookmarksnumbered, 
    hidelinks
]{hyperref}

%%%%%%%%%% Begin: JSON Syntax Definition %%%%%%%%%%
\makeatletter
% switch used as state variable
\newif\ifcolonfoundonthisline
\newcommand\JSONnumbervaluestyle{\color{blue!80!black}}
\newcommand\JSONstringvaluestyle{\color{red!80!black}}
\lstdefinestyle{json}{
  showstringspaces    = false,
  keywords            = {false,true},
  alsoletter          = 0123456789.,
  morestring          = [s]{"}{"},
  stringstyle         = \ifcolonfoundonthisline\JSONstringvaluestyle\fi,
  MoreSelectCharTable = \lst@DefSaveDef{`:}\colon@json{\processColon@json},
  basicstyle          = \footnotesize\ttfamily\linespread{0.5},
  keywordstyle        = \ttfamily\bfseries,
  numbers             = left,
  numberstyle         = \scriptsize,
  stepnumber          = 1,
  numbersep           = 6pt,
  showstringspaces    = false,
  breaklines          = true,
  frame               = lines,
  escapeinside        = {(*}{*)}
}
% flip the switch if a colon is found in Pmode
\newcommand\processColon@json{%
  \colon@json%
  \ifnum\lst@mode=\lst@Pmode%
    \global\colonfoundonthislinetrue%
  \fi
}
\lst@AddToHook{Output}{%
  \ifcolonfoundonthisline%
    \ifnum\lst@mode=\lst@Pmode%
      \def\lst@thestyle{\JSONnumbervaluestyle}%
    \fi
  \fi
  %override by keyword style if a keyword is detected!
  \lsthk@DetectKeywords% 
}
% reset the switch at the end of line
\lst@AddToHook{EOL}{\global\colonfoundonthislinefalse}
%%%%%%%%%% End:   JSON Syntax Definition %%%%%%%%%%

%%%%%%%%%% Begin: Finding Box Definition %%%%%%%%%%
\tcbset{
  my box2/.style={
    enhanced,
    colframe=#1!80,
    colback=#1!5,
  },
}
\newtcolorbox{finding}{
  my box2=black,
  boxrule=1pt,top=3pt,bottom=3pt,left=4pt,right=4pt
}
%%%%%%%%%% End:   Finding Box Definition %%%%%%%%%%

%%%%%%%%%% Begin: Custom Commands and Colors %%%%%%%%%%
\renewcommand*{\ttdefault}{lmtt} % Change texttt font

\newcommand{\TODO}[1]{{\textbf{\color{red}[[TODO: #1]]}}}
\newcommand{\Code}[1]{\texttt{#1}}
\newcommand{\SmallCode}[1]{\begin{small}\texttt{#1}\end{small}}
\newcommand{\SystemName}[0]{\texttt{StarScout}\xspace}
\newcommand{\RQOne}[0]{\textbf{RQ1} (\textsf{Prevalance}): \emph{How prevalent are fake GitHub stars?}\xspace}
\newcommand{\RQTwo}[0]{\textbf{RQ2} (\textsf{Activity Patterns}): \emph{What are the activity patterns of GitHub repositories and accounts with fake star campaigns?}}
\newcommand{\RQThree}[0]{\textbf{RQ3} (\textsf{Repository Charactersitics}): \emph{What are the GitHub repositories with fake star campaigns?}}
\newcommand{\RQFour}[0]{\textbf{RQ4} (\textsf{Promotional Effect}): \emph{To what extent are fake stars effective in promoting the target GitHub repositories?}}

\definecolor{darkred}{rgb}{0.9, 0.0, 0.0}
\definecolor{darkgreen}{rgb}{0.0, 0.8, 0.0}
\definecolor{amber}{rgb}{1.0, 0.75, 0.0}
%%%%%%%%%% End: Custom Commands and Colors %%%%%%%%%%

\addbibresource{references-pinning.bib}
\addbibresource{references-fake-stars.bib}
\addbibresource{references-cursor.bib} 

\begin{document} 
\frontmatter

\pagestyle{empty}

\title{{\it Thesis Proposal}\\
{\bf Settling Software Engineering Controversies through Robust Causal Inference Methods
}}
\author{Hao He}
\date{January 2026}
\Year{2026}
\trnumber{}

\committee{
Bogdan Vasilescu, Co-Chair \\
Christian Kästner, Co-Chair \\
Rohan Padhye \\
Narayan Ramasubbu
}

\support{}
\disclaimer{}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

%\keywords{Stuff, More Stuff}

\maketitle

%\begin{dedication}
%For my dog
%\end{dedication}

\pagestyle{plain} % for toc, was empty

\begin{abstract}
Software engineering practice is plagued with controversies and unverified beliefs about best practices. 
Unfortunately, empirical software engineering research often fails to resolve these controversies if it relies solely on descriptive or correlational evidence without distinguishing causation from confounding factors. 
In this thesis, I propose that \emph{software engineering controversies can be resolved with rigorous causal inference methods}, which are particularly useful in providing confidence and avoiding false positives in empirical software engineering findings.

Through three empirical studies, I demonstrate how the appropriate use of robust causal inference methods provides solid, yet surprising, answers to important software engineering problems. 
First, using counterfactual analysis and simulations on npm dependency data, I challenge the conventional wisdom that pinning dependency versions improves overall project security, revealing that it paradoxically increases attack exposure as long as upstream packages are not pinning their own dependencies. 
Second, through a longitudinal analysis of GitHub starring patterns, I uncover a massive inflation of fake GitHub stars and show that these artificial signals provide only short-term promotional benefits before becoming long-term liabilities. 
Finally, through a difference-in-differences framework, I quantify the causal impact of a widely popular AI Coding tool, Cursor, on software development velocity and code quality, finding that the initial velocity gains are offset by sustained technical debt accumulation, which eventually erodes velocity.
Collectively, these studies demonstrate that the application of robust causal inference methods moves empirical software engineering research toward trustworthy evidence and concrete practitioner guidance for navigating complex software engineering decisions.

Building on these contributions, I propose to further explore whether the use of robust causal inference methods leads to \emph{different} findings compared to the use of only descriptive or correlational evidence.
Consequently, this thesis would provide a new standard, a methodological blueprint, and concrete guidance for the design and implementation of future empirical software engineering research.
\end{abstract}

%\begin{acknowledgments}
%My advisor is cool.
%\end{acknowledgments}

\tableofcontents

%\listoffigures

%\listoftables

\mainmatter

\chapter{Introduction}

Motivation: The Prevalence of Myths and Unverified Beliefs in Software Engineering  
The Problem: Correlation vs. Causation in SE Research
Thesis Statement and Research Questions 
Overview of Approach: Three Case Studies  
Contributions  
Thesis Organization  

\chapter{Background and Related Work}

1.  Causal Inference in Observational Studies    
- The Fundamental Problem of Causal Inference    
-  Potential Outcomes Framework    
- Common Pitfalls in Correlational SE Research    
2. Causal Inference Methodologies    
- Counterfactual Analysis and Simulation    
- Difference-in-Differences Estimation    
- Propensity Score Matching    
- Panel Generalized Method of Moments    
- Threats to Causal Validity    
3. Empirical Software Engineering Research     
- Controlled Experiments  
- Large-Scale Observational Studies    
- The Need for Causal Methods on Observational Data  
4. Summary

\chapter{Empirical Study I:  Dependency Pinning and Software Supply Chain Security}

\include{main-pinning}

\chapter{Empirical Study II: Fake GitHub Stars and Repository Promotion}

\include{main-fake-stars}

\chapter{Empirical Study III: LLM Agent Assistants and the Velocity \& Quality Tradeoff}

- He, Hao, et al. "Speed at the Cost of Quality? The Impact of LLM Agent Assistance on Software Development." arXiv preprint arXiv:2511.04427 (2025).

\chapter{Proposed Work}

1. Strengthening Causal Evidence in the Fake Star Study (GMM Analysis)    
2. Sensitivity Analysis: DiD vs. Alternative Designs in the Cursor Study (Importance of causal inference techniques in )

\chapter{Conclusion}

1. Summary of Contributions    
2. Revisiting the Thesis Statement    
3. Final Thoughts

%\appendix
%\include{appendix}

\backmatter

\chapter{\bibname}

\printbibliography

\end{document}