```{r}
library(plm)
library(dplyr)
library(tidyr)
library(nlme)
library(tseries)
library(MuMIn)
library(stargazer)
library(texreg)
```

```{r}
repos = read.csv("data/model_stars.csv")

l1_repos <- repos %>%
  group_by(repo) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l1_fake = lag(n_stars_fake, 1),
    total_real = cumsum(n_stars_real),
    l2_total_real = lag(total_real, 2),
    total_fake = cumsum(n_stars_fake),
    l2_total_fake = lag(total_fake, 2),
  ) %>%
  ungroup() %>% 
  drop_na()

l2_repos <- repos %>%
  group_by(repo) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l2_real = lag(n_stars_real, 2),
    l1_fake = lag(n_stars_fake, 1),
    l2_fake = lag(n_stars_fake, 2),
    total_real = cumsum(n_stars_real),
    l3_total_real = lag(total_real, 3),
    total_fake = cumsum(n_stars_fake),
    l3_total_fake = lag(total_fake, 3)
  ) %>%
  ungroup() %>% 
  drop_na()

l3_repos <- repos %>%
  group_by(repo) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l2_real = lag(n_stars_real, 2),
    l3_real = lag(n_stars_real, 3),
    l1_fake = lag(n_stars_fake, 1),
    l2_fake = lag(n_stars_fake, 2),
    l3_fake = lag(n_stars_fake, 3),
    total_real = cumsum(n_stars_real),
    l4_total_real = lag(total_real, 3),
    total_fake = cumsum(n_stars_fake),
    l4_total_fake = lag(total_fake, 3),
  ) %>%
  ungroup() %>% 
  drop_na()

l4_repos <- repos %>%
  group_by(repo) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l2_real = lag(n_stars_real, 2),
    l3_real = lag(n_stars_real, 3),
    l4_real = lag(n_stars_real, 4),
    l1_fake = lag(n_stars_fake, 1),
    l2_fake = lag(n_stars_fake, 2),
    l3_fake = lag(n_stars_fake, 3),
    l4_fake = lag(n_stars_fake, 4),
    total_real = cumsum(n_stars_real),
    l5_total_real = lag(total_real, 5),
    total_fake = cumsum(n_stars_fake),
    l5_total_fake = lag(total_fake, 5)
  ) %>%
  ungroup() %>% 
  drop_na()

l5_repos <- repos %>%
  group_by(repo) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l2_real = lag(n_stars_real, 2),
    l3_real = lag(n_stars_real, 3),
    l4_real = lag(n_stars_real, 4),
    l5_real = lag(n_stars_real, 5),
    l1_fake = lag(n_stars_fake, 1),
    l2_fake = lag(n_stars_fake, 2),
    l3_fake = lag(n_stars_fake, 3),
    l4_fake = lag(n_stars_fake, 4),
    l5_fake = lag(n_stars_fake, 5),
    total_real = cumsum(n_stars_real),
    l6_total_real = lag(total_real, 6),
    total_fake = cumsum(n_stars_fake),
    l6_total_fake = lag(total_fake, 6)
  ) %>%
  ungroup() %>% 
  drop_na()

l6_repos <- repos %>%
  group_by(repo) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l2_real = lag(n_stars_real, 2),
    l3_real = lag(n_stars_real, 3),
    l4_real = lag(n_stars_real, 4),
    l5_real = lag(n_stars_real, 5),
    l6_real = lag(n_stars_real, 6),
    l1_fake = lag(n_stars_fake, 1),
    l2_fake = lag(n_stars_fake, 2),
    l3_fake = lag(n_stars_fake, 3),
    l4_fake = lag(n_stars_fake, 4),
    l5_fake = lag(n_stars_fake, 5),
    l6_fake = lag(n_stars_fake, 6),
    total_real = cumsum(n_stars_real),
    l7_total_real = lag(total_real, 7),
    total_fake = cumsum(n_stars_fake),
    l7_total_fake = lag(total_fake, 7)
  ) %>%
  ungroup() %>% 
  drop_na()
```

```{r}
ar1_stars_fe = plm(log(n_stars_real + 1) ~ log(l1_real + 1) + log(l1_fake + 1) + log(l2_total_real + 1) + log(l2_total_fake + 1), 
  data=l1_repos, index=c("repo", "month"), model="within")
summary(ar1_stars_fe)
```

```{r}
ar1_stars_re = plm(
  log(n_stars_real + 1) ~ 
    log(l1_real + 1) + log(l1_fake + 1) + 
    log(l2_total_real + 1) + log(l2_total_fake + 1)
    + age + had_release + log(activity + 1), 
  data=l1_repos, index=c("repo", "month"), model="random")
summary(ar1_stars_re)
```

```{r}
phtest(ar1_stars_fe, ar1_stars_re)
```

```{r}
ar2_stars_fe = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) + 
                     log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1), 
  data=l2_repos, index=c("repo", "month"), model="within")
summary(ar2_stars_fe)
```

```{r}
ar2_stars_re = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) 
                     + log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1)
                     + age + had_release + log(activity + 1), 
  data=l2_repos, index=c("repo", "month"), model="random")
summary(ar2_stars_re)
```

```{r}
phtest(ar2_stars_fe, ar2_stars_re)
```

```{r}
ar3_stars_fe = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_total_real + 1) + 
                     log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_total_fake + 1), 
  data=l3_repos, index=c("repo", "month"), model="within")
summary(ar3_stars_fe)
```

```{r}
ar3_stars_re = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_total_real + 1) 
                     + log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_total_fake + 1)
                     + age + had_release + log(activity + 1), 
  data=l3_repos, index=c("repo", "month"), model="random")
summary(ar3_stars_re)
```


```{r}
ar4_stars_fe = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_real + 1) + log(l5_total_real + 1) + 
                     log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_fake + 1) + log(l5_total_fake + 1), 
  data=l4_repos, index=c("repo", "month"), model="within")
summary(ar4_stars_fe)
```
```{r}
ar4_stars_re = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_real + 1) + log(l5_total_real + 1) 
                     + log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_fake + 1) + log(l5_total_fake + 1)
                     + age + had_release + log(activity + 1), 
  data=l4_repos, index=c("repo", "month"), model="random")
summary(ar4_stars_re)
```
```{r}
ar5_stars_fe = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_real + 1) + log(l5_real + 1) + log(l6_total_real + 1) + 
                     log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_fake + 1) + log(l5_fake + 1) + log(l6_total_fake + 1), 
  data=l5_repos, index=c("repo", "month"), model="within")
summary(ar5_stars_fe)
```
```{r}
ar5_stars_re = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_real + 1) + log(l5_real + 1) + log(l6_total_real + 1) 
                     + log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_fake + 1) + log(l5_fake + 1) + log(l6_total_fake + 1) 
                     + age + had_release + log(activity + 1), 
  data=l5_repos, index=c("repo", "month"), model="random")
summary(ar5_stars_re)
```
```{r}
ar6_stars_fe = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_real + 1) + log(l5_real + 1) + log(l6_real + 1) + log(l7_total_real + 1) + 
                     log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_fake + 1) + log(l5_fake + 1) + log(l6_fake + 1) + log(l7_total_fake + 1), 
  data=l6_repos, index=c("repo", "month"), model="within")
summary(ar6_stars_fe)
```

```{r}
ar6_stars_re = plm(log(n_stars_real + 1) ~ 
                     log(l1_real + 1) + log(l2_real + 1) + log(l3_real + 1) + log(l4_real + 1) + log(l5_real + 1) + log(l6_real + 1) + log(l7_total_real + 1)
                     + log(l1_fake + 1) + log(l2_fake + 1) + log(l3_fake + 1) + log(l4_fake + 1) + log(l5_fake + 1) + log(l6_fake + 1) + log(l7_total_fake + 1)
                     + age + had_release + log(activity + 1), 
  data=l6_repos, index=c("repo", "month"), model="random")
summary(ar6_stars_re)
```

```{r}
aicbic_plm <- function(object, criterion) {
  
  
  # object is "plm", "panelmodel" 
  # Lets panel data has index :index = c("Country", "Time")
  
  sp = summary(object)
  
  if(class(object)[1]=="plm"){
    u.hat <- residuals(sp) # extract residuals
    df <- cbind(as.vector(u.hat), attr(u.hat, "index"))
    names(df) <- c("resid", "repo", "month")
    c = length(levels(df$Country)) # extract country dimension 
t = length(levels(df$Time)) # extract time dimension 
    np = length(sp$coefficients[,1]) # number of parameters
n.N = nrow(sp$model) # number of data
    s.sq  <- log( (sum(u.hat^2)/(n.N))) # log sum of squares
    
    # effect = c("individual", "time", "twoways", "nested"),
    # model = c("within", "random", "ht", "between", "pooling", "fd")
    
    # I am making example only with some of the versions:
    
    if (sp$args$model == "within" & sp$args$effect == "individual"){
      n = c
      np = np+n+1 # update number of parameters
    }
    
    if (sp$args$model == "within" & sp$args$effect == "time"){
      T = t
      np = np+T+1 # update number of parameters
    }
    
    if (sp$args$model == "within" & sp$args$effect == "twoways"){
      n = c
      T = t
      np = np+n+T # update number of parameters
    }
    aic <- round(       2*np  +  n.N * (  log(2*pi) + s.sq  + 1 ),1)
    bic <- round(log(n.N)*np  +  n.N * (  log(2*pi) + s.sq  + 1 ),1)
    
    if(criterion=="AIC"){
      names(aic) = "AIC"
      return(aic)
    }
    if(criterion=="BIC"){
      names(bic) = "BIC"
      return(bic)
    }
  }
}

sprintf("AR(1) FE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar1_stars_fe, "AIC"), aicbic_plm(ar1_stars_fe, "BIC"))
sprintf("AR(1) RE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar1_stars_re, "AIC"), aicbic_plm(ar1_stars_re, "BIC"))
sprintf("AR(2) FE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar2_stars_fe, "AIC"), aicbic_plm(ar2_stars_fe, "BIC"))
sprintf("AR(2) RE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar2_stars_re, "AIC"), aicbic_plm(ar2_stars_re, "BIC"))
sprintf("AR(3) FE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar3_stars_fe, "AIC"), aicbic_plm(ar3_stars_fe, "BIC"))
sprintf("AR(3) RE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar3_stars_re, "AIC"), aicbic_plm(ar3_stars_re, "BIC"))
sprintf("AR(4) FE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar4_stars_fe, "AIC"), aicbic_plm(ar4_stars_fe, "BIC"))
sprintf("AR(4) RE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar4_stars_re, "AIC"), aicbic_plm(ar4_stars_re, "BIC"))
sprintf("AR(5) FE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar5_stars_fe, "AIC"), aicbic_plm(ar5_stars_fe, "BIC"))
sprintf("AR(5) RE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar5_stars_fe, "AIC"), aicbic_plm(ar5_stars_re, "BIC"))
sprintf("AR(6) FE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar6_stars_fe, "AIC"), aicbic_plm(ar6_stars_fe, "BIC"))
sprintf("AR(6) RE: AIC=%.2f, BIC=%.2f", aicbic_plm(ar6_stars_re, "AIC"), aicbic_plm(ar6_stars_re, "BIC"))
```

```{r}
rob_se <- list(sqrt(diag(vcovHC(ar2_stars_re, type = "HC1"))),
               sqrt(diag(vcovHC(ar2_stars_fe, type = "HC1"))))
stargazer(ar2_stars_re, 
          ar2_stars_fe, 
          digits = 3,
          header = FALSE,
          type = "latex", 
          se = rob_se,
          model.numbers = FALSE)
```

```{r}
rob_se <- list(sqrt(diag(vcovHC(ar1_stars_re, type = "HC1"))),
               sqrt(diag(vcovHC(ar1_stars_fe, type = "HC1"))),
               sqrt(diag(vcovHC(ar3_stars_re, type = "HC1"))),
               sqrt(diag(vcovHC(ar3_stars_fe, type = "HC1"))),
               sqrt(diag(vcovHC(ar4_stars_re, type = "HC1"))),
               sqrt(diag(vcovHC(ar4_stars_fe, type = "HC1"))),
               sqrt(diag(vcovHC(ar5_stars_re, type = "HC1"))),
               sqrt(diag(vcovHC(ar5_stars_fe, type = "HC1"))),
               sqrt(diag(vcovHC(ar6_stars_re, type = "HC1"))),
               sqrt(diag(vcovHC(ar6_stars_fe, type = "HC1"))))
texreg(list(ar1_stars_re,ar1_stars_fe,
            ar3_stars_re,ar3_stars_fe,
            ar4_stars_re,ar4_stars_fe,
            ar5_stars_re,ar5_stars_fe,
            ar6_stars_re,ar6_stars_fe,),
       dcolumn=FALSE,
       override.se=rob_se)
```

# Robustness Checks for Causal Credibility (R1-R4)

These chunks operationalise Step 4 of the causal credibility analysis (see
`plans/20260505 - Fake Stars Causal Credibility Analysis.md`). All four checks
re-fit the AR(2) fixed-effects specification on alternative samples or with a
swapped dependent variable.

## R1 -- Reverse-causality lead-lag test (threatens A1)

Swap the dependent variable: regress detected fake-star intensity on lagged
real-star intensity, conditional on the same fixed-effects structure. A
significantly positive coefficient on lagged real stars indicates that
purchasing behaviour responds to prior organic momentum (not stagnation).

```{r r1-reverse-causality}
rev_causality_fe <- plm(
  log(n_stars_fake + 1) ~
    log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) +
    log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1),
  data = l2_repos, index = c("repo", "month"), model = "within"
)
coeftest(rev_causality_fe, vcov = vcovHC(rev_causality_fe, type = "HC1"))
```

## R2 -- Survivor-length stratification (threatens A5)

Stratify campaign repositories by panel length as a proxy for eventual deletion
(short panels suggest early exit). Re-fit the AR(2) FE model on each stratum.

```{r r2-survivor-length}
panel_lengths <- repos %>%
  group_by(repo) %>%
  summarise(panel_length = n_distinct(month), .groups = "drop")

l2_repos_long  <- l2_repos %>% inner_join(filter(panel_lengths, panel_length >= 12), by = "repo")
l2_repos_short <- l2_repos %>% inner_join(filter(panel_lengths, panel_length <  12), by = "repo")

ar2_fe_long <- plm(log(n_stars_real + 1) ~
                   log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) +
                   log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1),
                   data = l2_repos_long, index = c("repo", "month"), model = "within")
ar2_fe_short <- plm(log(n_stars_real + 1) ~
                    log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) +
                    log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1),
                    data = l2_repos_short, index = c("repo", "month"), model = "within")
coeftest(ar2_fe_long,  vcov = vcovHC(ar2_fe_long,  type = "HC1"))
coeftest(ar2_fe_short, vcov = vcovHC(ar2_fe_short, type = "HC1"))
```

## R4 -- Calendar-period subset sensitivity (threatens A2)

Stratify by calendar period to bound the topic-by-time confounder. The 2024
LLM-surge window is when both fake-star activity and organic AI/LLM interest
peaked simultaneously; if the headline coefficient is materially larger in
this window than outside it, topic-momentum confounding is contributing to
the pooled estimate.

```{r r4-calendar-period}
fit_subset <- function(d) plm(log(n_stars_real + 1) ~
                              log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) +
                              log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1),
                              data = d, index = c("repo", "month"), model = "within")

l2_pre2024  <- l2_repos %>% filter(month <  "2024-01")
l2_2024     <- l2_repos %>% filter(month >= "2024-01" & month < "2024-09")
l2_post2024 <- l2_repos %>% filter(month >= "2024-09")

ar2_fe_pre2024   <- fit_subset(l2_pre2024)
ar2_fe_2024surge <- fit_subset(l2_2024)
ar2_fe_post2024  <- fit_subset(l2_post2024)

coeftest(ar2_fe_pre2024,   vcov = vcovHC(ar2_fe_pre2024,   type = "HC1"))
coeftest(ar2_fe_2024surge, vcov = vcovHC(ar2_fe_2024surge, type = "HC1"))
coeftest(ar2_fe_post2024,  vcov = vcovHC(ar2_fe_post2024,  type = "HC1"))
```

## R5 -- Topic stratification using `repo_labels.csv` (threatens A2)

Stratify the AR(2) FE model by the manual category labels (Section 4.3 of the
chapter) to disentangle topic-momentum confounding from the within-topic
treatment effect. The `repo_labels.csv` file carries one row per labelled
repository (580 in total), with `domain` as the primary categorical variable.
Strata smaller than 30 repositories or 100 panel observations are skipped
because within-FE estimation is unstable for such small panels.

```{r r5-topic-stratification}
labels <- read.csv("repo_labels.csv", stringsAsFactors = FALSE)
l2_labeled <- l2_repos %>% inner_join(labels[, c("repo", "domain")], by = "repo")

domain_counts <- l2_labeled %>%
  group_by(domain) %>%
  summarise(n_obs = n(), n_repos = n_distinct(repo), .groups = "drop") %>%
  arrange(desc(n_obs))
domain_counts

# Pooled estimate restricted to labelled repos (R5 baseline)
ar2_fe_labeled <- fit_subset(l2_labeled)
coeftest(ar2_fe_labeled, vcov = vcovHC(ar2_fe_labeled, type = "HC1"))

# Per-domain estimates (only domains with >= 30 repos and >= 100 obs)
domains_to_fit <- domain_counts %>% filter(n_obs >= 100, n_repos >= 30) %>% pull(domain)
domain_models <- lapply(domains_to_fit, function(dom) {
  d <- l2_labeled %>% filter(domain == dom)
  fit_subset(d)
})
names(domain_models) <- domains_to_fit
for (dom in domains_to_fit) {
  cat("\n--- domain:", dom, "---\n")
  print(coeftest(domain_models[[dom]], vcov = vcovHC(domain_models[[dom]], type = "HC1")))
}
```
