---
title: "Causal Credibility Robustness Checks (R1, R2, R4, R5)"
author: "Fake Stars chapter"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: show
    df_print: kable
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.align = "center")
```

# Overview

This notebook operationalises Step 4 of the causal credibility analysis plan
(`plans/20260505 - Fake Stars Causal Credibility Analysis.md`). All four
checks re-fit the AR(2) fixed-effects specification of the chapter on
alternative samples or with a swapped dependent variable, using the panel
data in `model_stars.csv` and the manual category labels in
`repo_labels.csv`.

| Check | Threatens | Question |
|---|---|---|
| R1 | A1 | Does past organic momentum predict subsequent fake-star buying? |
| R2 | A5 | Is the long-run negative coefficient driven by selective survival? |
| R4 | A2 | Is the headline coefficient concentrated in a topic-hype window? |
| R5 | A2 | Is the coefficient heterogeneous across manually labelled domains? |

R3 (star-monotonicity filter) was withdrawn because `n_stars_real` is computed
from GHArchive `WatchEvent` records, which are additive only --- the panel
cannot register a star removal, so the cumulative real-star series is monotone
by construction.

```{r packages}
suppressPackageStartupMessages({
  library(plm)
  library(dplyr)
  library(tidyr)
  library(sandwich)
  library(lmtest)
  library(knitr)
})
```

# Data

```{r load-data}
repos <- read.csv("model_stars.csv")

l2_repos <- repos %>%
  group_by(repo) %>%
  arrange(month, .by_group = TRUE) %>%
  mutate(
    l1_real = lag(n_stars_real, 1),
    l2_real = lag(n_stars_real, 2),
    l1_fake = lag(n_stars_fake, 1),
    l2_fake = lag(n_stars_fake, 2),
    total_real = cumsum(n_stars_real),
    l3_total_real = lag(total_real, 3),
    total_fake = cumsum(n_stars_fake),
    l3_total_fake = lag(total_fake, 3)
  ) %>%
  ungroup() %>%
  drop_na()

cat(sprintf("l2_repos: n = %d, repos = %d, months = [%s, %s]\n",
            nrow(l2_repos), n_distinct(l2_repos$repo),
            min(l2_repos$month), max(l2_repos$month)))
```

# Helpers

```{r helpers}
fit_ar2_fe <- function(d) {
  plm(log(n_stars_real + 1) ~
        log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) +
        log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1),
      data = d, index = c("repo", "month"), model = "within")
}

robust_coefs <- function(model, label) {
  cv <- coeftest(model, vcov = vcovHC(model, type = "HC1"))
  out <- as.data.frame(unclass(cv))
  out$term <- rownames(out)
  rownames(out) <- NULL
  data.frame(
    spec   = label,
    n_obs  = nobs(model),
    n_repos= length(unique(index(model)[, 1])),
    term   = out$term,
    est    = out$Estimate,
    se     = out$`Std. Error`,
    p      = out$`Pr(>|t|)`,
    stringsAsFactors = FALSE
  )
}

show_coefs <- function(model, label) {
  d <- robust_coefs(model, label) %>%
    mutate(sig = case_when(p < 0.001 ~ "***",
                           p < 0.01  ~ "**",
                           p < 0.05  ~ "*",
                           p < 0.1   ~ ".",
                           TRUE      ~ "")) %>%
    mutate(`est (SE)` = sprintf("%+.4f (%.4f) %s", est, se, sig)) %>%
    select(term, `est (SE)`)
  knitr::kable(d, caption = paste0(label, "  (n=", nobs(model), ")"))
}
```

# Baseline AR(2) FE (sanity check vs. paper)

This replicates Table~\ref{tab:regression} in `main-fake-stars.tex`.

```{r baseline}
ar2_fe <- fit_ar2_fe(l2_repos)
show_coefs(ar2_fe, "Baseline AR(2) FE")
```

# R1 -- Reverse-causality lead-lag test

Swap the dependent variable: regress `log(n_stars_fake + 1)` on lagged real-
star intensity. A significantly *positive* lagged-real coefficient supports
hype-amplification or rich-get-richer in purchasing (biases $\hat\beta_1$ in
the main regression *upward*); a significantly *negative* coefficient
supports panic-buying after stagnation (biases $\hat\beta_1$ *downward*).

```{r r1}
rev_causality_fe <- plm(
  log(n_stars_fake + 1) ~
    log(l1_real + 1) + log(l2_real + 1) + log(l3_total_real + 1) +
    log(l1_fake + 1) + log(l2_fake + 1) + log(l3_total_fake + 1),
  data = l2_repos, index = c("repo", "month"), model = "within"
)
show_coefs(rev_causality_fe, "R1: reverse-causality FE (DV = log(fake+1))")
```

**Reading.** Lagged real-star coefficients are positive and significant at
all three horizons. Past organic momentum predicts subsequent fake-star
buying, suggesting a hype-amplification or rich-get-richer dynamic in
purchasing. The bias on the headline $\hat\beta_1$ from this channel is
upward, so the headline estimate should be treated as an *upper bound* under
this reverse-causality story.

# R2 -- Survivor-length stratification

Stratify campaign repositories by panel length as a proxy for eventual
deletion. Short panels suggest early exit (deletion or abandonment); long
panels are repositories that survived the observation window. Refit AR(2)
FE on each stratum separately.

```{r r2}
panel_lengths <- repos %>%
  group_by(repo) %>%
  summarise(panel_length = n_distinct(month), .groups = "drop")

l2_long  <- l2_repos %>% inner_join(filter(panel_lengths, panel_length >= 12), by = "repo")
l2_short <- l2_repos %>% inner_join(filter(panel_lengths, panel_length <  12), by = "repo")

ar2_fe_long  <- fit_ar2_fe(l2_long)
ar2_fe_short <- fit_ar2_fe(l2_short)

show_coefs(ar2_fe_long,  "R2: panel_length >= 12")
show_coefs(ar2_fe_short, "R2: panel_length < 12")
```

**Reading.** The long-run negative coefficient on `log(l3_total_fake + 1)`
persists in long-survivors ($-0.045^{***}$) and *flips positive* in
short-survivors ($+0.061^{*}$). Both readings remain compatible: the negative
coefficient may reflect user-disengagement that requires time to materialise
(visible only in long panels), or it may reflect selective deletion of
repositories with successful-looking inflated histories before the negative
mechanism plays out (visible in long panels because deletions are filtered
out of long panels by definition).

# R4 -- Calendar-period subset sensitivity

Stratify by calendar period to bound the topic-by-time confounder $Z_t^{\text{topic}}$.

```{r r4}
l2_pre2024  <- l2_repos %>% filter(month <  "2024-01")
l2_2024     <- l2_repos %>% filter(month >= "2024-01" & month < "2024-09")
l2_post2024 <- l2_repos %>% filter(month >= "2024-09")

ar2_fe_pre2024   <- fit_ar2_fe(l2_pre2024)
ar2_fe_2024surge <- fit_ar2_fe(l2_2024)
ar2_fe_post2024  <- fit_ar2_fe(l2_post2024)

show_coefs(ar2_fe_pre2024,   "R4: month < 2024-01 (pre-surge)")
show_coefs(ar2_fe_2024surge, "R4: 2024-01 <= month < 2024-09 (LLM surge)")
show_coefs(ar2_fe_post2024,  "R4: month >= 2024-09 (post-surge)")
```

**Reading.** The headline positive short-run coefficient is significant
*only* in the 2024 LLM-surge window ($+0.080^{***}$). It is statistically
indistinguishable from zero pre-2024 ($+0.031$, n.s.) and post-2024
($+0.030$, n.s.). The long-run negative coefficient is concentrated in the
pre-2024 sample and absent during the surge. Strong evidence of topic-by-time
confounding contributing to the pooled estimate.

# R5 -- Topic stratification using `repo_labels.csv`

Stratify on the manual `domain` labels for the 580 categorised repositories
(of which 375 also appear in the AR(2) panel). Strata smaller than 30
repositories or 100 panel observations are skipped because within-FE
estimation is unstable on small panels.

```{r r5-data}
labels <- read.csv("repo_labels.csv", stringsAsFactors = FALSE)
l2_labeled <- l2_repos %>% inner_join(labels[, c("repo", "domain")], by = "repo")

domain_counts <- l2_labeled %>%
  group_by(domain) %>%
  summarise(n_obs = n(), n_repos = n_distinct(repo), .groups = "drop") %>%
  arrange(desc(n_obs))
knitr::kable(domain_counts, caption = "R5: panel observations and repos per domain")
```

```{r r5-pooled}
ar2_fe_labeled <- fit_ar2_fe(l2_labeled)
show_coefs(ar2_fe_labeled, "R5: pooled labelled repos")
```

```{r r5-strata, results="asis"}
domains_to_fit <- domain_counts %>% filter(n_obs >= 100, n_repos >= 30) %>% pull(domain)
domain_models <- list()
for (dom in domains_to_fit) {
  d <- l2_labeled %>% filter(domain == dom)
  m <- tryCatch(fit_ar2_fe(d), error = function(e) e)
  if (inherits(m, "plm")) {
    domain_models[[dom]] <- m
    cat("\n\n## Domain:", dom, "\n\n")
    print(show_coefs(m, sprintf("R5: domain = %s", dom)))
  }
}
skipped <- domain_counts %>% filter(n_obs < 100 | n_repos < 30) %>% pull(domain)
if (length(skipped) > 0) {
  cat("\n\n**Skipped strata** (below the 30-repo / 100-obs threshold):",
      paste(skipped, collapse = ", "), "\n")
}
```

**Reading.** The pooled $\hat\beta_{l1\_fake}$ on labelled repositories
($+0.067^{***}$) hides an order-of-magnitude variation across domains. The
short-run coefficient is significant in `tool/application` ($+0.120^{*}$),
`tutorial/demo` ($+0.192^{**}$), `ai` ($+0.099^{*}$), and `suspicious`
($+0.068^{*}$); it is null in `blockchain` ($+0.023$, n.s.) and `web`
($+0.003$, n.s.). The long-run negative coefficient concentrates in
`tool/application` ($-0.171^{**}$), `blockchain` ($-0.069^{***}$), and `web`
($-0.054^{.}$), and is absent in `ai`, `suspicious`, and `tutorial/demo`.
The pooled estimate is therefore not a single causal parameter; it is an
average over heterogeneous within-domain effects, complicating any uniform
causal interpretation.

# Combined Coefficient Summary

```{r summary-table}
all_specs <- bind_rows(
  robust_coefs(ar2_fe,           "Baseline AR(2) FE"),
  robust_coefs(ar2_fe_long,      "R2: panel_length >= 12"),
  robust_coefs(ar2_fe_short,     "R2: panel_length < 12"),
  robust_coefs(ar2_fe_pre2024,   "R4: pre-2024"),
  robust_coefs(ar2_fe_2024surge, "R4: 2024 surge"),
  robust_coefs(ar2_fe_post2024,  "R4: post-2024"),
  robust_coefs(ar2_fe_labeled,   "R5: pooled labelled")
) %>%
  bind_rows(do.call(rbind, lapply(names(domain_models), function(dom)
    robust_coefs(domain_models[[dom]], paste0("R5: ", dom)))))

key_terms <- c("log(l1_fake + 1)", "log(l2_fake + 1)", "log(l3_total_fake + 1)")

summary_wide <- all_specs %>%
  filter(term %in% key_terms) %>%
  mutate(sig = case_when(p < 0.001 ~ "***",
                         p < 0.01  ~ "**",
                         p < 0.05  ~ "*",
                         p < 0.1   ~ ".",
                         TRUE      ~ ""),
         cell = sprintf("%+.4f (%.4f) %s", est, se, sig)) %>%
  select(spec, n_obs, n_repos, term, cell) %>%
  pivot_wider(names_from = term, values_from = cell)

knitr::kable(summary_wide, caption = "All specifications: headline fake-star coefficients (HC1 SEs in parentheses)")
```

# R1 reverse regression: lagged-real coefficients

```{r r1-summary}
r1_summary <- robust_coefs(rev_causality_fe, "R1") %>%
  filter(term %in% c("log(l1_real + 1)", "log(l2_real + 1)", "log(l3_total_real + 1)")) %>%
  mutate(sig = case_when(p < 0.001 ~ "***",
                         p < 0.01  ~ "**",
                         p < 0.05  ~ "*",
                         p < 0.1   ~ ".",
                         TRUE      ~ ""),
         cell = sprintf("%+.4f (%.4f) %s", est, se, sig)) %>%
  select(term, cell)
knitr::kable(r1_summary, caption = "R1: lagged-real coefficients in the swapped regression")
```

# Session info

```{r session}
sessionInfo()
```
